### 2015-10-08 paper/paper.tex
Outline of the paper.

\documentclass[smallabstract,smallcaptions]{dccpaper}

\usepackage[utf8]{inputenc}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{textgreek}
\usepackage{url}

\newlength{\figurewidth}
\newlength{\smallfigurewidth}
\setlength{\smallfigurewidth}{2.75in}
\setlength{\figurewidth}{6in}

\begin{document}

\title
{\large
\textbf{Large-scale Burrows-Wheeler compression \\ of sequence data}
}

\author{%
Jouni Sirén \\[0.5em]
{\small\begin{minipage}{\linewidth}\begin{center}
\begin{tabular}{c}
Wellcome Trust Sanger Institute \\
Wellcome Trust Genome Campus \\
Hinxton, Cambridge, CB10 1SA, UK \\
\url{jouni.siren@iki.fi}
\end{tabular}
\end{center}\end{minipage}}
}

\maketitle
\thispagestyle{empty}

\begin{abstract}
Fooo
\end{abstract}

\Section{Introduction}

Sequence data

BWT, compression, FM-index, applications

ReadServer; multiple BWTs

BWT construction: from SA, space efficient algorithms, DNA sequences

Large scale: speed, space, hardware, efficiency

This paper: terabases

\Section{Background}

Definitions

SA, BWT, FM-index

Incremental BWT construction: insert vs extend, batch updates vs dynamic BWT

BWT merging: subcollections, extract and search, build RA, interleave

\Section{Space-efficient BWT merging}

search: trie vs individual sequences, multiple threads

buffering: run buffer, thread buffer (RLE, differential encoding, dynamic paged array), merge buffers, to disk

interleaving: multiway merge from disk in one thread, interleave BWTs in another, dynamic paged array

\Section{Experiments}

implementation, environment

data: ReadServer BWTs, NA12878 trio

ReadServer: trade-offs with different parameters

trio: ropebwt + this vs ropebwt vs ropebwt2

\Section{Discussion}

conclusions

sequence ordering: RLO/RCLO for compression, RLCO for mapping between reverse complements, mapping position for encoding pairing information

remove duplicates

%\Section{References}
%\bibliographystyle{IEEEtran}
%\bibliography{refs}

\end{document}

### 2015-10-09 bwt_convert.cpp
RopeBWT format; input/output formats for bwt_merge

@@ -113,20 +113,7 @@ printUsage()
   std::cerr << "  -o format      Write the output in the given format (default: native)" << std::endl;
   std::cerr << std::endl;
 
-  std::cerr << "Formats supporting any alphabetic order:" << std::endl;
-  printFormat<NativeFormat>(std::cerr);
-  std::cerr << std::endl;
-
-  std::cerr << "Formats using the default alphabet:" << std::endl;
-  printFormat<PlainFormatD>(std::cerr);
-  printFormat<SGAFormat>(std::cerr);
-  std::cerr << std::endl;
-
-  std::cerr << "Formats using sorted alphabet:" << std::endl;
-  printFormat<PlainFormatS>(std::cerr);
-  printFormat<RFMFormat>(std::cerr);
-  printFormat<SDSLFormat>(std::cerr);
-  std::cerr << std::endl;
+  printFormats(std::cerr);
 }
 
 //------------------------------------------------------------------------------

### 2015-10-09 bwt_merge.cpp

@@ -22,6 +22,7 @@
   SOFTWARE.
 */
 
+#include <sstream>
 #include <string>
 #include <unistd.h>
 
@@ -63,8 +64,9 @@ main(int argc, char** argv)
   int c = 0;
   bool verify = false;
   MergeParameters parameters;
-  std::string pattern_name;
-  while((c = getopt(argc, argv, "b:m:r:s:t:d:v:")) != -1)
+  std::string pattern_name, output_format;
+  std::vector<std::string> input_formats;
+  while((c = getopt(argc, argv, "b:m:r:s:t:d:v:i:o:")) != -1)
   {
     switch(c)
     {
@@ -89,24 +91,47 @@ main(int argc, char** argv)
     case 'v':
       pattern_name = optarg; verify = true;
       break;
+    case 'i':
+      tokenize(optarg, input_formats, ',');
+      break;
+    case 'o':
+      output_format = optarg;
+      break;
     case '?':
     default:
       std::exit(EXIT_FAILURE);
     }
   }
-  if(argc - optind < 3)
+
+  int inputs = (argc - 1) - optind;
+  if(inputs < 2)
   {
     std::cerr << "bwt_merge: Output file not specified" << std::endl;
     std::exit(EXIT_FAILURE);
   }
+  if(input_formats.empty())
+  {
+    for(int i = 0; i < inputs; i++) { input_formats.push_back(NativeFormat::tag); }
+  }
+  if(input_formats.size() == 1)
+  {
+    for(int i = 1; i < inputs; i++) { input_formats.push_back(input_formats[0]); }
+  }
+  if(input_formats.size() != (unsigned)inputs)
+  {
+    std::cerr << "bwt_merge: Specified " << input_formats.size() << " formats for "
+              << inputs << " inputs" << std::endl;
+    std::exit(EXIT_FAILURE);
+  }
+  if(output_format.length() == 0) { output_format = NativeFormat::tag; }
   parameters.sanitize();
   Parallel::max_threads = parameters.threads;
 
   for(int i = optind; i < argc - 1; i++)
   {
-    std::cout << "Input:            " << argv[i] << std::endl;
+    std::cout << "Input:            " << argv[i] << " (" << input_formats[i - optind] << ")" << std::endl;
   }
-  std::cout << "Output:           " << argv[argc - 1] << std::endl;
+  std::cout << "Output:           " << argv[argc - 1] << " (" << output_format << ")" << std::endl;
   if(verify)
[…]

### 2015-10-09 fmi.cpp

@@ -393,6 +393,10 @@ serialize(const FMI& fmi, const std::string& filename, const std::string& format
   {
     fmi.serialize<SDSLFormat>(filename);
   }
+  else if(format == RopeFormat::tag)
+  {
+    fmi.serialize<RopeFormat>(filename);
+  }
   else if(format == SGAFormat::tag)
   {
     fmi.serialize<SGAFormat>(filename);
@@ -426,6 +430,10 @@ load(FMI& fmi, const std::string& filename, const std::string& format)
   else if(format == SDSLFormat::tag)
   {
     fmi.load<SDSLFormat>(filename);
+  }
+  else if(format == RopeFormat::tag)
+  {
+    fmi.load<RopeFormat>(filename);
   }
   else if(format == SGAFormat::tag)
   {

### 2015-10-09 formats.cpp

@@ -115,6 +115,9 @@ const std::string RFMFormat::tag = "rfm";
 const std::string SDSLFormat::name = "SDSL format";
 const std::string SDSLFormat::tag = "sdsl";
 
+const std::string RopeFormat::name = "RopeBWT format";
+const std::string RopeFormat::tag = "ropebwt";
+
 const std::string SGAFormat::name = "SGA format";
 const std::string SGAFormat::tag = "sga";
 
@@ -275,7 +278,7 @@ SDSLFormat::write(std::ofstream& out, const BlockArray& data, const NativeHeader
 
 //------------------------------------------------------------------------------
 
-struct SGAData
+struct RopeData
 {
   typedef bwtmerge::comp_type comp_type;
   typedef bwtmerge::size_type length_type;
@@ -289,43 +292,64 @@ struct SGAData
   const static size_type RUN_BITS = 5;
   const static size_type MAX_RUN = 31;
   const static size_type SIGMA = 6;
+
+  static void read(std::ifstream& in, size_type bytes, BlockArray& data, sdsl::int_vector<64>& counts)
+  {
+    data.clear();
+    counts = sdsl::int_vector<64>(SIGMA, 0);
+
+    code_type buffer[MEGABYTE];
+    RunBuffer run_buffer;
+    for(size_type offset = 0; offset < bytes; offset += MEGABYTE)
+    {
+      size_type bytes = std::min(MEGABYTE, bytes - offset);
+      in.read((char*)buffer, bytes);
+      for(size_type i = 0; i < bytes; i++)
+      {
+        if(run_buffer.add(comp(buffer[i]), length(buffer[i])))
+        {
+          Run::write(data, run_buffer.run);
+          counts[run_buffer.run.first] += run_buffer.run.second;
+        }
+      }
+    }
+    run_buffer.flush();
+    Run::write(data, run_buffer.run);
+    counts[run_buffer.run.first] += run_buffer.run.second;
+  }
+
+  static void write(std::ofstream& out, const BlockArray& data)
+  {
+    size_type rle_pos = 0;
+    std::vector<code_type> buffer; buffer.reserve(MEGABYTE);
+    while(rle_pos < data.size())
+    {
+      range_type run = Run::read(data, rle_pos);
+      while(run.second > MAX_RUN)
+      {
+        buffer.push_back(encode(run.first, MAX_RUN));
+        run.second -= MAX_RUN;
+        if(buffer.size() >= MEGABYTE)
+        {
[…]

### 2015-10-09 formats.h

@@ -81,6 +81,7 @@ std::ostream& operator<<(std::ostream& stream, const NativeHeader& header);
   PlainFormatS  BWT as a character array; AO_SORTED
   RFMFormat     BWT as int_vector<8> of comp values; AO_SORTED
   SDSLFormat    BWT as int_vector<8> of characters; AO_SORTED
+  RopeFormat    RopeBWT; AO_DEFAULT
   SGAFormat     SGA assembler; AO_DEFAULT
 */
 
@@ -134,6 +135,16 @@ struct SDSLFormat
   const static std::string tag;
 };
 
+struct RopeFormat
+{
+  static void read(std::ifstream& in, BlockArray& data, sdsl::int_vector<64>& counts);
+  static void write(std::ofstream& out, const BlockArray& data, const NativeHeader& info);
+  inline static AlphabeticOrder order() { return AO_DEFAULT; }
+
+  const static std::string name;
+  const static std::string tag;
+};
+
 struct SGAFormat
 {
   static void read(std::ifstream& in, BlockArray& data, sdsl::int_vector<64>& counts);
@@ -146,6 +157,8 @@ struct SGAFormat
 
 //------------------------------------------------------------------------------
 
+void printFormats(std::ostream& stream);
+
 template<class Format>
 void
 printFormat(std::ostream& stream)
@@ -156,6 +169,22 @@ printFormat(std::ostream& stream)
 }
 
 //------------------------------------------------------------------------------
+
+struct RopeHeader
+{
+  uint32_t tag;
+
+  const static uint32_t DEFAULT_TAG = 0x06454C52;
+  const static size_type SIZE = 4;
+
+  RopeHeader();
+
+  size_type serialize(std::ostream& out, sdsl::structure_tree_node* v = nullptr, std::string name = "") const;
+  void load(std::istream& in);
+  bool check() const;
+};
+
+std::ostream& operator<<(std::ostream& stream, const RopeHeader& header);
 
 struct SGAHeader
 {

### 2015-10-09 utils.cpp

@@ -66,6 +66,14 @@ printTime(const std::string& header, size_type queries, double seconds, size_typ
             << inMicroseconds(seconds / queries) << " µs/query)" << std::endl;
 }
 
+void
+tokenize(const std::string& source, std::vector<std::string>& tokens, char delim)
+{
+  std::istringstream ss(source);
+  std::string token;
+  while(std::getline(ss, token, delim)) { tokens.push_back(token); }
+}
+
 //------------------------------------------------------------------------------
 
 double

### 2015-10-09 utils.h

@@ -208,6 +208,8 @@ void printSize(const std::string& header, size_type bytes, size_type data_size, 
 void printTime(const std::string& header, size_type found, size_type matches, size_type bytes, double seconds, size_type indent = DEFAULT_INDENT);
 void printTime(const std::string& header, size_type queries, double seconds, size_type indent = DEFAULT_INDENT);
 
+void tokenize(const std::string& source, std::vector<std::string>& tokens, char delim);
+
 //------------------------------------------------------------------------------
 
 double readTimer();       // Seconds from an arbitrary time point.

### 2015-10-09 formats.cpp
Fixed a silly bug

@@ -302,9 +302,9 @@ struct RopeData
     RunBuffer run_buffer;
     for(size_type offset = 0; offset < bytes; offset += MEGABYTE)
     {
-      size_type bytes = std::min(MEGABYTE, bytes - offset);
-      in.read((char*)buffer, bytes);
-      for(size_type i = 0; i < bytes; i++)
+      size_type block_bytes = std::min(MEGABYTE, bytes - offset);
+      in.read((char*)buffer, block_bytes);
+      for(size_type i = 0; i < block_bytes; i++)
       {
         if(run_buffer.add(comp(buffer[i]), length(buffer[i])))
         {

### 2015-10-10 formats.cpp
Fixed the RopeBWT format.

@@ -280,25 +280,16 @@ SDSLFormat::write(std::ofstream& out, const BlockArray& data, const NativeHeader
 
 struct RopeData
 {
-  typedef bwtmerge::comp_type comp_type;
-  typedef bwtmerge::size_type length_type;
-  typedef uint8_t             code_type;
-
-  inline static code_type encode(comp_type comp, length_type length) { return (comp << RUN_BITS) | length; }
-  inline static comp_type comp(code_type code) { return (code >> RUN_BITS); }
-  inline static length_type length(code_type code) { return (code & RUN_MASK); }
-
-  const static code_type RUN_MASK = 0x1F;
-  const static size_type RUN_BITS = 5;
   const static size_type MAX_RUN = 31;
   const static size_type SIGMA = 6;
 
+  template<class Coder>
   static void read(std::ifstream& in, size_type bytes, BlockArray& data, sdsl::int_vector<64>& counts)
   {
     data.clear();
     counts = sdsl::int_vector<64>(SIGMA, 0);
 
-    code_type buffer[MEGABYTE];
+    typename Coder::code_type buffer[MEGABYTE];
     RunBuffer run_buffer;
     for(size_type offset = 0; offset < bytes; offset += MEGABYTE)
     {
@@ -306,7 +297,7 @@ struct RopeData
       in.read((char*)buffer, block_bytes);
       for(size_type i = 0; i < block_bytes; i++)
       {
-        if(run_buffer.add(comp(buffer[i]), length(buffer[i])))
+        if(run_buffer.add(Coder::comp(buffer[i]), Coder::length(buffer[i])))
         {
           Run::write(data, run_buffer.run);
           counts[run_buffer.run.first] += run_buffer.run.second;
@@ -318,16 +309,17 @@ struct RopeData
     counts[run_buffer.run.first] += run_buffer.run.second;
   }
 
+  template<class Coder>
   static void write(std::ofstream& out, const BlockArray& data)
   {
     size_type rle_pos = 0;
-    std::vector<code_type> buffer; buffer.reserve(MEGABYTE);
+    std::vector<typename Coder::code_type> buffer; buffer.reserve(MEGABYTE);
     while(rle_pos < data.size())
     {
       range_type run = Run::read(data, rle_pos);
       while(run.second > MAX_RUN)
       {
-        buffer.push_back(encode(run.first, MAX_RUN));
+        buffer.push_back(Coder::encode(run.first, MAX_RUN));
[…]

### 2015-10-20 paper/paper.tex
Started writing the paper.

@@
 
+% Mathematics
+\newcommand{\set}[1]{\ensuremath{\{ #1 \}}}
+\newcommand{\abs}[1]{\ensuremath{\lvert #1 \rvert}}
+\newcommand{\Oh}{\ensuremath{\mathsf{O}}}
+\newcommand{\oh}{\ensuremath{\mathsf{o}}}
+\newcommand{\Th}{\ensuremath{\mathsf{\Theta}}}
+
+% Structures
+\newcommand{\SA}{\textsf{SA}}
+\newcommand{\ISA}{\textsf{ISA}}
+\newcommand{\BWT}{\textsf{BWT}}
+\newcommand{\FMI}{\textsf{FMI}}
+\newcommand{\C}{\textsf{C}}
+\newcommand{\mSA}{\ensuremath{\mathsf{SA}}}
+\newcommand{\mISA}{\ensuremath{\mathsf{ISA}}}
+\newcommand{\mBWT}{\ensuremath{\mathsf{BWT}}}
+\newcommand{\mC}{\ensuremath{\mathsf{C}}}
+
+% Operations
+\newcommand{\LF}{\textsf{LF}}
+\newcommand{\rank}{\textsf{rank}}
+\newcommand{\select}{\textsf{select}}
+\newcommand{\mLF}{\ensuremath{\mathsf{LF}}}
+\newcommand{\mrank}{\ensuremath{\mathsf{rank}}}
+\newcommand{\mselect}{\ensuremath{\mathsf{select}}}
+
 \begin{document}
@@
 {\large
-\textbf{Large-scale Burrows-Wheeler compression \\ of sequence data}
+\textbf{Burrows-Wheeler transform for terabases}
 }
@@
 
-BWT, compression, FM-index, applications
+BWT, compression, FM-index, applications, repetitive, rle
 
@@
 
-BWT construction: from SA, space efficient algorithms, DNA sequences
+Large scale construction: speed, space, hardware, efficiency
 
-Large scale: speed, space, hardware, efficiency
+BWT construction: from SA \cite{Mori2008,Nong2011}
+external SA \cite{Gonnet1992,Bingmann2013,Kaerkkaeinen2014a,Nong2014,Kaerkkaeinen2015a}
+space efficient algorithms \cite{Hon2007,Kaerkkaeinen2007,Siren2009}
+external BWT \cite{Ferragina2012,Beller2013}
+
+Specialized for DNA: \cite{Bauer2013,Liu2014,Li2014a,Pantaleoni2014,Wang2015}
 
@@
 
-Definitions
+A \emph{string} $S[1,n] = s_{1} \dotsm s_{n}$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{1, \dotsc, \sigma}$.
+For indexing purposes, we consider \emph{text} strings $T[1,n]$ terminated by an endmarker $T[n] = \$ = 0$ not occurring anywhere else in the text.
+\emph{Binary} sequences are strings over the alphabet $\set{0, 1}$.
+A \emph{substring} of string $S$ is a sequence of the form $S[i,j] = s_{i} \dotsm s_{j}$.
+Substrings of the type $S[1,j]$ and $S[i,n]$ are called \emph{prefixes} and \emph{suffixes}, respectively.
 
-SA, BWT, FM-index
+The \emph{suffix array} (\SA) \cite{Manber1993} is a simple full-text index of a string.
+Given a text $T$, its suffix array $\mSA_{T}[1,n]$ is an array of pointers to the suffixes of the text in \emph{lexicographic order}.
+(If the text is evident from the context, we will omit it and write just \SA.) We can build the suffix array in $\Oh(n)$ time using approximately $n(\log n + \log \sigma + 2)$ bits of space, or just $2n$ bits of working space over the text and the suffix array \cite{Nong2011}.
+Given a \emph{pattern} $P$, we can find the \emph{lexicographic range} $[sp,ep]$ of suffixes of the text prefixed by the pattern in $\Oh(\abs{P} \log n)$ time by binary searching in the suffix array.
+The corresponding range of pointers $\mSA[sp,ep]$ lists the \emph{occurrences} of the pattern in the text.
 
-Incremental BWT construction: insert vs extend, batch updates vs dynamic BWT
+While the suffix array is simple, it requires several times more memory than the original text.
+For large texts, this can be a serious drawback.
+The more space-efficient alternatives to the suffix array are based on the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994}, an easily reversible permutation of the text with a similar combinatorial structure to the suffix array.
+Given a text $T[1,n]$ and its suffix array, we can easily produce the \BWT{} as $\mBWT[i] = T[\mSA[i]-1]$ (with $\mBWT[i] = T[n]$, if $\mSA[i] = 1$).
+
+A key observation is that if suffix $X$ precedes suffix $Y$ in lexicographic order, suffix $cX$ will also precede suffix $cY$, for any character $c$.
+Therefore, if $\mBWT[i]$ is the $j$\nobreakdash-th occurrence of character $c$ in the \BWT, and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order.
+This forms the essence of \LF\emph{\nobreakdash-mapping}.
+The suffix preceding $T[\mSA[i],n]$ in text order is the one starting at $\mSA[\mLF(i)]$, where
+$$
+\mLF(i) = \mC[\mBWT[i]] + \mBWT.\mrank(i, \mBWT[i]),
+$$
+$\mC[c]$ is the number of suffixes starting with a character smaller than $c$, and $S.\mrank(i,c)$ is the number of occurrences of character $c$ in the prefix $S[1,i]$.
+For searching, we use a more general definition of \LF\nobreakdash-mapping:
+$$
+\mLF(i,c) = \mC[c] + \mBWT.\mrank(i, c),
+$$
+which tells the number of suffixes $X$ of text $T$ with $X \le c T[\mSA[i],n]$ in lexicographic order.
+This is known as the \emph{lexicographic rank} $\mrank(cT[\mSA[i],n], T)$ of string $cT[\mSA[i],n]$ among the suffixes of text $T$.
+
+The \emph{FM-index} (\FMI) \cite{Ferragina2005a} uses the Burrows-Wheeler transform as a full-text index by augmenting it with a few additional structures.
+The lexicographic range $[sp,ep]$ corresponding to pattern $P$ is found through \emph{backward searching}.
+Let $[sp_{i},ep_{i}]$ be the lexicographic range of suffixes of text $T$ matching suffix $P[i, \abs{P}]$ of the pattern.
+We can find $[sp_{i-1},ep_{i-1}]$ as
+$$
+[sp_{i-1},ep_{i-1}] = [\mLF(sp_{i}-1, P[i-1]) + 1, \mLF(ep_{i}, P[i-1])].
+$$
+By starting from $[sp_{\abs{P}}, ep_{\abs{P}}] = [\mC[P[\abs{P}]]+1, \mC[P[\abs{P}]+1]]$, we can find the lexicographic range of suffixes starting with the pattern in $\Oh(\abs{P} \cdot t_{r})$ time, where $t_{r}$ is the time required to answer \rank{} queries on the \BWT.
+In practice, the time complexity ranges from $\Oh(\abs{P})$ to $\Oh(\abs{P} \log n)$, depending on the encoding.
+
+In order to find the occurrences of the pattern in the text, the FM-index \emph{samples} some suffix array pointers, and uses \LF\nobreakdash-mapping to derive the unsampled pointers.
+The set of sampled pointers always includes a pointer to the beginning of each text.
+If $\mSA[i]$ is not sampled, the FM-index proceeds to $\mLF(i)$ and continues from there.
+If $\mSA[\mLF^{k}(i)]$ is the first sample encountered, $\mSA[i] = \mSA[\mLF^{k}(i)] + k$.
+Depending on the way the samples are selected, we may need a binary sequence to mark the pointers that have been sampled.
+
+Assume that we have an ordered \emph{collection} of texts $\mathcal{A} = (T_{1}, \dotsc, T_{m})$ of total length $n = \abs{\mathcal{A}} = \sum_{i} \abs{T_{i}}$.
+We want to build a (generalized) \BWT{} for the collection.
+The usual way is to assume that the endmarkers of all text are distinct, with the one at the end of text $T_{i}$ having character value $(0,i)$.
+This guarantees that each suffix has a unique lexicographic rank among the suffixes of the collection.
+We still encode each endmarker as $0$ in the \BWT{} in order to save space.
+Because of this, \LF\nobreakdash-mapping does not work with $c = 0$, and we cannot match patterns spanning text boundaries.
+
+With large collections of short texts (such as short reads), there are more space-efficient alternatives to sampling.
+Because all endmarkers had distinct character values during sorting, we know that $\mSA[i]$ (with $i \le m)$ points to the endmarker of text $T_{i}$.
+To reach the endmarker, we iterate
+$$
+\Psi(i) = \mBWT.\mselect(i - \mC[c], c),
+$$
+where $c$ is the largest character value with $\mC[c] < i$ and $S.\mselect(i,c)$ finds the $i$\nobreakdash-th occurrence of character $c$ in string $S$.
+If $k \ge 0$ is the smallest value for which $j = \Psi^{k}(i) \le m$, we know that $\mSA[i]$ points to offset $\abs{T_{j}} - k$ in text $T_{j}$.
+
+\Section{Space-efficient \BWT{} construction}
+
+The FM-index was introduced as a more space-efficient alternative to the suffix array.
+If we have to build a suffix array first in order to construct an FM-index, we lose a large part of the benefits, and index construction becomes the overall bottleneck.
+To overcome this bottleneck, we can use \emph{incremental construction algorithms} that build the FM-index directly.
+Some of them use an adjustable amount of working space on top of the FM-index, making it possible to index text colletions that are too large to fit in memory in an uncompressed form.
+
+Assume that we have built the \BWT{} of text $T$, and we want to \emph{transform} the \BWT{} into that of text $cT$, where $c$ is a character.
+The transformation \cite{Hon2007} is quite straightforward.
+We first find the pointer $\mSA[i]$ to the beginning of text $T$ (the position $i$, where $\mBWT[i] = 0$).
+Then we determine the lexicographic rank $j = \mrank(cT, T) = \mC[c] + \mBWT.\mrank(i, c)$ of text $cT$ among the suffixes of text $T$.
+Finally we transform the \BWT{} by \emph{replacing} the endmarker at $\mBWT[i]$ with the inserted character $c$, and by \emph{inserting} a new endmarker between $\mBWT[j]$ and $\mBWT[j+1]$.
+
+There are six basic ways of using the transformation in a \BWT{} construction algorithm.
+Instead of inserting a single character at a time, we may use \emph{batch updates} and transform the \BWT{} of text $T$ directly into that of text $XT$, where $X$ is any string \cite{Hon2007}.
+We can start with the \BWT{}s of text collections $\mathcal{A}$ and $\mathcal{B}$, and \emph{merge} them into the \BWT{} of collection $\mathcal{A} \cup \mathcal{B}$ \cite{Siren2009}.
+We can also have the \BWT{} of a text collection, and \emph{extend} multiple texts by inserting a new character to the beginning of each of them in a single update \cite{Bauer2013}.
+In all three cases, we may use either \emph{static} or \emph{dynamic} \cite{Chan2007} structures for the \BWT.
+
+\Section{Merging \BWT{}s of text collections}
+
+% FIXME: a running example with 2 texts
+Assume that we have
+
+The \BWT{} merging algorithm \cite{Siren2009} consists of three phases.
 
 BWT merging: subcollections, extract and search, build RA, interleave
+rank array / interleaving bitvector / gap array \cite{Gonnet1992,Ferragina2012}
 
-\Section{Space-efficient BWT merging}
+\Section{Large-scale \BWT{} merging}
 
@@
 
-%\Section{References}
-%\bibliographystyle{IEEEtran}
-%\bibliography{refs}
+\Section{References}
+\bibliographystyle{IEEEtran}
+\bibliography{paper}
 

### 2015-10-21 paper/paper.tex
Merging algorithm.

@@
 \newcommand{\C}{\textsf{C}}
+\newcommand{\RA}{\textsf{RA}}
 \newcommand{\mSA}{\ensuremath{\mathsf{SA}}}
@@
 \newcommand{\mC}{\ensuremath{\mathsf{C}}}
+\newcommand{\mRA}{\ensuremath{\mathsf{RA}}}
 
@@
 \newcommand{\mselect}{\ensuremath{\mathsf{select}}}
+
+% Shorthands
+\newcommand{\Acoll}{\ensuremath{\mathcal{A}}}
+\newcommand{\Bcoll}{\ensuremath{\mathcal{B}}}
 
@@
 external SA \cite{Gonnet1992,Bingmann2013,Kaerkkaeinen2014a,Nong2014,Kaerkkaeinen2015a}
-space efficient algorithms \cite{Hon2007,Kaerkkaeinen2007,Siren2009}
+space efficient algorithms \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}
 external BWT \cite{Ferragina2012,Beller2013}
@@
 
-Assume that we have an ordered \emph{collection} of texts $\mathcal{A} = (T_{1}, \dotsc, T_{m})$ of total length $n = \abs{\mathcal{A}} = \sum_{i} \abs{T_{i}}$.
+Assume that we have an ordered \emph{collection} of texts $\Acoll = (T_{1}, \dotsc, T_{m})$ of total length $n = \abs{\Acoll} = \sum_{i} \abs{T_{i}}$.
 We want to build a (generalized) \BWT{} for the collection.
@@
 If $k \ge 0$ is the smallest value for which $j = \Psi^{k}(i) \le m$, we know that $\mSA[i]$ points to offset $\abs{T_{j}} - k$ in text $T_{j}$.
+
+We can \emph{extract} text $T_{i}$ in $\Oh(\abs{T_{i}} \cdot t_{r})$ time by using \LF\nobreakdash-mapping \cite{Burrows1994}.
+We start from the endmarker at $\mBWT[i]$ and extract the text backwards as
+$$
+T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)] \,\, \textrm{for} \,\, 1 \le j < \abs{T_{i}}.
+$$
 
@@
 Instead of inserting a single character at a time, we may use \emph{batch updates} and transform the \BWT{} of text $T$ directly into that of text $XT$, where $X$ is any string \cite{Hon2007}.
-We can start with the \BWT{}s of text collections $\mathcal{A}$ and $\mathcal{B}$, and \emph{merge} them into the \BWT{} of collection $\mathcal{A} \cup \mathcal{B}$ \cite{Siren2009}.
+We can start with the \BWT{}s of text collections $\Acoll$ and $\Bcoll$, and \emph{merge} them into the \BWT{} of collection $\Acoll \cup \Bcoll$ \cite{Siren2009}.
 We can also have the \BWT{} of a text collection, and \emph{extend} multiple texts by inserting a new character to the beginning of each of them in a single update \cite{Bauer2013}.
@@
 % FIXME: a running example with 2 texts
-Assume that we have
+Assume that we have two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively, and that we have already built \BWT{}s for them.
+The algorithm for merging $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ into $\mBWT_{\Acoll \cup \Bcoll}$ works in three phases \cite{Siren2009}:
+\begin{enumerate}
+\item \textbf{Search.} We search for all texts in collection $\Bcoll$ using $\mBWT_{\Acoll}$ and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of collection $\Bcoll$.
+This takes $\Oh(n_{\Bcoll} t_{r})$ time.
+If we do not have collection $\Bcoll$ in plain form, we can extract the texts from $\mBWT_{\Bcoll}$ in the same asymptotic time.
+\item \textbf{Sort.} We sort the ranks produced in the search phase to build the \emph{rank array} (\RA) of collection $\Bcoll$ relative to collection $\Acoll$.
+The rank array is defined as $\mRA_{\Bcoll \mid \Acoll}[i] = \mrank(X, \Acoll)$, where $X$ is the suffix pointed by $\mSA_{\Bcoll}[i]$.
+The array requires $n_{\Bcoll} \log n_{\Acoll}$ bits of space, and we can build it in $\Oh(sort(n_{\Bcoll}, n_{\Acoll}))$ time, where $sort(n, u)$ is the time required to sort $n$ integers from universe $[0,u]$.
+\item \textbf{Merge.} We interleave $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ according to the rank array.
+If $\mRA_{\Bcoll \mid \Acoll}[i] = j$, the merged \BWT{} will have $j$ characters from $\mBWT_{\Acoll}$ before $\mBWT_{\Bcoll}[i]$.
+This phase takes $\Oh(n_{\Acoll} + n_{\Bcoll})$ time and uses $\Oh(b)$ bits of working space, where $b$ is the size of virtual memory pages.
+\end{enumerate}
+The overall time complexity of \BWT{} merging is $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r} + sort(n_{\Bcoll}, n_{\Acoll}))$.
+The algorithm uses $n_{\Bcoll} \log n_{\Acoll} + \Oh(b)$ bits of working space in addition to the \BWT{}s, the structures required to use them as FM-indexes, and the working space used by the sorting algorithm.
 
-The \BWT{} merging algorithm \cite{Siren2009} consists of three phases.
-
-BWT merging: subcollections, extract and search, build RA, interleave
-rank array / interleaving bitvector / gap array \cite{Gonnet1992,Ferragina2012}
+In an alternative approach, we build the \emph{interleaving bitvector} $B_{\Acoll \cup \Bcoll}$.
+The bitvector is a binary sequence of length $n_{\Acoll} + n_{\Bcoll}$ and encodes the same information as the rank array.
+We set $B_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]] = 1$ for $1 \le i \le n_{\Bcoll}$.
+If we extract the texts from $\mBWT_{\Bcoll}$ in the search phase, this can be done directly using the information available at that point.
+This approach uses $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r})$ time and $n_{\Acoll} + n_{\Bcoll} + \Oh(b)$ bits of working space.
 
 \Section{Large-scale \BWT{} merging}
+
+main problem: memory size may be $\oh(\min(n_{\Bcoll} \log n_{\Acoll}, n_{\Acoll} + n_{\Bcoll}))$ bits even if a run-length encoded \BWT{} fits in memory; space overhead from dynamic representations also too large
 

### 2015-10-21 paper/paper.tex
Sketched some details.

@@
 
-ReadServer; multiple BWTs
+ReadServer; multiple BWTs due to construction issues; less BWTs would mean less memory, less querying
 
@@
 
+sort is interleaved with search and merge
+
 search: trie vs individual sequences, multiple threads
 
-buffering: run buffer, thread buffer (RLE, differential encoding, dynamic paged array), merge buffers, to disk
+buffering: run buffer, thread buffer (RLE, differential encoding), merge buffers, to disk
 
-interleaving: multiway merge from disk in one thread, interleave BWTs in another, dynamic paged array
+interleaving: multiway merge from disk in one thread, interleave BWTs in another
+
+\Section{Implementation}
+
+C++, compiler, github
+
+details of run-length encoding of BWT, RA
+
+number of threads
+
+use 8~MB blocks instead of virtual memory pages
 
@@
 
-implementation, environment
+environment
 
-data: ReadServer BWTs, NA12878 trio
+data: ReadServer BWTs, NA12878 trio \cite{1000GP2015}
+error correction using bfc \cite{Li2015}
 
@@
 
-sequence ordering: RLO/RCLO for compression, RLCO for mapping between reverse complements, mapping position for encoding pairing information
+sequence ordering: original orderin may be important, RLO/RCLO for compression, RLCO for mapping between reverse complements, mapping position for encoding pairing information
 

### 2015-10-22 paper/paper.tex
Large-scale merging.

@@
 $$
+As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also extract the lexicographic ranks of all suffixes of text $T_{i}$ in the process.
 
@@
 Assume that we have two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively, and that we have already built \BWT{}s for them.
-The algorithm for merging $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ into $\mBWT_{\Acoll \cup \Bcoll}$ works in three phases \cite{Siren2009}:
+Assume that we store the \BWT{}s in a two-level array, where the first level contains pointers to \emph{blocks} of size $b$ (bits).
+If the \BWT{} takes $x$ bits, the two-level array imposes $\frac{x}{b} \log x + \Oh(b)$ bits of space overhead.
+With $b = \sqrt{x \log x}$, the overhead becomes $\Oh(\sqrt{x \log x})$ bits.
+
+The algorithm for merging $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ into $\mBWT_{\Acoll \cup \Bcoll}$ has three phases \cite{Siren2009}:
 \begin{enumerate}
+
 \item \textbf{Search.} We search for all texts in collection $\Bcoll$ using $\mBWT_{\Acoll}$ and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of collection $\Bcoll$.
@@
 If we do not have collection $\Bcoll$ in plain form, we can extract the texts from $\mBWT_{\Bcoll}$ in the same asymptotic time.
+
 \item \textbf{Sort.} We sort the ranks produced in the search phase to build the \emph{rank array} (\RA) of collection $\Bcoll$ relative to collection $\Acoll$.
@@
 The array requires $n_{\Bcoll} \log n_{\Acoll}$ bits of space, and we can build it in $\Oh(sort(n_{\Bcoll}, n_{\Acoll}))$ time, where $sort(n, u)$ is the time required to sort $n$ integers from universe $[0,u]$.
+If we extract the texts from $\mBWT_{\Bcoll}$ in the search phase, we can write the ranks directly into the rank array, making this phase unnecessary.
+
 \item \textbf{Merge.} We interleave $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ according to the rank array.
 If $\mRA_{\Bcoll \mid \Acoll}[i] = j$, the merged \BWT{} will have $j$ characters from $\mBWT_{\Acoll}$ before $\mBWT_{\Bcoll}[i]$.
-This phase takes $\Oh(n_{\Acoll} + n_{\Bcoll})$ time and uses $\Oh(b)$ bits of working space, where $b$ is the size of virtual memory pages.
+This phase takes $\Oh(n_{\Acoll} + n_{\Bcoll})$ time.
+By reusing the blocks of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ for $\mBWT_{\Acoll \cup \Bcoll}$ once we have processed them, we can merge the \BWT{}s almost in-place, using $\Oh(\sqrt{x \log x})$ bits of working space, where $x$ is the maximum of the sizes of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ in bits.
+
 \end{enumerate}
-The overall time complexity of \BWT{} merging is $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r} + sort(n_{\Bcoll}, n_{\Acoll}))$.
-The algorithm uses $n_{\Bcoll} \log n_{\Acoll} + \Oh(b)$ bits of working space in addition to the \BWT{}s, the structures required to use them as FM-indexes, and the working space used by the sorting algorithm.
+The overall time complexity of \BWT{} merging is $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r})$.
+The algorithm uses $n_{\Bcoll} \log n_{\Acoll} + \Oh(\sqrt{x \log x})$ bits of working space in addition to the \BWT{}s and the structures required to use them as FM-indexes.
 
-In an alternative approach, we build the \emph{interleaving bitvector} $B_{\Acoll \cup \Bcoll}$.
-The bitvector is a binary sequence of length $n_{\Acoll} + n_{\Bcoll}$ and encodes the same information as the rank array.
-We set $B_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]] = 1$ for $1 \le i \le n_{\Bcoll}$.
-If we extract the texts from $\mBWT_{\Bcoll}$ in the search phase, this can be done directly using the information available at that point.
-This approach uses $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r})$ time and $n_{\Acoll} + n_{\Bcoll} + \Oh(b)$ bits of working space.
+Alternatively, we can build the \emph{interleaving bitvector} $B_{\Acoll \cup \Bcoll}$.
+This bitvector is a binary sequence of length $n_{\Acoll} + n_{\Bcoll}$ that encodes the same information as the rank array.
+We set $B_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]] = 1$ for $1 \le i \le n_{\Bcoll}$, denoting that $\mSA_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]]$ points to a suffix of collection $\Bcoll$.
+This approach uses $n_{\Acoll} + n_{\Bcoll} + \Oh(\sqrt{x \log x})$ bits of working space.
 
@@
 
-main problem: memory size may be $\oh(\min(n_{\Bcoll} \log n_{\Acoll}, n_{\Acoll} + n_{\Bcoll}))$ bits even if a run-length encoded \BWT{} fits in memory; space overhead from dynamic representations also too large
+If we split text collection $\Acoll$ of total length $n$ into $p$ \emph{subcollections} of equal size, we can build $\mBWT_{\Acoll}$ by merging the \BWT{}s of the subcollections incrementally.
+Merging the subcollections takes $\Oh((p+t_{r})n)$ time and uses (essentially) $\min(n, \frac{n}{p} \log n)$ bits of working space.
 
-sort is interleaved with search and merge
+When the collection is large, the space overhead of the construction algorithm often determines whether we can build the \BWT{} of the collection.
+Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large.
+The space overhead from the rank array or the interleaving bitvector (or their equivalents in the other space-efficient algorithms) may also be too much.
+We can make the rank array fit in memory by increasing the number of subcollections, but that can make the construction too slow.
 
-search: trie vs individual sequences, multiple threads
+The key to reducing the space overhead is that we do not need to keep the rank array in memory.
+We can write the lexicographic ranks to \emph{disk} in the search phase, sort them to build the rank array in the sort phase, and scan the rank array once during the merge phase.
+In practice, we may want to \emph{interleave} the sort phase with the search/merge phases and store the lexicographic ranks on disk in a \emph{compressed form}, because the amount of fast disk space can also become a bottleneck.
+In the following, we describe the key ideas that make large-scale \BWT{} merging fast and space-efficient.
 
-buffering: run buffer, thread buffer (RLE, differential encoding), merge buffers, to disk
+\SubSection{Search}
 
-interleaving: multiway merge from disk in one thread, interleave BWTs in another
+Instead of searching for every text in collection $\Bcoll$ separately, we can search for the \emph{reverse trie} of the collection.
+Assume that there are $m_{\Acoll}$ texts in collection $\Acoll$ and $m_{\Bcoll}$ texts in collection $\Bcoll$.
+The \emph{root} of the trie corresponds to suffix $\$$, which has lexicographic rank $m_{\Acoll}$ in $\Acoll$ and corresponds to lexicographic range $[1,m_{\Bcoll}]$ in $\Bcoll$.
+
+% FIXME pseudocode if we have space for it
+Assume that we have a \emph{node} of the trie corresponding to suffix $X$, lexicographic rank $r$, and lexicographic range $[sp,ep]$.
+This means that suffix $X$ occurs $sp+1-ep$ times in collection $\Bcoll$, so we can output a \emph{run} of lexicographic ranks $(r, sp+1-ep)$.
+Afterwards, we proceed to the \emph{children} of the node.
+For each character $c \in \Sigma$, we create a node corresponding to suffix $cX$, lexicographic rank $\mLF_{\Acoll}(r,c)$, and lexicographic range $[\mLF_{\Bcoll}(sp-1, c) + 1, \mLF_{\Bcoll}(ep, c)]$.
+Searching the branches of the trie can be done in parallel using multiple \emph{threads}.
+
+\SubSection{Buffering}
+
+To reduce disk usage and the amount of disk I/O, we buffer and compress the lexicographic ranks before writing them to disk.
+Each thread has two buffers: a \emph{run buffer} and a \emph{thread buffer}.
+The run buffer stores the runs as pairs of integers $(r, \ell)$.
+Once the run buffer becomes full, we sort the runs by \emph{run heads} $r$, use \emph{differential encoding} for the run heads, and encode the differences and run lengths with a \emph{prefix-free code}.
+The compressed run buffer is then merged with the similarly compressed thread buffer.
+
+Once a thread buffer becomes full, we insert it into the global \emph{merge buffers}.
+There are $k$ merge buffers, with buffer $i$ containing $2^{i-1}$ merged thread buffers.
+The insertion starts at buffer $1$.
+If buffer $i$ is empty, the thread inserts its thread buffer there and returns back to the search phase.
+Otherwise it takes the contents of merge buffer $i$, merges them with its thread buffer, and proceeds to buffer $i+1$.
+If a thread reaches buffer $k+1$, it writes its thread buffer to disk and returns back to work.
+
+\SubSection{Merge}
+
+The merge phase starts with the rank array stored in multiple files on disk.
+Each of the files is sorted and compressed.
+For interleaving the \BWT{}s, we need to merge the files to produce an increasing run-length encoded stream of lexicographic ranks.
+
+This work can also be parallelized between multiple threads.
+One thread reads the files and performs a \emph{multiway merge} using a priority queue, producing a stream of lexicographic ranks.
+Another thread consumes the stream and uses it to interleave the \BWT{}s.
+Because the producer thread is usually slower than the consumer thread, we may want to use multiple threads for the multiway merge, if the disk is fast enough.
 
@@
 
-details of run-length encoding of BWT, RA
+details of run-length encoding of BWT, RA; search optimizations
 

### 2015-10-22 paper/paper.tex
Implementation.

@@
 \newcommand{\Bcoll}{\ensuremath{\mathcal{B}}}
+
+% Names
+\newcommand{\BWTmerge}{\textsf{BWT\nobreakdash-merge}}
+\newcommand{\sdarray}{\textsf{sdarray}}
 
@@
 
-C++, compiler, github
+We have implemented the proposed enhancements to the \BWT{} merging algorithm in a tool (\BWTmerge) intended for merging the \BWT{}s of large collections of short reads.
+\BWTmerge{} is written in C++, and the source code is available at GitHub.\footnote{\url{https://github.com/jltsiren/bwt-merge}} The implementation is built on top of the \emph{SDSL library} \cite{Gog2014b} and uses the features of C++11 extensively.
+As a result, it needs a fairly recent C++11 compiler to compile.
+We have built \BWTmerge{} on Linux and OS~X using g++.
 
-details of run-length encoding of BWT, RA; search optimizations
+The target environment of \BWTmerge{} is a \emph{single node} of a \emph{computer cluster}.
+The system should have tens of CPU cores and hundreds of gigabytes of memory.
+The amount of local disk space might not be much larger than memory size, while there can be plenty of shared disk space available.
+The number of search threads is equal to the number of CPU cores, while the merge phase uses just one producer thread and one consumer thread.
+By adjusting the sizes of run buffers and thread buffers and the number of merge buffers, \BWTmerge{} should work reasonably well in different environments.
 
-number of threads
+The internal alphabet of \BWTmerge{} is \texttt{012345}, which corresponds to either \texttt{\$ACGTN} or \texttt{\$ACGNT}, depending on where the \BWT{}s come from.
+\BWT{}s using different alphabetic orders cannot be merged.
+We use simple byte-level codes for \emph{run-length encoding} the \BWT{}s.
+The encoding of run $(c, \ell)$, where $c$ is the character value and $\ell$ is the length, depends on the length of the run.
+If $\ell \le 41$, the run is encoded in a single byte as $6 \cdot (\ell-1) + c$.
+Longer runs start with byte $6 \cdot 41 + c$, followed by the encoding of $\ell-42$.
+The remaining run length is encoded as a sequence of bytes, with the low 7 bits containing data and the high bit telling whether the encoding continues in the next byte.
+The compressed buffers use the same 7+1\nobreakdash-bit code for both differentially encoded run heads and run lengths.
 
-use 8~MB blocks instead of virtual memory pages
+For \rank/\select{} support, we split the \BWT{}s into 64\nobreakdash-byte blocks of compressed data, so that no run continues from one block to the next one.
+For each block $i$, we store the total number of characters in blocks $1$ to $i-1$ as $n_{i}$, as well as the cumulative character counts $c_{i} = \mBWT.\mrank(n_{i},c)$ for $0 \le c \le 5$.
+Each of these seven increasing sequences is encoded using the \sdarray{} encoding \cite{Okanohara2007}.
+To compute $\mBWT.\mrank(j,c)$, we first use a \rank{} query on the $n_{i}$ sequence to find the correct block.
+A \select{} query on the same sequence transforms position $j$ into an offset within the block, while a \select{} query on the $c_{i}$ sequence gives the rank up to the beginning of the block.
+Finally, we decompress the block to answer the query.
+\select{} queries and random accesses to the \BWT{} work in a similar way.
+There are also optimizations for e.g.~computing $\mrank(i,c)$ for all characters $c$, and for finding the children of a reverse trie node corresponding to a relatively short lexicographic range.
+
+We use two-level arrays with 8\nobreakdash-megabyte blocks to store the \BWT{}s and the compressed buffers, managing the blocks using \texttt{mmap()} and \texttt{munmap()}.
+In our experiments, this direct memory management reduced the space overhead by tens of gigabytes.
+When multiple threads allocate memory in small enough blocks, the multithreaded \emph{glibc} implementation of \texttt{malloc()} creates a number of additional heaps that will never grow larger than 64~MB.
+Each thread tries to reuse the heap it used for the last allocation.
+If a heap is full or the thread cannot acquire the mutex, it moves to the next heap.
+With our workload of tens of threads allocating and freeing hundreds of gigabytes of memory, this created thousands of heaps with holes in the middle.
+Most calls to \texttt{free()} could not release the memory back to the operating system, while it took a while before a thread reached the heap where the \texttt{free()} took place.
 
@@
 
-environment
+environment, compiler
 
@@
 
-sequence ordering: original orderin may be important, RLO/RCLO for compression, RLCO for mapping between reverse complements, mapping position for encoding pairing information
+sequence ordering: original order may be important, RLO/RCLO for compression, RLCO for mapping between reverse complements, mapping position for encoding pairing information
 

### 2015-10-22 paper/paper.tex
Multiple sections.

@@
 \newcommand{\sdarray}{\textsf{sdarray}}
+\newcommand{\CEU}{\textsf{CEU}}
+\newcommand{\RS}{\textsf{RS}}
 
@@
 \begin{abstract}
-Fooo
+In order to avoid the reference bias introduced by mapping reads to a reference genome, bioinformaticians are investigating reference-free methods for analyzing sequenced genomes.
+With large projects sequencing thousands of individuals, this raises the need for tools capable of processing terabases of sequence data.
+A key method is the Burrows-Wheeler transform (BWT), which is widely used for compressing and indexing reads.
+We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of smaller subcollections.
+The algorithm is capable of merging over 700 Gbp/day on a single system, with 20--25 gigabytes of memory overhead on top of the run-length encoded BWTs.
 \end{abstract}
@@
 
-\SubSection{Search}
-
-Instead of searching for every text in collection $\Bcoll$ separately, we can search for the \emph{reverse trie} of the collection.
+\smallbreak\noindent\textbf{Search.} Instead of searching for every text in collection $\Bcoll$ separately, we can search for the \emph{reverse trie} of the collection.
 Assume that there are $m_{\Acoll}$ texts in collection $\Acoll$ and $m_{\Bcoll}$ texts in collection $\Bcoll$.
@@
 
-\SubSection{Buffering}
-
-To reduce disk usage and the amount of disk I/O, we buffer and compress the lexicographic ranks before writing them to disk.
+\smallbreak\noindent\textbf{Buffering.} To reduce disk usage and the amount of disk I/O, we buffer and compress the lexicographic ranks before writing them to disk.
 Each thread has two buffers: a \emph{run buffer} and a \emph{thread buffer}.
@@
 
-\SubSection{Merge}
-
-The merge phase starts with the rank array stored in multiple files on disk.
+\smallbreak\noindent\textbf{Merge.} The merge phase starts with the rank array stored in multiple files on disk.
 Each of the files is sorted and compressed.
 For interleaving the \BWT{}s, we need to merge the files to produce an increasing run-length encoded stream of lexicographic ranks.
-
-This work can also be parallelized between multiple threads.
+We can also use multiple threads.
 One thread reads the files and performs a \emph{multiway merge} using a priority queue, producing a stream of lexicographic ranks.
@@
 The remaining run length is encoded as a sequence of bytes, with the low 7 bits containing data and the high bit telling whether the encoding continues in the next byte.
-The compressed buffers use the same 7+1\nobreakdash-bit code for both differentially encoded run heads and run lengths.
+The compressed buffers use the same 7+1\nobreakdash-bit code for both the differentially encoded run heads and the run lengths.
 
@@
 
-environment, compiler
+We ran our experiments on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256 gigabytes of memory.
+The system was running Ubuntu 12.04 on Linux kernel 3.2.0.
+All software was compiled with gcc/g++ version 4.9.2.
 
-data: ReadServer BWTs, NA12878 trio \cite{1000GP2015}
-error correction using bfc \cite{Li2015}
+\begin{table}[t!]
+\begin{center}
+\caption{Datasets used in the experiments.}\label{table:datasets}\smallskip%
+{
+\renewcommand{\baselinestretch}{1}\footnotesize
+\begin{tabular}{cc|ccc}
+\hline
+\textbf{Dataset} & \textbf{Collection} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} \\
+\hline
+\CEU             & NA12878             &       284 Gbp &          2.81G &               101 bp \\
+\CEU             & NA12891             &       242 Gbp &          2.40G &               101 bp \\
+\CEU             & NA12892             &       245 Gbp &          2.42G &               101 bp \\
+\CEU             & Merged              &       771 Gbp &          7.63G &               101 bp \\
+\hline
+\RS              & AA                  &       433 Gbp &          4.69G &         73 or 100 bp \\
+\RS              & TT                  &       432 Gbp &          4.68G &         73 or 100 bp \\
+\RS              & AT                  &       275 Gbp &          2.98G &         73 or 100 bp \\
+\RS              & TA                  &       355 Gbp &          3.84G &         73 or 100 bp \\
+\RS              & Merged              &      1.49 Tbp &          16.2G &         73 or 100 bp \\
+\hline
+\end{tabular}}
+\end{center}
+\end{table}
 
-ReadServer: trade-offs with different parameters
+We used two datasets from the phase 3 of the \emph{1000 Genomes Project} \cite{1000GP2015}:
+\begin{itemize}
 
-trio: ropebwt + this vs ropebwt vs ropebwt2
+\item \CEU{} contains reads from high-coverage sequencing of the \emph{CEU trio} (individuals NA12878, NA12891, and NA12892).
+We downloaded the gzipped FASTQ files (run accessions SRR622457, SRR622458, and SRR622459) and concatenated the files representing the same individual.
+Then we corrected the sequencing errors with BFC \cite{Li2015} (\texttt{bfc -s 3g -t 16}) separately for each individual.
+
+\item \RS{} is from the ReadServer project, which uses all sequence data from the phase 3.
+After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp.
+The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} \cite{Simpson2012}, distributed according to the last two bases.
+We chose four of the files (AA, TT, AT, and TA).
+
+\end{itemize}
+See Table~\ref{table:datasets} for further details on the datasets.
+
+CEU: ropebwt + this vs ropebwt vs ropebwt2
+
+RS: trade-offs with different parameters
+
+Note: input/output formats, BWT-merge times include verification
 

### 2015-10-25 paper/paper.tex
Introduction.

@@
 
-Sequence data
+The dramatic decrease in the cost of DNA sequencing has flooded the world with \emph{sequence data}.
+The \emph{1000 Genomes Project} \cite{1000GP2015} sequenced the genomes of over 2500 humans, and there are many other projects that are similar or greater in scale.
+For each sequenced genome, a sequencing machine produces a large number of \emph{reads} (short sequences) that cover the genome many times over.
+For a \~3~Gbp human genome, the total length of the reads is often \~100~Gbp or more.
 
-BWT, compression, FM-index, applications, repetitive, rle
+\emph{De novo assembly} of sequenced genomes is still too difficult to be routinely done.
+As a practical alternative, bioinformaticians typically align the reads are aligned to a \emph{reference genome} of the same species.
+Because most reference genomes have been derived from the genomes of a small number of individuals, this introduces \emph{reference bias}, which may adversely affect the results of subsequent analysis.
+Switching from reference sequences to \emph{reference graphs} can help to reduce the bias, but such transition will likely take years \cite{Church2015}.
 
-ReadServer; multiple BWTs due to construction issues; less BWTs would mean less memory, less querying
+The \emph{ReadServer project} at the Wellcome Trust Sanger Institute aims to avoid the reference bias entirely by developing tools for large-scale \emph{reference-free} genome analysis.
+Unique reads are compressed and indexed using the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994}, while metadata databases contain information about the original reads.
+In the pilot phase, the ReadServer project works with the low-coverage and exome data from phase 3 of the 1000 Genomes Project.
+After error correction and trimming the reads to either 73~bp or 100~bp, the 86~Tbp of raw data consisting of 922~billion original reads is reduced to 4.88~Tbp and 53.0~billion unique sequences.
+These sequences are stored in 16 \BWT-based indexes \cite{Ferragina2005a} taking a total of 561.5~gigabytes.
 
-Large scale construction: speed, space, hardware, efficiency
+Preprocessing large datasets can take days or even weeks.
+As a result, it is not always feasible to rebuild everything when new methods of analysis require new functionalities from the indexes.
+\BWT-based methods were chosen due to their versatility.
+A \emph{run-length encoded} \BWT{} compresses repetitive sequence collections quite well \cite{Maekinen2010}, while the similarities to the suffix tree and the suffix array make \BWT-based indexes suitable for a wide variety of \emph{pattern matching} and \emph{sequence analysis} tasks \cite{Ohlebusch2013,Maekinen2015}.
 
-BWT construction: from SA \cite{Mori2008,Nong2011}
-external SA \cite{Gonnet1992,Bingmann2013,Kaerkkaeinen2014a,Nong2014,Kaerkkaeinen2015a}
-space efficient algorithms \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}
-external BWT \cite{Ferragina2012,Beller2013}
+ReadServer stores the unique reads in 16 \BWT-based indexes according to the last two bases of the read.
+This means that every query must be repeated 16 times, while the indexes require more space, as the compression cannot take advantage of the similarities between the reads in different indexes.
+Reducing the number of indexes would improve both index size and performance.
+However, the lack of \BWT{} construction algorithms that work with terabases of sequence data has made it infeasible so far.
 
-Specialized for DNA: \cite{Bauer2013,Liu2014,Li2014a,Pantaleoni2014,Wang2015}
+There are four often contradictory requirements for large-scale \BWT{} construction:
+\begin{itemize}
 
-This paper: terabases
+\item \textbf{Speed.} Larger datasets require faster algorithms. As a rough guideline, an algorithm processing 1~Mbp/s is good for up to 100~Gbp, while remaining somewhat useful until 1~Tbp of data.
+
+\item \textbf{Memory.} We may have to process $n$~bp datasets on systems with less than $n$~bits of memory.
+
+\item \textbf{Hardware.} A single node in a computer cluster typically has tens of CPU cores, hundreds of gigabytes of memory, a limited amount of local disk space, and access to a large amount of shared disk space with no performance guarantees.
+Algorithms using a GPU or a large amount of fast disk space require special-purpose hardware.
+
+\item \textbf{Efficiency.} Large \BWT{}s can be built by doing a lot of redundant work on multiple nodes.
+As most computer clusters do not have large amounts of unused capacity, such inefficient algorithms are unsuitable for repeated use.
+
+\end{itemize}
+
+The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} for the sequences using a fast general-purpose algorithm \cite{Mori2008,Nong2011}, and then transform the suffix array into the \BWT.
+These algorithms cannot be used with large datasets, as they require several times more memory than the size of the dataset.
+Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index more than 1\nobreakdash--2~Mbp/s \cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Kaerkkaeinen2015a}.
+
+There are several algorithms for constructing the \BWT{} \emph{directly} without having to build the suffix array first.
+They typically require only a limited amount of working space on top of a plain or compressed representation of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009} or use the disk as additional working space \cite{Ferragina2012,Beller2013}.
+General-purpose implementations of these algorithms rarely exceed 1\nobreakdash--2~Mbp/s with large datasets.
+
+\emph{Specialized algorithms} for DNA sequences are much faster.
+Some algorithms achieve 5\nobreakdash--10~Mbps on general-purpose hardware, with their memory usage becoming a major bottleneck for datasets larger than about 1~Tbp \cite{Bauer2013,Li2014a}.
+GPU-based algorithms can be much faster, but their memory usage is also greater \cite{Liu2014,Pantaleoni2014}.
+Distributing the \BWT{} construction on multiple nodes can remove the obvious bottlenecks, but such algorithms generally do not use the hardware as efficiently as those running on  a single system \cite{Wang2015}.
+
+In this paper, we propose a practical algorithm for building the \BWT{} for terabases of sequence data.
+The algorithm is based on splitting the sequence collection into a number of smaller subcollections, building the \BWT{} for each subcollection, and \emph{merging} the \BWT{}s into a single structure \cite{Siren2009}.
+The merging algorithm is faster than \BWT{} construction for the subcollections, while having a relatively small memory overhead on top of the final \BWT-based index.
+As the index must be loaded in memory before it can be used, index construction can be done on the same system as the index is used.
 
@@
 All software was compiled with gcc/g++ version 4.9.2.
+We stored the input/output files on a distributed Lustre file system and used a local 0.5~TB hard disk for temporary files.
 
@@
 
-\item \RS{} is from the ReadServer project, which uses all sequence data from the phase 3.
+\item \RS{} is from the ReadServer project, which uses all low-coverage and exome data from the phase 3.
 After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp.

### 2015-10-26 paper/paper.tex
More experiments; better default parameters.

@@
 \newcommand{\BWTmerge}{\textsf{BWT\nobreakdash-merge}}
-\newcommand{\sdarray}{\textsf{sdarray}}
+\newcommand{\ropebwt}{\textsf{RopeBWT}}
+\newcommand{\ropebwtii}{\textsf{RopeBWT2}}
 \newcommand{\CEU}{\textsf{CEU}}
@@
 \author{%
-Jouni Sirén \\[0.5em]
+Jouni Sirén\thanks{This work was supported by the Wellcome Trust grant [098051].} \\[0.5em]
 {\small\begin{minipage}{\linewidth}\begin{center}
@@
 For each sequenced genome, a sequencing machine produces a large number of \emph{reads} (short sequences) that cover the genome many times over.
-For a \~3~Gbp human genome, the total length of the reads is often \~100~Gbp or more.
+For a 3~Gbp human genome, the total length of the reads is often 100~Gbp or more.
 
@@
 For each block $i$, we store the total number of characters in blocks $1$ to $i-1$ as $n_{i}$, as well as the cumulative character counts $c_{i} = \mBWT.\mrank(n_{i},c)$ for $0 \le c \le 5$.
-Each of these seven increasing sequences is encoded using the \sdarray{} encoding \cite{Okanohara2007}.
+Each of these seven increasing sequences is encoded using the \emph{sdarray} encoding \cite{Okanohara2007}.
 To compute $\mBWT.\mrank(j,c)$, we first use a \rank{} query on the $n_{i}$ sequence to find the correct block.
@@
 We stored the input/output files on a distributed Lustre file system and used a local 0.5~TB hard disk for temporary files.
+\BWTmerge{} was allowed to use 32 threads, while the other \BWT{} construction tools were limited to 4 or 5 parallel threads by design.
+All reported merging times include verification by querying the \BWT{}s with 2~million $32$\nobreakdash-mers extracted from the human reference genome.
 
@@
 \hline
-\textbf{Dataset} & \textbf{Collection} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} \\
+\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} \\
 \hline
 \CEU & NA12878 &  284 Gbp & 2.81G &       101 bp \\
-\CEU             & NA12891             &       242 Gbp &          2.40G &               101 bp \\
-\CEU             & NA12892             &       245 Gbp &          2.42G &               101 bp \\
-\CEU             & Merged              &       771 Gbp &          7.63G &               101 bp \\
+     & NA12891 &  242 Gbp & 2.40G &       101 bp \\
+     & NA12892 &  245 Gbp & 2.42G &       101 bp \\
+     & Merged  &  771 Gbp & 7.63G &       101 bp \\
 \hline
 \RS  & AA      &  433 Gbp & 4.69G & 73 or 100 bp \\
-\RS              & TT                  &       432 Gbp &          4.68G &         73 or 100 bp \\
-\RS              & AT                  &       275 Gbp &          2.98G &         73 or 100 bp \\
-\RS              & TA                  &       355 Gbp &          3.84G &         73 or 100 bp \\
-\RS              & Merged              &      1.49 Tbp &          16.2G &         73 or 100 bp \\
+     & TT      &  432 Gbp & 4.68G & 73 or 100 bp \\
+     & AT      &  275 Gbp & 2.98G & 73 or 100 bp \\
+     & TA      &  355 Gbp & 3.84G & 73 or 100 bp \\
+     & Merged  & 1.49 Tbp & 16.2G & 73 or 100 bp \\
+\hline
+\RS  & *A, *C  & 2.45 Tbp & 26.5G & 73 or 100 bp \\
+     & *G, *T  & 2.44 Tbp & 26.5G & 73 or 100 bp \\
 \hline
@@
 
-\item \RS{} is from the ReadServer project, which uses all low-coverage and exome data from the phase 3.
+\item \RS{} is from the \emph{ReadServer project}, which uses all low-coverage and exome data from the phase 3.
 After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp.
 The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} \cite{Simpson2012}, distributed according to the last two bases.
-We chose four of the files (AA, TT, AT, and TA).
 
@@
 
-CEU: ropebwt + this vs ropebwt vs ropebwt2
+\smallbreak\noindent\textbf{Benchmarking.} For benchmarking with different parameter values, we converted four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset to the native format used by \BWTmerge.
+Then we merged the \BWT{}s (in the given order).
+We used 128~MB or 256~MB run buffers and 256~MB or 512~MB thread buffers.
+The number of merge buffers was 4 or 5 with 512~MB thread buffers and 5 or 6 with 256~MB thread buffers, so that the files on disk were always merged from either 8~GB or 16~GB of thread buffers.
 
-RS: trade-offs with different parameters
+\begin{figure}[t!]
+\begin{center}
+\includegraphics{benchmark.pdf}%
+\hspace{-0.3in}%
+\includegraphics{comparison.pdf}
+\end{center}
+\caption{Left: Time/space trade-offs for merging four \BWT{} files containing a total of 1.49~Tbp.
+Label rXbYmZ denotes $X$~MB run buffers, $Y$~MB threads buffers, and $Z$ merge buffers.
+The dashed line represents the total size of the last pair of \BWT{}s to be merged.
+Right: Time/space trade-offs for building the \BWT{} of the 771~Gbp \CEU{} dataset.}\label{fig:benchmark}
+\end{figure}
 
-Note: input/output formats, BWT-merge times include verification
+The results of the benchmarks can be seen in Figure~\ref{fig:benchmark} (left).
+Merging took 31.4 to 35.7~hours and required 145 to 166~gigabytes of memory, depending on the parameter values.
+This corresponds to an average speed of 8.27 to 9.40~Mbp/s for inserting 1.06~Tbp into the initial file AA.
+As the total size of the last pair of \BWT{}s was 124.2~gigabytes, the memory overhead was 21.1 to 41.5~gigabytes.
+The temporary files required 287 to 306~gigabytes of disk space.
+For the further experiments, we chose 128\nobreakdash-megabyte run buffers and 256\nobreakdash-megabyte thread buffers.
+We use 5 merge buffers in the \emph{small} variant (memory overhead 21.1~GB) and 6 merge buffers in the \emph{fast} variant (overhead 30.8~GB).
+
+\smallbreak\noindent\textbf{Comparison.} In the next experiment, we compared the performance of \BWTmerge{} to the \ropebwt{} \cite{Li2011-2013} implementation of the BCR algorithm \cite{Bauer2013} and to \ropebwtii{} \cite{Li2014a}.
+These two tools are the fastest existing \BWT{} construction tools for read collections on general hardware \cite{Li2014a}.
+We ran \ropebwt{} with parameters \texttt{-btORf -abcr} and \ropebwtii{} with parameters \texttt{-bRm10g}.
+Due to memory constraints, we used the smaller \CEU{} dataset.
+
+We built the \BWT{} of the dataset in four ways.
+We concatenated all input files, and built the \BWT{} directly for the entire collection using both \ropebwt{} and \ropebwtii{}.
+We built the \BWT{} for each input file separately using \ropebwt{}, and then merged the results using \BWTmerge{} with small settings.
+Finally, we built the individual \BWT{}s using parallel \ropebwt{} jobs, and then merged them using \BWTmerge{} with fast settings.
+In each case, the tools were allowed to output the \BWT{} files in their preferred formats.
+
+The results can be seen in Figure~\ref{fig:benchmark} (right).
+% FIXME discussion of the results; note that ropebwt2 did not finish
+
+\smallbreak\noindent\textbf{ReadServer.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files using the fast settings.
+% FIXME sizes in native/sga format, merging time/space
 
@@
 
-conclusions
+conclusions: fast, low memory overhead. because BWTs are in-memory data structures, they can be built on the same system they are going to be used
 
-sequence ordering: original order may be important, RLO/RCLO for compression, RLCO for mapping between reverse complements, mapping position for encoding pairing information
+sequence ordering: original order may be important, RLO/RCLO for compression, RLCO for mapping between reverse complements, mapping position for encoding pairing information; ReadServer currently uses RLO
 

### 2015-10-26 fmi.h

@@ -47,8 +47,8 @@ struct MergeParameters
   typedef typename RLArray<BlockArray>::run_type run_type;
 
   const static size_type RUN_BUFFER_SIZE = 8 * MEGABYTE;      // Runs.
-  const static size_type THREAD_BUFFER_SIZE = 512 * MEGABYTE; // Bytes.
-  const static size_type MERGE_BUFFERS = 5;
+  const static size_type THREAD_BUFFER_SIZE = 256 * MEGABYTE; // Bytes.
+  const static size_type MERGE_BUFFERS = 6;
   const static size_type BLOCKS_PER_THREAD = 4;
 
   const static std::string DEFAULT_TEMP_DIR;  // .

### 2015-10-26 paper/benchmark.R

@@ -1,0 +1,55 @@
+# Use R --slave --args name < benchmark.R
+
+args = commandArgs()
+
+name = args[4]
+
+x = 3.3
+y = 3
+
+data <- read.csv(file = paste(name, ".csv", sep = ""), head = FALSE, sep = ",", dec = ".", check.names = FALSE)
+pdf(file = paste(name, ".pdf", sep = ""), width = x, height = y, paper = "special",
+  family = "Helvetica", pointsize = 11)
+par(mar=c(4, 4, 1, 1))
+
+xrange = c(120, 170)
+xscale = c(120, 130, 140, 150, 160, 170)
+xtitle = "Memory usage (GB)"
+xlabs = xscale
+
+yrange = c(30, 36)
+yscale = c(30, 31, 32, 33, 34, 35, 36)
+ytitle = "Time (h)"
+ylabs = yscale
+
+plot(c(1),
+  c(1),
+  type = "n",
+  axes = F,
+  main = "",
+  xlab = xtitle,
+  ylab = ytitle,
+  xlim = xrange,
+  ylim = yrange)
+
+axis(1, at = xscale, lab = xlabs, cex.axis = 0.8)
+axis(2, at = yscale, lab = ylabs, cex.axis = 0.8)
+box()
+
+nr = nrow(data)
+nc = ncol(data)
+
+points(data[1:nr, 3], data[1:nr, 2] / 3600, type = "p", pch = 20)
+
+text(data[1, 3], data[1, 2] / 3600, data[1, 1], cex = 0.8, pos = 2)
+text(data[2, 3], data[2, 2] / 3600, data[2, 1], cex = 0.8, pos = 4)
+text(data[3, 3], data[3, 2] / 3600, data[3, 1], cex = 0.8, pos = 2)
+text(data[4, 3], data[4, 2] / 3600, data[4, 1], cex = 0.8, pos = 2)
+text(data[5, 3], data[5, 2] / 3600, data[5, 1], cex = 0.8, adj = c(0.65, 2))
+text(data[6, 3], data[6, 2] / 3600, data[6, 1], cex = 0.8, pos = 2)
+text(data[7, 3], data[7, 2] / 3600, data[7, 1], cex = 0.8, pos = 2)
+
+segments(124.2, 29, 124.2, 37, lty = "dashed")
+
+dev.off()
+q()

### 2015-10-26 paper/comparison.R

@@ -1,0 +1,48 @@
+# Use R --slave --args name < comparison.R
+
+args = commandArgs()
+
+name = args[4]
+
+x = 3
+y = 3
+
+data <- read.csv(file = paste(name, ".csv", sep = ""), head = FALSE, sep = ",", dec = ".", check.names = FALSE)
+pdf(file = paste(name, ".pdf", sep = ""), width = x, height = y, paper = "special",
+  family = "Helvetica", pointsize = 11)
+par(mar=c(4, 4, 1, 1))
+
+xrange = c(0, 256)
+xscale = c(0, 64, 128, 192, 256)
+xtitle = "Memory usage (GB)"
+xlabs = xscale
+
+yrange = c(0, 60)
+yscale = c(0, 12, 24, 36, 48, 60)
+ytitle = ""
+ylabs = yscale
+
+plot(c(1),
+  c(1),
+  type = "n",
+  axes = F,
+  main = "",
+  xlab = xtitle,
+  ylab = ytitle,
+  xlim = xrange,
+  ylim = yrange)
+
+axis(1, at = xscale, lab = xlabs, cex.axis = 0.8)
+axis(2, at = yscale, lab = ylabs, cex.axis = 0.8)
+box()
+
+nr = nrow(data)
+nc = ncol(data)
+
+points(data[1:nr, 3], data[1:nr, 2] / 3600, type = "p", pch = 20)
+
+text(data[1, 3], data[1, 2] / 3600, data[1, 1], cex = 0.8, pos = 2)
+text(data[2, 3], data[2, 2] / 3600, data[2, 1], cex = 0.8, pos = 2)
+
+dev.off()
+q()

### 2015-10-26 paper/paper.tex
Verification that the format exists.

@@
 
-We have implemented the proposed enhancements to the \BWT{} merging algorithm in a tool (\BWTmerge) intended for merging the \BWT{}s of large collections of short reads.
-\BWTmerge{} is written in C++, and the source code is available at GitHub.\footnote{\url{https://github.com/jltsiren/bwt-merge}} The implementation is built on top of the \emph{SDSL library} \cite{Gog2014b} and uses the features of C++11 extensively.
+We have implemented the improved \BWT{} merging algorithm as a tool for merging the \BWT{}s of large collections of short reads.
+The tool, \BWTmerge{}, is written in C++, and the source code is available at GitHub.\footnote{\url{https://github.com/jltsiren/bwt-merge}} The implementation is built on top of the \emph{SDSL library} \cite{Gog2014b} and uses the features of C++11 extensively.
 As a result, it needs a fairly recent C++11 compiler to compile.
-We have built \BWTmerge{} on Linux and OS~X using g++.
+We have successfully built \BWTmerge{} on Linux and OS~X using g++.
 
@@
 See Table~\ref{table:datasets} for further details on the datasets.
+We used a development version of \BWTmerge{} that was essentially equivalent to v0.3 for the experiments.
+For the other tools, we used the versions that were available on GitHub in October~2015.
 
-\smallbreak\noindent\textbf{Benchmarking.} For benchmarking with different parameter values, we converted four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset to the native format used by \BWTmerge.
-Then we merged the \BWT{}s (in the given order).
+\smallbreak\noindent\textbf{Benchmarking.} For benchmarking with different parameter values, we converted four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset to the \emph{native format} of \BWTmerge.
+This format includes the \BWT{} and the \rank/\select{} structures required by the FM-index.
+We then merged the \BWT{}s (in the given order).
 We used 128~MB or 256~MB run buffers and 256~MB or 512~MB thread buffers.

### 2015-10-26 bwt_convert.cpp

@@ -56,9 +56,19 @@ main(int argc, char** argv)
     {
     case 'i':
       input_tag = optarg;
+      if(!formatExists(input_tag))
+      {
+        std::cerr << "bwt_convert: Invalid input format: " << input_tag << std::endl;
+        std::exit(EXIT_FAILURE);
+      }
       break;
     case 'o':
       output_tag = optarg;
+      if(!formatExists(output_tag))
+      {
+        std::cerr << "bwt_convert: Invalid output format: " << output_tag << std::endl;
+        std::exit(EXIT_FAILURE);
+      }
       break;
     case '?':
     default:
@@ -66,17 +76,13 @@ main(int argc, char** argv)
     }
   }
 
-  if(optind < argc) { input_name = argv[optind]; }
-  else
-  {
-    std::cerr << "bwt_convert: Input file not specified" << std::endl;
-  }
-  if(optind + 1 < argc) { output_name = argv[optind + 1]; }
-  else
+  if(optind + 1 >= argc)
   {
     std::cerr << "bwt_convert: Output file not specified" << std::endl;
     std::exit(EXIT_FAILURE);
   }
+  input_name = argv[optind];
+  output_name = argv[optind + 1];
 
   std::cout << "Input:   " << input_name << " (" << input_tag << ")" << std::endl;
   std::cout << "Output:  " << output_name << " (" << output_tag << ")" << std::endl;

### 2015-10-26 bwt_merge.cpp

@@ -89,9 +89,22 @@ main(int argc, char** argv)
       break;
     case 'i':
       tokenize(optarg, input_formats, ',');
+      for(size_type i = 0; i < input_formats.size(); i++)
+      {
+        if(!formatExists(input_formats[i]))
+        {
+          std::cerr << "bwt_merge: Invalid input format: " << input_formats[i] << std::endl;
+          std::exit(EXIT_FAILURE);
+        }
+      }
       break;
     case 'o':
       output_format = optarg;
+      if(!formatExists(output_format))
+      {
+        std::cerr << "bwt_merge: Invalid output format: " << output_format << std::endl;
+        std::exit(EXIT_FAILURE);
+      }
       break;
     case '?':
     default:

### 2015-10-26 formats.cpp

@@ -446,6 +446,18 @@ SGAFormat::write(std::ofstream& out, const BlockArray& data, const NativeHeader&
 
 //------------------------------------------------------------------------------
 
+bool
+formatExists(const std::string& format)
+{
+  return (format == NativeFormat::tag)
+      || (format == PlainFormatD::tag)
+      || (format == PlainFormatS::tag)
+      || (format == RFMFormat::tag)
+      || (format == SDSLFormat::tag)
+      || (format == RopeFormat::tag)
+      || (format == SGAFormat::tag);
+}
+
 void
 printFormats(std::ostream& stream)
 {

### 2015-10-26 formats.h

@@ -157,6 +157,8 @@ struct SGAFormat
 
 //------------------------------------------------------------------------------
 
+bool formatExists(const std::string& format);
+
 void printFormats(std::ostream& stream);
 
 template<class Format>

### 2015-10-27 paper/paper.tex
Discussion.

@@
 
-The \emph{ReadServer project} at the Wellcome Trust Sanger Institute aims to avoid the reference bias entirely by developing tools for large-scale \emph{reference-free} genome analysis.
+The \emph{Read Server project} at the Wellcome Trust Sanger Institute aims to avoid the reference bias entirely by developing tools for large-scale \emph{reference-free} genome analysis.
 Unique reads are compressed and indexed using the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994}, while metadata databases contain information about the original reads.
-In the pilot phase, the ReadServer project works with the low-coverage and exome data from phase 3 of the 1000 Genomes Project.
+In the pilot phase, the Read Server project works with the low-coverage and exome data from phase 3 of the 1000 Genomes Project.
 After error correction and trimming the reads to either 73~bp or 100~bp, the 86~Tbp of raw data consisting of 922~billion original reads is reduced to 4.88~Tbp and 53.0~billion unique sequences.
@@
 
-ReadServer stores the unique reads in 16 \BWT-based indexes according to the last two bases of the read.
+Read Server stores the unique reads in 16 \BWT-based indexes according to the last two bases of the read.
 This means that every query must be repeated 16 times, while the indexes require more space, as the compression cannot take advantage of the similarities between the reads in different indexes.
@@
 
-\item \RS{} is from the \emph{ReadServer project}, which uses all low-coverage and exome data from the phase 3.
+\item \RS{} is from the \emph{Read Server project}, which uses all low-coverage and exome data from the phase 3.
 After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp.
@@
 The results of the benchmarks can be seen in Figure~\ref{fig:benchmark} (left).
-Merging took 31.4 to 35.7~hours and required 145 to 166~gigabytes of memory, depending on the parameter values.
-This corresponds to an average speed of 8.27 to 9.40~Mbp/s for inserting 1.06~Tbp into the initial file AA.
-As the total size of the last pair of \BWT{}s was 124.2~gigabytes, the memory overhead was 21.1 to 41.5~gigabytes.
+The average speed for inserting 1.06~Tbp into the initial file AA ranged from 8.27 to 9.40~Mbp/s, depending on the parameter values.
+Memory overhead was from 21.1 to 41.5~GB on top of the 124.2~GB required by the last pair of \BWT{}s.
 The temporary files required 287 to 306~gigabytes of disk space.
-For the further experiments, we chose 128\nobreakdash-megabyte run buffers and 256\nobreakdash-megabyte thread buffers.
+Thread buffer size and the number of merge buffers were the most important parameters.
+The larger the individual files are, the more memory the search phase uses and the faster the merge phase is.
+Increasing run buffer size to 256~MB made the search phase faster with 512~MB thread buffers and slower with 256~MB thread buffers.
+
+For the further experiments, we chose 128~MB run buffers and 256~MB thread buffers.
 We use 5 merge buffers in the \emph{small} variant (memory overhead 21.1~GB) and 6 merge buffers in the \emph{fast} variant (overhead 30.8~GB).
@@
 
-\smallbreak\noindent\textbf{ReadServer.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files using the fast settings.
+\smallbreak\noindent\textbf{Read Server.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files using the fast settings.
 % FIXME sizes in native/sga format, merging time/space
@@
 
-conclusions: fast, low memory overhead. because BWTs are in-memory data structures, they can be built on the same system they are going to be used
+We have proposed an improved \BWT{} merging algorithm for large read collections.
+Our implementation of the algorithm in the \BWTmerge{} tool is fast enough to be used with terabases of sequence data.
+It requires only 20\nobreakdash--30~gigabytes of memory on top of the \BWT{}s to be merged.
+As \BWT-based indexes do random accesses to large arrays, they must reside in memory in most applications.
+Hence \BWTmerge{} can build the index on the same system as it is going to be used.
 
-sequence ordering: original order may be important, RLO/RCLO for compression, RLCO for mapping between reverse complements, mapping position for encoding pairing information; ReadServer currently uses RLO
+\BWTmerge{} can be used as a building block of a \BWT{} construction algorithm.
+We split the read collection into multiple subcollections, build the \BWT{}s of the subcollections with other tools, and merge the results.
+Depending on whether we run one or more \BWT{} construction jobs in parallel, the resulting algorithm can be slower but more space-efficient than the existing algorithms, or faster than them with similar memory usage.
 
-remove duplicates
+The most important feature of our algorithm is its low memory usage.
+With it, we can build the \BWT{}s of much larger read collections than before on commonly available hardware.
+As a concrete example, we merged the 16 Read Server \BWT{} files into two files.
+This reduced the number of servers required to house the indexes from three to two, and also improved the query performance of the servers.
+
+In the future, we are going to extend \BWTmerge{} to support different \emph{text orders}, and to optionally \emph{remove duplicate texts} from the merged collection.
+The current implementation maintains the existing order by inserting the texts from $\mBWT_{\Bcoll}$ after the texts in $\mBWT_{\Acoll}$.
+This makes it easy to determine the original identifiers of the texts without having to store a permutation.
+Other text orders can be useful for different purposes.
+
+Read Server stores the reads in \emph{reverse lexicographic order} to improve compression \cite{Cox2012}.
+We can maintain this order with a few changes to the search phase \cite{Li2014a}.
+Sorting the reads by their \emph{reverse complements} also improves compression in a similar way.
+In this order, $\mSA[\mBWT.\mselect(i,0)]$ points to the beginning of the reverse complement of read $T_{i}$, if the collection includes the reverse complement of each read \cite{Li2014a}.
+
+We can also sort the reads by their likely positions in a reference genome.
+This \emph{position order} is useful, if we also want to store \emph{pairing information} for the reads.
+Consider a graph with reads as its nodes and edges between paired reads.
+If we sort the reads in position order, most edges will be close to the diagonal of the edge matrix.
+Such matrices are very easy to compress.
 

### 2015-10-28 paper/paper.tex
Polishing the paper.

@@
 We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of smaller subcollections.
-The algorithm is capable of merging over 700 Gbp/day on a single system, with 20--25 gigabytes of memory overhead on top of the run-length encoded BWTs.
+The algorithm is capable of merging over 700 Gbp/day on a single system, with 20--30 gigabytes of memory overhead on top of the run-length encoded BWTs.
 \end{abstract}
@@
 
-The \emph{Read Server project} at the Wellcome Trust Sanger Institute aims to avoid the reference bias entirely by developing tools for large-scale \emph{reference-free} genome analysis.
+The \emph{Read Server} project at the Wellcome Trust Sanger Institute aims to avoid the reference bias entirely by developing tools for large-scale \emph{reference-free} genome analysis.
 Unique reads are compressed and indexed using the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994}, while metadata databases contain information about the original reads.
@@
 \begin{center}
-\caption{Datasets used in the experiments.}\label{table:datasets}\smallskip%
+\caption{Datasets used in the experiments.
+\BWT{} sizes are for the native format.
+RLO indicates that the reads are sorted in reverse lexicographic order in each input file.}\label{table:datasets}\smallskip%
 {
 \renewcommand{\baselinestretch}{1}\footnotesize
-\begin{tabular}{cc|ccc}
+\begin{tabular}{cc|ccc|cc}
 \hline
-\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} \\
+ & & & & & \multicolumn{2}{c}{\textbf{\BWT{} size}} \\
+\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} & \textbf{Unsorted} & \textbf{RLO} \\
 \hline
-\CEU & NA12878 &  284 Gbp & 2.81G &       101 bp \\
-     & NA12891 &  242 Gbp & 2.40G &       101 bp \\
-     & NA12892 &  245 Gbp & 2.42G &       101 bp \\
-     & Merged  &  771 Gbp & 7.63G &       101 bp \\
+\CEU & NA12878 &  284 Gbp & 2.81G &       101 bp & 50.3 GB \\
+     & NA12891 &  242 Gbp & 2.40G &       101 bp & 42.4 GB \\
+     & NA12892 &  245 Gbp & 2.42G &       101 bp & 43.8 GB \\
+     & Merged  &  771 Gbp & 7.63G &       101 bp & \\
 \hline
-\RS  & AA      &  433 Gbp & 4.69G & 73 or 100 bp \\
-     & TT      &  432 Gbp & 4.68G & 73 or 100 bp \\
-     & AT      &  275 Gbp & 2.98G & 73 or 100 bp \\
-     & TA      &  355 Gbp & 3.84G & 73 or 100 bp \\
-     & Merged  & 1.49 Tbp & 16.2G & 73 or 100 bp \\
+\RS  & AA      &  433 Gbp & 4.69G & 73 or 100 bp &      -- & 38.5 GB \\
+     & TT      &  432 Gbp & 4.68G & 73 or 100 bp &      -- & 38.7 GB \\
+     & AT      &  275 Gbp & 2.98G & 73 or 100 bp &      -- & 26.6 GB \\
+     & TA      &  355 Gbp & 3.84G & 73 or 100 bp &      -- & 32.7 GB \\
+     & Merged  & 1.49 Tbp & 16.2G & 73 or 100 bp &      -- &  117 GB \\
 \hline
-\RS  & *A, *C  & 2.45 Tbp & 26.5G & 73 or 100 bp \\
-     & *G, *T  & 2.44 Tbp & 26.5G & 73 or 100 bp \\
+\RS  & *A, *C  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- &  225 GB \\
+     & Merged  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- & \\
+\hline
+\RS  & *G, *T  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- &  226 GB \\
+     & Merged  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- & \\
 \hline
@@
 
-\item \RS{} is from the \emph{Read Server project}, which uses all low-coverage and exome data from the phase 3.
+\item \RS{} is from the \emph{Read Server} project, which uses all low-coverage and exome data from the phase 3.
 After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp.
@@
 
-\smallbreak\noindent\textbf{Benchmarking.} For benchmarking with different parameter values, we converted four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset to the \emph{native format} of \BWTmerge.
+\smallbreak\noindent\textbf{Parameters.} For benchmarking with different parameter values, we converted four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset to the \emph{native format} of \BWTmerge.
 This format includes the \BWT{} and the \rank/\select{} structures required by the FM-index.
@@
 The dashed line represents the total size of the last pair of \BWT{}s to be merged.
-Right: Time/space trade-offs for building the \BWT{} of the 771~Gbp \CEU{} dataset.}\label{fig:benchmark}
+Right: Time/space trade-offs for building the \BWT{} of the 771~Gbp \CEU{} dataset.
+RLO indicates reverse lexicographic order.}\label{fig:benchmark}
 \end{figure}
@@
 
-For the further experiments, we chose 128~MB run buffers and 256~MB thread buffers.
-We use 5 merge buffers in the \emph{small} variant (memory overhead 21.1~GB) and 6 merge buffers in the \emph{fast} variant (overhead 30.8~GB).
+For the further experiments, we chose 128~MB run buffers, and 256~MB thread buffers and 6 merge buffers (overhead 30.8~GB).
 
@@
 We ran \ropebwt{} with parameters \texttt{-btORf -abcr} and \ropebwtii{} with parameters \texttt{-bRm10g}.
+For some runs, we also used \texttt{-s} with both tools to sort the reads in \emph{reverse lexicographic order} (RLO) to improve compression.
 Due to memory constraints, we used the smaller \CEU{} dataset.
 
-We built the \BWT{} of the dataset in four ways.
+We built the \BWT{} of the dataset in several ways.
 We concatenated all input files, and built the \BWT{} directly for the entire collection using both \ropebwt{} and \ropebwtii{}.
-We built the \BWT{} for each input file separately using \ropebwt{}, and then merged the results using \BWTmerge{} with small settings.
-Finally, we built the individual \BWT{}s using parallel \ropebwt{} jobs, and then merged them using \BWTmerge{} with fast settings.
+We also built the \BWT{} for each input file separately using \ropebwt{}, and then merged the results using \BWTmerge{}.
+Finally, we repeated everything with the reads sorted in reverse lexicographic order.
 In each case, the tools were allowed to output the \BWT{} files in their preferred formats.
@@
 The results can be seen in Figure~\ref{fig:benchmark} (right).
-% FIXME discussion of the results; note that ropebwt2 did not finish
+% FIXME discussion of the results; note that ropebwt2 did not finish without RLO; BWT-merge does not do RLO
 
-\smallbreak\noindent\textbf{Read Server.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files using the fast settings.
+\smallbreak\noindent\textbf{Read Server.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files.
 % FIXME sizes in native/sga format, merging time/space
@@
 We split the read collection into multiple subcollections, build the \BWT{}s of the subcollections with other tools, and merge the results.
-Depending on whether we run one or more \BWT{} construction jobs in parallel, the resulting algorithm can be slower but more space-efficient than the existing algorithms, or faster than them with similar memory usage.
+The resulting algorithm is typically slower but more space-efficient than the existing algorithms.
 
@@
 As a concrete example, we merged the 16 Read Server \BWT{} files into two files.
-This reduced the number of servers required to house the indexes from three to two, and also improved the query performance of the servers.
+This reduced the number of servers required to use the indexes from three to two, and also improved the query performance of the servers.
 
@@
 In this order, $\mSA[\mBWT.\mselect(i,0)]$ points to the beginning of the reverse complement of read $T_{i}$, if the collection includes the reverse complement of each read \cite{Li2014a}.
+With \emph{lexicographic order}, we can determine $\mSA[i]$ without samples by using $\mLF$ instead of $\Psi$, which is often faster in practice.
 

### 2015-10-28 paper/benchmark.R

@@ -42,12 +42,13 @@ nc = ncol(data)
 points(data[1:nr, 3], data[1:nr, 2] / 3600, type = "p", pch = 20)
 
 text(data[1, 3], data[1, 2] / 3600, data[1, 1], cex = 0.8, pos = 2)
-text(data[2, 3], data[2, 2] / 3600, data[2, 1], cex = 0.8, pos = 4)
-text(data[3, 3], data[3, 2] / 3600, data[3, 1], cex = 0.8, pos = 2)
+text(data[2, 3], data[2, 2] / 3600, data[2, 1], cex = 0.8, pos = 1)
+text(data[3, 3], data[3, 2] / 3600, data[3, 1], cex = 0.8, pos = 1)
 text(data[4, 3], data[4, 2] / 3600, data[4, 1], cex = 0.8, pos = 2)
 text(data[5, 3], data[5, 2] / 3600, data[5, 1], cex = 0.8, adj = c(0.65, 2))
 text(data[6, 3], data[6, 2] / 3600, data[6, 1], cex = 0.8, pos = 2)
 text(data[7, 3], data[7, 2] / 3600, data[7, 1], cex = 0.8, pos = 2)
+text(data[8, 3], data[8, 2] / 3600, data[8, 1], cex = 0.8, pos = 2)
 
 segments(124.2, 29, 124.2, 37, lty = "dashed")
 

### 2015-10-29 paper/paper.tex
Merge example.

@@
 
-% FIXME: a running example with 2 texts
+\begin{figure}[t!]
+\includegraphics[width=\textwidth]{merge_example.pdf}
+\caption{Merging the \BWT{}s of texts $R$ and $S$.
+Rank array \RA{} counts the number of suffixes of $R$ that are lexicographically smaller than or equal to the given suffix of $S$.
+We fill it by starting with $\mRA[1] = 1$ and iterating $\mRA[\mLF_{S}(i)] = \mLF_{R}(\mRA[i], \mBWT_{S}[i])$.
+Interleaving bitvector $B_{RS}$ tells whether the source of a character in the merged \BWT{} is in $\mBWT_{R}$ or $\mBWT_{S}$.
+We build it by setting bits $i+\mRA[i]$ to 1 for all $i$.}\label{fig:merge}
+\end{figure}
+
 Assume that we have two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively, and that we have already built \BWT{}s for them.
-Assume that we store the \BWT{}s in a two-level array, where the first level contains pointers to \emph{blocks} of size $b$ (bits).
+Assume that we store the \BWT{}s in two-level arrays, where the first level contains pointers to \emph{blocks} of size $b$ (bits).
 If the \BWT{} takes $x$ bits, the two-level array imposes $\frac{x}{b} \log x + \Oh(b)$ bits of space overhead.
@@
 \end{enumerate}
+See Figure~\ref{fig:merge} for an example with two texts.
 The overall time complexity of \BWT{} merging is $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r})$.
@@
 
-\smallbreak\noindent\textbf{Buffering.} To reduce disk usage and the amount of disk I/O, we buffer and compress the lexicographic ranks before writing them to disk.
+\smallbreak\noindent\textbf{Buffering.} To reduce disk I/O and disk usage, we buffer and compress the lexicographic ranks before writing them to disk.
 Each thread has two buffers: a \emph{run buffer} and a \emph{thread buffer}.
@@
 \begin{center}
-\caption{Datasets used in the experiments.
-\BWT{} sizes are for the native format.
-RLO indicates that the reads are sorted in reverse lexicographic order in each input file.}\label{table:datasets}\smallskip%
+\caption{Datasets used in the experiments. \BWT{} sizes are for the native format. RLO indicates that the reads are sorted in reverse lexicographic order.}\label{table:datasets}\smallskip%
 {
@@
 \hline
-\CEU & NA12878 &  284 Gbp & 2.81G &       101 bp & 50.3 GB \\
-     & NA12891 &  242 Gbp & 2.40G &       101 bp & 42.4 GB \\
-     & NA12892 &  245 Gbp & 2.42G &       101 bp & 43.8 GB \\
-     & Merged  &  771 Gbp & 7.63G &       101 bp & \\
+\CEU & NA12878 &  284 Gbp & 2.81G &       101 bp & 50.3 GB & 25.5 GB \\
+     & NA12891 &  242 Gbp & 2.40G &       101 bp & 42.4 GB & 19.7 GB \\
+     & NA12892 &  245 Gbp & 2.42G &       101 bp & 43.8 GB & 20.7 GB \\
+     & Merged  &  771 Gbp & 7.63G &       101 bp &  129 GB & \\
 \hline

### 2015-10-29 paper/comparison.R

@@ -17,8 +17,8 @@ xscale = c(0, 64, 128, 192, 256)
 xtitle = "Memory usage (GB)"
 xlabs = xscale
 
-yrange = c(0, 60)
-yscale = c(0, 12, 24, 36, 48, 60)
+yrange = c(0, 72)
+yscale = c(0, 12, 24, 36, 48, 60, 72)
 ytitle = ""
 ylabs = yscale
 
@@ -41,8 +41,9 @@ nc = ncol(data)
 
 points(data[1:nr, 3], data[1:nr, 2] / 3600, type = "p", pch = 20)
 
-text(data[1, 3], data[1, 2] / 3600, data[1, 1], cex = 0.8, pos = 2)
-text(data[2, 3], data[2, 2] / 3600, data[2, 1], cex = 0.8, pos = 2)
+text(data[1, 3], data[1, 2] / 3600, data[1, 1], cex = 0.8, pos = 2) # RopeBWT
+text(data[2, 3], data[2, 2] / 3600, data[2, 1], cex = 0.8, pos = 2) # RopeBWT2
+text(data[3, 3], data[3, 2] / 3600, data[3, 1], cex = 0.8, pos = 2) # BWT-merge
 
 dev.off()
 q()

### 2015-10-30 paper/paper.tex
Updated the experiments

@@
 With our workload of tens of threads allocating and freeing hundreds of gigabytes of memory, this created thousands of heaps with holes in the middle.
-Most calls to \texttt{free()} could not release the memory back to the operating system, while it took a while before a thread reached the heap where the \texttt{free()} took place.
+Most \texttt{free()} calls could not release the memory back to the operating system, while it took a while before a thread reached the heap where the \texttt{free()} took place.
 
@@
 \begin{center}
-\caption{Datasets used in the experiments. \BWT{} sizes are for the native format. RLO indicates that the reads are sorted in reverse lexicographic order.}\label{table:datasets}\smallskip%
+\caption{Datasets used in the experiments.
+The amount of sequence data and the number of reads in each file, and the size of the \BWT{} in the native format and in the Read Server format.
+RLO indicates that the reads are sorted in reverse lexicographic order.}\label{table:datasets}\smallskip%
 {
 \renewcommand{\baselinestretch}{1}\footnotesize
-\begin{tabular}{cc|ccc|cc}
+\begin{tabular}{cc|cc|cc|cc}
 \hline
- & & & & & \multicolumn{2}{c}{\textbf{\BWT{} size}} \\
-\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} & \textbf{Unsorted} & \textbf{RLO} \\
+ & & \multicolumn{2}{c|}{\textbf{Data}} & \multicolumn{2}{c|}{\textbf{Native \BWT}} & \multicolumn{2}{c}{\textbf{Read Server}} \\
+\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Unsorted} & \textbf{RLO} & \textbf{\BWT} & \textbf{\FMI} \\
 \hline
-\CEU & NA12878 &  284 Gbp & 2.81G &       101 bp & 50.3 GB & 25.5 GB \\
-     & NA12891 &  242 Gbp & 2.40G &       101 bp & 42.4 GB & 19.7 GB \\
-     & NA12892 &  245 Gbp & 2.42G &       101 bp & 43.8 GB & 20.7 GB \\
-     & Merged  &  771 Gbp & 7.63G &       101 bp &  129 GB & \\
+\CEU & NA12878 &  284 Gbp & 2.81G & 50.3 GB & 25.5 GB &      -- &      -- \\
+     & NA12891 &  242 Gbp & 2.40G & 42.4 GB & 19.7 GB &      -- &      -- \\
+     & NA12892 &  245 Gbp & 2.42G & 43.8 GB & 20.7 GB &      -- &      -- \\
+     & Merged  &  771 Gbp & 7.63G &  129 GB & 58.9 GB &      -- &      -- \\
 \hline
-\RS  & AA      &  433 Gbp & 4.69G & 73 or 100 bp &      -- & 38.5 GB \\
-     & TT      &  432 Gbp & 4.68G & 73 or 100 bp &      -- & 38.7 GB \\
-     & AT      &  275 Gbp & 2.98G & 73 or 100 bp &      -- & 26.6 GB \\
-     & TA      &  355 Gbp & 3.84G & 73 or 100 bp &      -- & 32.7 GB \\
-     & Merged  & 1.49 Tbp & 16.2G & 73 or 100 bp &      -- &  117 GB \\
+\RS  & AA      &  433 Gbp & 4.69G &      -- & 38.5 GB & 39.9 GB & 48.3 GB \\
+     & TT      &  432 Gbp & 4.68G &      -- & 38.7 GB & 40.0 GB & 48.4 GB \\
+     & AT      &  275 Gbp & 2.98G &      -- & 26.6 GB & 26.9 GB & 32.6 GB \\
+     & TA      &  355 Gbp & 3.84G &      -- & 32.7 GB & 33.5 GB & 40.6 GB \\
+     & Merged  & 1.49 Tbp & 16.2G &      -- &  117 GB &  126 GB &  152 GB \\
 \hline
-\RS  & *A, *C  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- &  225 GB \\
-     & Merged  & 2.45 Tbp & 26.5G & 73 or 100 bp &      -- & \\
+\RS  & *A, *C  & 2.45 Tbp & 26.5G &      -- &  225 GB &  232 GB &  281 GB \\
+     & Merged  & 2.45 Tbp & 26.5G &      -- &         &         &         \\
 \hline
-\RS  & *G, *T  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- &  226 GB \\
-     & Merged  & 2.44 Tbp & 26.5G & 73 or 100 bp &      -- & \\
+\RS  & *G, *T  & 2.44 Tbp & 26.5G &      -- &  226 GB &  232 GB &  281 GB \\
+     & Merged  & 2.44 Tbp & 26.5G &      -- &         &         &         \\
 \hline
@@
 
-\item \CEU{} contains reads from high-coverage sequencing of the \emph{CEU trio} (individuals NA12878, NA12891, and NA12892).
-We downloaded the gzipped FASTQ files (run accessions SRR622457, SRR622458, and SRR622459) and concatenated the files representing the same individual.
+\item \CEU{} contains 101~bp reads from high-coverage sequencing of the \emph{CEU trio} (individuals NA12878, NA12891, and NA12892).
+We downloaded the gzipped FASTQ files (run accessions SRR622457, SRR622458, and SRR622459) and concatenated the files for each individual.
 Then we corrected the sequencing errors with BFC \cite{Li2015} (\texttt{bfc -s 3g -t 16}) separately for each individual.
@@
 After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp.
-The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} \cite{Simpson2012}, distributed according to the last two bases.
+The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} (SGA) \cite{Simpson2012}, partitioned according to the last two bases.
 
@@
 We can also sort the reads by their likely positions in a reference genome.
-This \emph{position order} is useful, if we also want to store \emph{pairing information} for the reads.
+This \emph{position order} is useful for both compression and storing the \emph{pairing information} for the reads.
 Consider a graph with reads as its nodes and edges between paired reads.

### 2015-10-30 paper/comparison.R

@@ -44,6 +44,7 @@ points(data[1:nr, 3], data[1:nr, 2] / 3600, type = "p", pch = 20)
 text(data[1, 3], data[1, 2] / 3600, data[1, 1], cex = 0.8, pos = 2) # RopeBWT
 text(data[2, 3], data[2, 2] / 3600, data[2, 1], cex = 0.8, pos = 2) # RopeBWT2
 text(data[3, 3], data[3, 2] / 3600, data[3, 1], cex = 0.8, pos = 2) # BWT-merge
+text(data[6, 3], data[6, 2] / 3600, data[6, 1], cex = 0.8, pos = 1) # BWT-merge (RLO)
 
 dev.off()
 q()

### 2015-11-01 paper/paper.tex
Last experiments.

@@
 We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of smaller subcollections.
-The algorithm is capable of merging over 700 Gbp/day on a single system, with 20--30 gigabytes of memory overhead on top of the run-length encoded BWTs.
+With our largest 2.4 Tbp datasets, the algorithm is capable of merging 600 Gbp/day on a single system, using 30 gigabytes of memory overhead on top of the run-length encoded BWTs.
 \end{abstract}
@@
 The amount of sequence data and the number of reads in each file, and the size of the \BWT{} in the native format and in the Read Server format.
-RLO indicates that the reads are sorted in reverse lexicographic order.}\label{table:datasets}\smallskip%
+RLO indicates that the reads are sorted in reverse lexicographic order.
+The numbers in parentheses are estimates.}\label{table:datasets}\smallskip%
 {
@@
      & TA      &  355 Gbp & 3.84G &      -- & 32.7 GB & 33.5 GB &  40.6 GB \\
-     & Merged  & 1.49 Tbp & 16.2G &      -- &  117 GB &  126 GB &  152 GB \\
+     & Merged  & 1.49 Tbp & 16.2G &      -- &  117 GB &  126 GB & (152 GB) \\
 \hline
 \RS  & *A, *C  & 2.45 Tbp & 26.5G &      -- &  225 GB &  232 GB &   281 GB \\
-     & Merged  & 2.45 Tbp & 26.5G &      -- &         &         &         \\
+     & Merged  & 2.45 Tbp & 26.5G &      -- &  181 GB &  197 GB & (239 GB) \\
 \hline
 \RS  & *G, *T  & 2.44 Tbp & 26.5G &      -- &  226 GB &  232 GB &   281 GB \\
-     & Merged  & 2.44 Tbp & 26.5G &      -- &         &         &         \\
+     & Merged  & 2.44 Tbp & 26.5G &      -- &  180 GB &  197 GB & (238 GB) \\
 \hline
@@
 
-For the further experiments, we chose 128~MB run buffers, and 256~MB thread buffers and 6 merge buffers (overhead 30.8~GB).
+For the further experiments, we chose 128~MB run buffers, 256~MB thread buffers, and 6 merge buffers (overhead 30.8~GB).
 
@@
 The results can be seen in Figure~\ref{fig:benchmark} (right).
-% FIXME discussion of the results; note that ropebwt2 did not finish without RLO; BWT-merge does not do RLO
+When the reads are in the original order, \BWTmerge{} is 1.85x slower and 1.46x more space-efficient than building the final \BWT{} directly with \ropebwt.
+\ropebwt2{} ran out of memory when processing the last 10~Gbpp batch of reads.
+\BWTmerge{} would have been about 1.2x slower and 1.5x more space-efficient than it.
+The running time of \BWTmerge{} was split almost evenly between building the \BWT{}s of individual files and merging them.
 
-\smallbreak\noindent\textbf{Read Server.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files.
-% FIXME sizes in native/sga format, merging time/space
+When we tell \ropebwt{} and \ropebwt2{} to sort the reads in reverse lexicographic order, all tools improve their performance.
+\BWTmerge{} is now 1.70x slower and 2.12x more space-efficient than \ropebwt{} and 1.21x \emph{faster} and 1.09x more space-efficient than \ropebwt2.
+Again, \BWTmerge{} spent around half of the time building the individual \BWT{}s and another half merging them.
+Note that the final \BWT{} built by \BWTmerge{} is not in reverse lexicographic order, unlike the ones built with \ropebwt{} and \ropebwt2{}.
+When such functionality is implemented, \BWTmerge{} should be somewhat slower and somewhat more space-efficient than when maintaining the existing order of reads.
+
+\smallbreak\noindent\textbf{Read Server.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files (AA, CA, TA, GA, AC, CC, GC, and TC in the first collection; TT, GT, CT, AT, TG, GG, CG, AG in the second one).
+Merging the \BWT{}s took 81.3~hours and 83.0~hours, required 221~gigabytes and 219~gigabytes of memory, and used 297~gigabytes and 300~gigabytes of disk space, respectively.
+We reduced the size of the FM-indexes from around 560~GB to 480~GB in the format used in the Read Server.
+With the native format of \BWTmerge{}, the total size of the indexes was down to 360~GB.
+This will make it possible to host the indexes on two servers instead of the original three.
 
@@
 Our implementation of the algorithm in the \BWTmerge{} tool is fast enough to be used with terabases of sequence data.
-It requires only 20\nobreakdash--30~gigabytes of memory on top of the \BWT{}s to be merged.
+It requires only 30~gigabytes of memory on top of the \BWT{}s to be merged.
 As \BWT-based indexes do random accesses to large arrays, they must reside in memory in most applications.

### 2015-11-01 paper/comparison.R

@@ -43,7 +43,9 @@ points(data[1:nr, 3], data[1:nr, 2] / 3600, type = "p", pch = 20)
 
 text(data[1, 3], data[1, 2] / 3600, data[1, 1], cex = 0.8, pos = 2) # RopeBWT
 text(data[2, 3], data[2, 2] / 3600, data[2, 1], cex = 0.8, pos = 2) # RopeBWT2
-text(data[3, 3], data[3, 2] / 3600, data[3, 1], cex = 0.8, pos = 2) # BWT-merge
+text(data[3, 3], data[3, 2] / 3600, data[3, 1], cex = 0.8, pos = 3) # BWT-merge
+text(data[4, 3], data[4, 2] / 3600, data[4, 1], cex = 0.8, pos = 1) # RopeBWT (RLO)
+text(data[5, 3], data[5, 2] / 3600, data[5, 1], cex = 0.8, pos = 1) # RopeBWT2 (RLO)
 text(data[6, 3], data[6, 2] / 3600, data[6, 1], cex = 0.8, pos = 1) # BWT-merge (RLO)
 
 dev.off()

### 2015-11-03 paper/paper.tex
The first arXiv version of the paper.

@@
 We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of smaller subcollections.
-With our largest 2.4 Tbp datasets, the algorithm is capable of merging 600 Gbp/day on a single system, using 30 gigabytes of memory overhead on top of the run-length encoded BWTs.
+With our largest 2.4 Tbp datasets, the algorithm can merge 600 Gbp/day on a single system, using 30 gigabytes of memory overhead on top of the run-length encoded BWTs.
 \end{abstract}
@@
 The \emph{1000 Genomes Project} \cite{1000GP2015} sequenced the genomes of over 2500 humans, and there are many other projects that are similar or greater in scale.
-For each sequenced genome, a sequencing machine produces a large number of \emph{reads} (short sequences) that cover the genome many times over.
+A sequencing machine produces a large number of \emph{reads} (short sequences) that cover the sequenced genome many times over.
 For a 3~Gbp human genome, the total length of the reads is often 100~Gbp or more.
@@
 \emph{De novo assembly} of sequenced genomes is still too difficult to be routinely done.
-As a practical alternative, bioinformaticians typically align the reads are aligned to a \emph{reference genome} of the same species.
+As a practical alternative, bioinformaticians typically align the reads to a \emph{reference genome} of the same species.
 Because most reference genomes have been derived from the genomes of a small number of individuals, this introduces \emph{reference bias}, which may adversely affect the results of subsequent analysis.
@@
 Unique reads are compressed and indexed using the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994}, while metadata databases contain information about the original reads.
-In the pilot phase, the Read Server project works with the low-coverage and exome data from phase 3 of the 1000 Genomes Project.
+Initially, the project works with the low-coverage and exome data from phase 3 of the 1000 Genomes Project.
 After error correction and trimming the reads to either 73~bp or 100~bp, the 86~Tbp of raw data consisting of 922~billion original reads is reduced to 4.88~Tbp and 53.0~billion unique sequences.
@@
 Preprocessing large datasets can take days or even weeks.
-As a result, it is not always feasible to rebuild everything when new methods of analysis require new functionalities from the indexes.
+It is often not feasible to rebuild everything when new methods of analysis require new functionalities from the indexes.
 \BWT-based methods were chosen due to their versatility.
@@
 
-Read Server stores the unique reads in 16 \BWT-based indexes according to the last two bases of the read.
-This means that every query must be repeated 16 times, while the indexes require more space, as the compression cannot take advantage of the similarities between the reads in different indexes.
-Reducing the number of indexes would improve both index size and performance.
+Read Server partitions the unique reads based on the last two bases of the read.
+Every query must be repeated in all 16 indexes, and the \BWT{}s require more space, as the compression cannot take advantage of the similarities between the reads in different indexes.
+Reducing the number of indexes would improve both memory usage and query performance.
 However, the lack of \BWT{} construction algorithms that work with terabases of sequence data has made it infeasible so far.
@@
 
-\item \textbf{Hardware.} A single node in a computer cluster typically has tens of CPU cores, hundreds of gigabytes of memory, a limited amount of local disk space, and access to a large amount of shared disk space with no performance guarantees.
+\item \textbf{Hardware.} A single node in a computer cluster typically has tens of CPU cores, from tens to hundreds of gigabytes of memory, a limited amount of local disk space, and access to a large amount of shared disk space with no performance guarantees.
 Algorithms using a GPU or a large amount of fast disk space require special-purpose hardware.
@@
 \item \textbf{Efficiency.} Large \BWT{}s can be built by doing a lot of redundant work on multiple nodes.
-As most computer clusters do not have large amounts of unused capacity, such inefficient algorithms are unsuitable for repeated use.
+As most computer clusters do not have large amounts of unused capacity, such inefficient algorithms are not suitable for repeated use.
 
@@
 
-The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} for the sequences using a fast general-purpose algorithm \cite{Mori2008,Nong2011}, and then transform the suffix array into the \BWT.
-These algorithms cannot be used with large datasets, as they require several times more memory than the size of the dataset.
-Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index more than 1\nobreakdash--2~Mbp/s \cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Kaerkkaeinen2015a}.
+The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} for the sequences using a fast general-purpose algorithm \cite{Mori2008,Nong2011}.
+Given the sequences and their suffix array, we can easily build the \BWT{} of the sequences.
+These algorithms cannot be used with large datasets, as they require several times more memory than the sequences themselves.
+Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s \cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Kaerkkaeinen2015a}.
 
 There are several algorithms for constructing the \BWT{} \emph{directly} without having to build the suffix array first.
-They typically require only a limited amount of working space on top of a plain or compressed representation of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009} or use the disk as additional working space \cite{Ferragina2012,Beller2013}.
+They may require only a limited amount of working space on top of a plain or compressed representation of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}, or use the disk as additional working space \cite{Ferragina2012,Beller2013}.
 General-purpose implementations of these algorithms rarely exceed 1\nobreakdash--2~Mbp/s with large datasets.
 
-\emph{Specialized algorithms} for DNA sequences are much faster.
-Some algorithms achieve 5\nobreakdash--10~Mbps on general-purpose hardware, with their memory usage becoming a major bottleneck for datasets larger than about 1~Tbp \cite{Bauer2013,Li2014a}.
-GPU-based algorithms can be much faster, but their memory usage is also greater \cite{Liu2014,Pantaleoni2014}.
+\emph{Specialized algorithms} for DNA sequences achieve better time/space trade-offs.
+Some algorithms can index 5\nobreakdash--10~Mbps on general-purpose hardware, with their memory usage becoming a major bottleneck for datasets larger than about 1~Tbp \cite{Bauer2013,Li2014a}.
+GPU-based algorithms can be much faster, but their memory usage is also higher \cite{Liu2014,Pantaleoni2014}.
 Distributing the \BWT{} construction on multiple nodes can remove the obvious bottlenecks, but such algorithms generally do not use the hardware as efficiently as those running on  a single system \cite{Wang2015}.
@@
 In this paper, we propose a practical algorithm for building the \BWT{} for terabases of sequence data.
-The algorithm is based on splitting the sequence collection into a number of smaller subcollections, building the \BWT{} for each subcollection, and \emph{merging} the \BWT{}s into a single structure \cite{Siren2009}.
+The algorithm is based on dividing the sequence collection into a number of smaller subcollections, building the \BWT{} for each subcollection, and \emph{merging} the \BWT{}s into a single structure \cite{Siren2009}.
 The merging algorithm is faster than \BWT{} construction for the subcollections, while having a relatively small memory overhead on top of the final \BWT-based index.
-As the index must be loaded in memory before it can be used, index construction can be done on the same system as the index is used.
+As the index must be loaded in memory before it can be used, index construction can be done on the same system as the index is going to be used.
 
@@
 A \emph{string} $S[1,n] = s_{1} \dotsm s_{n}$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{1, \dotsc, \sigma}$.
-For indexing purposes, we consider \emph{text} strings $T[1,n]$ terminated by an endmarker $T[n] = \$ = 0$ not occurring anywhere else in the text.
+For indexing purposes, we consider \emph{text} strings $T[1,n]$ terminated by an endmarker $T[n] = \$ = 0$ not occurring elsewhere in the text.
 \emph{Binary} sequences are strings over the alphabet $\set{0, 1}$.
 A \emph{substring} of string $S$ is a sequence of the form $S[i,j] = s_{i} \dotsm s_{j}$.
-Substrings of the type $S[1,j]$ and $S[i,n]$ are called \emph{prefixes} and \emph{suffixes}, respectively.
+We call substrings of the type $S[1,j]$ and $S[i,n]$ \emph{prefixes} and \emph{suffixes}, respectively.
 
@@
 (If the text is evident from the context, we will omit it and write just \SA.) We can build the suffix array in $\Oh(n)$ time using approximately $n(\log n + \log \sigma + 2)$ bits of space, or just $2n$ bits of working space over the text and the suffix array \cite{Nong2011}.
-Given a \emph{pattern} $P$, we can find the \emph{lexicographic range} $[sp,ep]$ of suffixes of the text prefixed by the pattern in $\Oh(\abs{P} \log n)$ time by binary searching in the suffix array.
+Given a \emph{pattern} $P$, we can binary search the suffix array to find the \emph{lexicographic range} $[sp,ep]$ of suffixes prefixed by the pattern in $\Oh(\abs{P} \log n)$ time.
 The corresponding range of pointers $\mSA[sp,ep]$ lists the \emph{occurrences} of the pattern in the text.
@@
 For large texts, this can be a serious drawback.
-The more space-efficient alternatives to the suffix array are based on the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994}, an easily reversible permutation of the text with a similar combinatorial structure to the suffix array.
+We can use the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994} as a more space-efficient alternative to the suffix array.
+The \BWT{} is an easily reversible permutation of the text with a similar combinatorial structure to the suffix array.
 Given a text $T[1,n]$ and its suffix array, we can easily produce the \BWT{} as $\mBWT[i] = T[\mSA[i]-1]$ (with $\mBWT[i] = T[n]$, if $\mSA[i] = 1$).
@@
 A key observation is that if suffix $X$ precedes suffix $Y$ in lexicographic order, suffix $cX$ will also precede suffix $cY$, for any character $c$.
-Therefore, if $\mBWT[i]$ is the $j$\nobreakdash-th occurrence of character $c$ in the \BWT, and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order.
+Therefore, if $\mBWT[i]$ is the $j$\nobreakdash-th occurrence of character $c$ in the \BWT and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order.
 This forms the essence of \LF\emph{\nobreakdash-mapping}.
-The suffix preceding $T[\mSA[i],n]$ in text order is the one starting at $\mSA[\mLF(i)]$, where
+The pointer to suffix $T[\mSA[i]-1,n]$ is stored at $\mSA[\mLF(i)]$, where
 $$
@@
 By starting from $[sp_{\abs{P}}, ep_{\abs{P}}] = [\mC[P[\abs{P}]]+1, \mC[P[\abs{P}]+1]]$, we can find the lexicographic range of suffixes starting with the pattern in $\Oh(\abs{P} \cdot t_{r})$ time, where $t_{r}$ is the time required to answer \rank{} queries on the \BWT.
-In practice, the time complexity ranges from $\Oh(\abs{P})$ to $\Oh(\abs{P} \log n)$, depending on the encoding.
+In practice, the time complexity ranges from $\Oh(\abs{P})$ to $\Oh(\abs{P} \log n)$, depending on the encoding of the \BWT{}.
 
-In order to find the occurrences of the pattern in the text, the FM-index \emph{samples} some suffix array pointers, and uses \LF\nobreakdash-mapping to derive the unsampled pointers.
-The set of sampled pointers always includes a pointer to the beginning of each text.
+In order to find the occurrences of the pattern in the text, the FM-index \emph{samples} some suffix array pointers, including the one to the beginning of the text.
+Unsampled pointers can be derived by using \LF\nobreakdash-mapping.
 If $\mSA[i]$ is not sampled, the FM-index proceeds to $\mLF(i)$ and continues from there.
@@
 We want to build a (generalized) \BWT{} for the collection.
-The usual way is to assume that the endmarkers of all text are distinct, with the one at the end of text $T_{i}$ having character value $(0,i)$.
+The usual way is to assume that the endmarkers of all texts are distinct, with the one at the end of text $T_{i}$ having character value $(0,i)$.
 This guarantees that each suffix has a unique lexicographic rank among the suffixes of the collection.
-We still encode each endmarker as $0$ in the \BWT{} in order to save space.
+To save space, we still encode each endmarker as a $0$ in the \BWT{}.
 Because of this, \LF\nobreakdash-mapping does not work with $c = 0$, and we cannot match patterns spanning text boundaries.
 
-With large collections of short texts (such as short reads), there are more space-efficient alternatives to sampling.
-Because all endmarkers had distinct character values during sorting, we know that $\mSA[i]$ (with $i \le m)$ points to the endmarker of text $T_{i}$.
+When the texts are short (e.g.~reads), there are more space-efficient alternatives to sampling.
+Because all endmarkers have distinct character values during sorting, we know that $\mSA[i]$ (with $i \le m)$ points to the endmarker of text $T_{i}$.
 To reach the endmarker, we iterate
@@
 To overcome this bottleneck, we can use \emph{incremental construction algorithms} that build the FM-index directly.
-Some of them use an adjustable amount of working space on top of the FM-index, making it possible to index text colletions that are too large to fit in memory in an uncompressed form.
+Some of them use an adjustable amount of working space on top of the FM-index, making it possible to index text collections that are too large to fit in memory in an uncompressed form.
 
@@
 We can also have the \BWT{} of a text collection, and \emph{extend} multiple texts by inserting a new character to the beginning of each of them in a single update \cite{Bauer2013}.
+
 In all three cases, we may use either \emph{static} or \emph{dynamic} \cite{Chan2007} structures for the \BWT.
+Dynamic representations increase the size of the \BWT{} by a constant factor to facilitate fast updates (around 1.5x for \ropebwtii{} \cite{Li2014a}), while static representations require more space overhead for buffering the updates.
 
@@
 
-Assume that we have two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively, and that we have already built \BWT{}s for them.
-Assume that we store the \BWT{}s in two-level arrays, where the first level contains pointers to \emph{blocks} of size $b$ (bits).
+Assume that we have two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively, and that we have already built the \BWT{}s for them.
+Assume that we store the \BWT{}s in two-level arrays, where the first level contains pointers $b$\nobreakdash-bit \emph{blocks}.
 If the \BWT{} takes $x$ bits, the two-level array imposes $\frac{x}{b} \log x + \Oh(b)$ bits of space overhead.
@@
 
-\item \textbf{Search.} We search for all texts in collection $\Bcoll$ using $\mBWT_{\Acoll}$ and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of collection $\Bcoll$.
+\item \textbf{Search.} We search for all texts in collection $\Bcoll$ by using $\mBWT_{\Acoll}$, and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of collection $\Bcoll$.
 This takes $\Oh(n_{\Bcoll} t_{r})$ time.
@@
 \item \textbf{Sort.} We sort the ranks produced in the search phase to build the \emph{rank array} (\RA) of collection $\Bcoll$ relative to collection $\Acoll$.
-The rank array is defined as $\mRA_{\Bcoll \mid \Acoll}[i] = \mrank(X, \Acoll)$, where $X$ is the suffix pointed by $\mSA_{\Bcoll}[i]$.
+The rank array is defined as $\mRA_{\Bcoll \mid \Acoll}[i] = \mrank(X, \Acoll)$, where $\mSA_{\Bcoll}[i]$ points to suffix $X$.
 The array requires $n_{\Bcoll} \log n_{\Acoll}$ bits of space, and we can build it in $\Oh(sort(n_{\Bcoll}, n_{\Acoll}))$ time, where $sort(n, u)$ is the time required to sort $n$ integers from universe $[0,u]$.
@@
 This phase takes $\Oh(n_{\Acoll} + n_{\Bcoll})$ time.
-By reusing the blocks of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ for $\mBWT_{\Acoll \cup \Bcoll}$ once we have processed them, we can merge the \BWT{}s almost in-place, using $\Oh(\sqrt{x \log x})$ bits of working space, where $x$ is the maximum of the sizes of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ in bits.
+By reusing the blocks of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ for $\mBWT_{\Acoll \cup \Bcoll}$ once we have processed them, we can merge the \BWT{}s almost in-place.
+The total working space is $\Oh(\sqrt{x \log x})$ bits, where $x$ is the maximum of the sizes of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ in bits.
 
@@
 
-If we split text collection $\Acoll$ of total length $n$ into $p$ \emph{subcollections} of equal size, we can build $\mBWT_{\Acoll}$ by merging the \BWT{}s of the subcollections incrementally.
+If we split text collection $\Acoll$ of total length $n$ into $p$ \emph{subcollections} of equal size, we can build $\mBWT_{\Acoll}$ incrementally by merging the \BWT{}s of the subcollections.
 Merging the subcollections takes $\Oh((p+t_{r})n)$ time and uses (essentially) $\min(n, \frac{n}{p} \log n)$ bits of working space.
@@
 
-The key to reducing the space overhead is that we do not need to keep the rank array in memory.
-We can write the lexicographic ranks to \emph{disk} in the search phase, sort them to build the rank array in the sort phase, and scan the rank array once during the merge phase.
-In practice, we may want to \emph{interleave} the sort phase with the search/merge phases and store the lexicographic ranks on disk in a \emph{compressed form}, because the amount of fast disk space can also become a bottleneck.
+The key to reducing the space overhead is writing the lexicographic ranks to \emph{disk} in the search phase.
+If we can perform the sort phase on disk, we just need to scan the rank array once during the merge phase.
+In practice, as fast disk space is often scarce, we may want to \emph{interleave} the sort phase with the search and merge phases, and to \emph{compress} the lexicographic ranks before writing them to disk.
 In the following, we describe the key ideas that make large-scale \BWT{} merging fast and space-efficient.
@@
 
-% FIXME pseudocode if we have space for it
 Assume that we have a \emph{node} of the trie corresponding to suffix $X$, lexicographic rank $r$, and lexicographic range $[sp,ep]$.
-This means that suffix $X$ occurs $sp+1-ep$ times in collection $\Bcoll$, so we can output a \emph{run} of lexicographic ranks $(r, sp+1-ep)$.
+As suffix $X$ occurs $sp+1-ep$ times in collection $\Bcoll$, we can output a \emph{run} of lexicographic ranks $(r, sp+1-ep)$.
 Afterwards, we proceed to the \emph{children} of the node.
@@
 
-\smallbreak\noindent\textbf{Buffering.} To reduce disk I/O and disk usage, we buffer and compress the lexicographic ranks before writing them to disk.
+\smallbreak\noindent\textbf{Buffering.} To reduce disk I/O and space usage, we buffer and compress the lexicographic ranks before writing them to disk.
 Each thread has two buffers: a \emph{run buffer} and a \emph{thread buffer}.
@@
 
-Once a thread buffer becomes full, we insert it into the global \emph{merge buffers}.
-There are $k$ merge buffers, with buffer $i$ containing $2^{i-1}$ merged thread buffers.
-The insertion starts at buffer $1$.
-If buffer $i$ is empty, the thread inserts its thread buffer there and returns back to the search phase.
-Otherwise it takes the contents of merge buffer $i$, merges them with its thread buffer, and proceeds to buffer $i+1$.
-If a thread reaches buffer $k+1$, it writes its thread buffer to disk and returns back to work.
+Once the thread buffer becomes full, we merge it with the global \emph{merge buffers}.
+There are $k$ merge buffers $M_{1}$ to $M_{k}$, with buffer $M_{i}$ containing $2^{i-1}$ merged thread buffers.
+The merging starts from buffer $M_{1}$.
+If $M_{i}$ is empty, the thread swaps its thread buffer with the empty buffer and returns to the search phase.
+Otherwise it swaps $M_{i}$ with an empty buffer, merges the former $M_{i}$ with its thread buffer, and proceeds to $M_{i+1}$.
+If the merging reaches $M_{k+1}$, the thread writes its thread buffer to disk and returns back to work.
 
@@
 Each of the files is sorted and compressed.
-For interleaving the \BWT{}s, we need to merge the files to produce an increasing run-length encoded stream of lexicographic ranks.
-We can also use multiple threads.
+For interleaving the \BWT{}s, we need to merge the files to scan through the rank array.
+We can also use multiple threads here.
 One thread reads the files and performs a \emph{multiway merge} using a priority queue, producing a stream of lexicographic ranks.
 Another thread consumes the stream and uses it to interleave the \BWT{}s.
-Because the producer thread is usually slower than the consumer thread, we may want to use multiple threads for the multiway merge, if the disk is fast enough.
+If the disk is fast enough, we may want to use multiple threads for the multiway merge.
 
@@
 
-We have implemented the improved \BWT{} merging algorithm as a tool for merging the \BWT{}s of large collections of short reads.
+We have implemented the improved merging algorithm as a tool for merging the \BWT{}s of large collections of short reads.
 The tool, \BWTmerge{}, is written in C++, and the source code is available at GitHub.\footnote{\url{https://github.com/jltsiren/bwt-merge}} The implementation is built on top of the \emph{SDSL library} \cite{Gog2014b} and uses the features of C++11 extensively.
@@
 
-For \rank/\select{} support, we split the \BWT{}s into 64\nobreakdash-byte blocks of compressed data, so that no run continues from one block to the next one.
+For \rank/\select{} support, we split the \BWT{}s into 64\nobreakdash-byte blocks of compressed data, ensuring that no run continues from one block to the next one.
 For each block $i$, we store the total number of characters in blocks $1$ to $i-1$ as $n_{i}$, as well as the cumulative character counts $c_{i} = \mBWT.\mrank(n_{i},c)$ for $0 \le c \le 5$.
@@
 To compute $\mBWT.\mrank(j,c)$, we first use a \rank{} query on the $n_{i}$ sequence to find the correct block.
-A \select{} query on the same sequence transforms position $j$ into an offset within the block, while a \select{} query on the $c_{i}$ sequence gives the rank up to the beginning of the block.
+A \select{} query on the same sequence transforms the position $j$ into an offset within the block, while a \select{} query on the $c_{i}$ sequence gives the rank up to the beginning of the block.
 Finally, we decompress the block to answer the query.
@@
 With our workload of tens of threads allocating and freeing hundreds of gigabytes of memory, this created thousands of heaps with holes in the middle.
-Most \texttt{free()} calls could not release the memory back to the operating system, while it took a while before a thread reached the heap where the \texttt{free()} took place.
+Most \texttt{free()} calls did not release the memory back to the operating system, while it took a while before any thread could reuse the holes created by the \texttt{free()} calls.
 
@@
 
-We used two datasets from the phase 3 of the \emph{1000 Genomes Project} \cite{1000GP2015}:
+We used two datasets from phase 3 of the \emph{1000 Genomes Project} \cite{1000GP2015}:
 \begin{itemize}
@@
 
-\smallbreak\noindent\textbf{Parameters.} For benchmarking with different parameter values, we converted four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset to the \emph{native format} of \BWTmerge.
+\smallbreak\noindent\textbf{Parameters.} For benchmarking with different parameter values, we took four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset, and converted them to the \emph{native format} of \BWTmerge.
 This format includes the \BWT{} and the \rank/\select{} structures required by the FM-index.
@@
 
-The results of the benchmarks can be seen in Figure~\ref{fig:benchmark} (left).
-The average speed for inserting 1.06~Tbp into the initial file AA ranged from 8.27 to 9.40~Mbp/s, depending on the parameter values.
-Memory overhead was from 21.1 to 41.5~GB on top of the 124.2~GB required by the last pair of \BWT{}s.
-The temporary files required 287 to 306~gigabytes of disk space.
+The results can be seen in Figure~\ref{fig:benchmark} (left).
+The average speed for inserting 1.06~Tbp into file AA ranged from 8.27 to 9.40~Mbp/s, depending on the parameter values.
+Memory overhead was 21.1 to 41.5~GB on top of the 124.2~GB required by the last pair of \BWT{}s.
+The temporary files used 287 to 306~gigabytes of disk space.
 Thread buffer size and the number of merge buffers were the most important parameters.
@@
 \smallbreak\noindent\textbf{Comparison.} In the next experiment, we compared the performance of \BWTmerge{} to the \ropebwt{} \cite{Li2011-2013} implementation of the BCR algorithm \cite{Bauer2013} and to \ropebwtii{} \cite{Li2014a}.
-These two tools are the fastest existing \BWT{} construction tools for read collections on general hardware \cite{Li2014a}.
+These two are the fastest existing \BWT{} construction tools for read collections on general hardware \cite{Li2014a}.
 We ran \ropebwt{} with parameters \texttt{-btORf -abcr} and \ropebwtii{} with parameters \texttt{-bRm10g}.
@@
 We built the \BWT{} of the dataset in several ways.
-We concatenated all input files, and built the \BWT{} directly for the entire collection using both \ropebwt{} and \ropebwtii{}.
+We concatenated all input files and built the \BWT{} directly for the entire collection using both \ropebwt{} and \ropebwtii{}.
 We also built the \BWT{} for each input file separately using \ropebwt{}, and then merged the results using \BWTmerge{}.
@@
 When the reads are in the original order, \BWTmerge{} is 1.85x slower and 1.46x more space-efficient than building the final \BWT{} directly with \ropebwt.
-\ropebwt2{} ran out of memory when processing the last 10~Gbpp batch of reads.
+\ropebwt2{} ran out of memory when processing the last 10~GB batch of reads.
 \BWTmerge{} would have been about 1.2x slower and 1.5x more space-efficient than it.
-The running time of \BWTmerge{} was split almost evenly between building the \BWT{}s of individual files and merging them.
+The running time of \BWTmerge{} was split almost evenly between building the \BWT{}s of the individual files and merging them.
 
@@
 Again, \BWTmerge{} spent around half of the time building the individual \BWT{}s and another half merging them.
-Note that the final \BWT{} built by \BWTmerge{} is not in reverse lexicographic order, unlike the ones built with \ropebwt{} and \ropebwt2{}.
-When such functionality is implemented, \BWTmerge{} should be somewhat slower and somewhat more space-efficient than when maintaining the existing order of reads.
+Note that \BWTmerge{} builds a \BWT{} for the concatenation of three input files that are in RLO, while the \BWT{}s produced by the other tools are completely in RLO.
+Maintaining RLO during merging would reduce the size of the final \BWT{} from 58.9~GB to 54.4~GB.
 
-\smallbreak\noindent\textbf{Read Server.} In the final experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files (AA, CA, TA, GA, AC, CC, GC, and TC in the first collection; TT, GT, CT, AT, TG, GG, CG, AG in the second one).
+\smallbreak\noindent\textbf{Read Server.} In the last experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files (AA, CA, TA, GA, AC, CC, GC, and TC into the first file; TT, GT, CT, AT, TG, GG, CG, AG into the second one).
 Merging the \BWT{}s took 81.3~hours and 83.0~hours, required 221~gigabytes and 219~gigabytes of memory, and used 297~gigabytes and 300~gigabytes of disk space, respectively.
-We reduced the size of the FM-indexes from around 560~GB to 480~GB in the format used in the Read Server.
-With the native format of \BWTmerge{}, the total size of the indexes was down to 360~GB.
-This will make it possible to host the indexes on two servers instead of the original three.
+This reduced the size of the FM-indexes from around 560~GB to 480~GB.
+By converting the \BWT{}s to the native format of \BWTmerge{}, we further reduced the size of the indexes to 360~GB.
+This makes it possible to host the indexes on two servers instead of the original three.
 
@@
 It requires only 30~gigabytes of memory on top of the \BWT{}s to be merged.
-As \BWT-based indexes do random accesses to large arrays, they must reside in memory in most applications.
+As \BWT-based indexes access large arrays in a random fashion, they must reside in memory in most applications.
 Hence \BWTmerge{} can build the index on the same system as it is going to be used.
@@
 As a concrete example, we merged the 16 Read Server \BWT{} files into two files.
-This reduced the number of servers required to use the indexes from three to two, and also improved the query performance of the servers.
+This reduced the number of servers required to host the indexes from three to two, and also improved the query performance of the servers.
 
 In the future, we are going to extend \BWTmerge{} to support different \emph{text orders}, and to optionally \emph{remove duplicate texts} from the merged collection.
-The current implementation maintains the existing order by inserting the texts from $\mBWT_{\Bcoll}$ after the texts in $\mBWT_{\Acoll}$.
-This makes it easy to determine the original identifiers of the texts without having to store a permutation.
-Other text orders can be useful for different purposes.
+The current algorithm maintains the existing order by inserting the texts from $\mBWT_{\Bcoll}$ after the texts in $\mBWT_{\Acoll}$.
+This makes it easy to determine the original text identifiers without having to store a permutation.
+Other text orders are useful for different purposes.
 
@@
 Sorting the reads by their \emph{reverse complements} also improves compression in a similar way.
-In this order, $\mSA[\mBWT.\mselect(i,0)]$ points to the beginning of the reverse complement of read $T_{i}$, if the collection includes the reverse complement of each read \cite{Li2014a}.
+In this order, $\mSA[\mBWT.\mselect(i,0)]$ points to the beginning of the reverse complement of read $T_{i}$, if the collection includes the reverse complement of every read \cite{Li2014a}.
 With \emph{lexicographic order}, we can determine $\mSA[i]$ without samples by using $\mLF$ instead of $\Psi$, which is often faster in practice.

### 2015-11-04 paper/paper.tex
References; link to arXiv version of the paper.

@@
 These algorithms cannot be used with large datasets, as they require several times more memory than the sequences themselves.
-Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s \cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Kaerkkaeinen2015a}.
+Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s \cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Nong2015,Kaerkkaeinen2015a,Liu2015a}.
 

### 2015-11-05 paper/paper.tex
Almost down to the 10-page limit.

--- 2015-11-03 paper/paper.tex
@@
 In order to avoid the reference bias introduced by mapping reads to a reference genome, bioinformaticians are investigating reference-free methods for analyzing sequenced genomes.
-With large projects sequencing thousands of individuals, this raises the need for tools capable of processing terabases of sequence data.
+With large projects sequencing thousands of individuals, this raises the need for tools capable of handling terabases of sequence data.
 A key method is the Burrows-Wheeler transform (BWT), which is widely used for compressing and indexing reads.
-We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of smaller subcollections.
-With our largest 2.4 Tbp datasets, the algorithm can merge 600 Gbp/day on a single system, using 30 gigabytes of memory overhead on top of the run-length encoded BWTs.
+We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of subcollections.
+With our 2.4 Tbp datasets, the algorithm can merge 600 Gbp/day on a single system, using 30 gigabytes of memory overhead on top of the run-length encoded BWTs.
 \end{abstract}
@@
 \emph{De novo assembly} of sequenced genomes is still too difficult to be routinely done.
-As a practical alternative, bioinformaticians typically align the reads to a \emph{reference genome} of the same species.
-Because most reference genomes have been derived from the genomes of a small number of individuals, this introduces \emph{reference bias}, which may adversely affect the results of subsequent analysis.
-Switching from reference sequences to \emph{reference graphs} can help to reduce the bias, but such transition will likely take years \cite{Church2015}.
+As a practical alternative, bioinformaticians usually align the reads to a \emph{reference genome} of the same species.
+Because most reference genomes come from the genomes of a small number of individuals, this introduces \emph{reference bias}, which may adversely affect the results of subsequent analysis.
+Switching from reference sequences to \emph{reference graphs} can reduce the bias, but such transition will likely take years \cite{Church2015}.
 
@@
 Read Server partitions the unique reads based on the last two bases of the read.
-Every query must be repeated in all 16 indexes, and the \BWT{}s require more space, as the compression cannot take advantage of the similarities between the reads in different indexes.
+Every query must be repeated in all 16 indexes, and the \BWT{}s require more space, as we cannot compress the similarities between the reads in different indexes.
 Reducing the number of indexes would improve both memory usage and query performance.
@@
 There are four often contradictory requirements for large-scale \BWT{} construction:
-\begin{itemize}
-
-\item \textbf{Speed.} Larger datasets require faster algorithms. As a rough guideline, an algorithm processing 1~Mbp/s is good for up to 100~Gbp, while remaining somewhat useful until 1~Tbp of data.
-
-\item \textbf{Memory.} We may have to process $n$~bp datasets on systems with less than $n$~bits of memory.
-
-\item \textbf{Hardware.} A single node in a computer cluster typically has tens of CPU cores, from tens to hundreds of gigabytes of memory, a limited amount of local disk space, and access to a large amount of shared disk space with no performance guarantees.
+\textbf{Speed.} Larger datasets require faster algorithms. As a rough guideline, an algorithm processing 1~Mbp/s is good for up to 100~Gbp, while remaining somewhat useful until 1~Tbp of data.
+\textbf{Memory.} We may have to process $n$~bp datasets on systems with less than $n$~bits of memory.
+\textbf{Hardware.} A single node in a typical computer cluster has tens of CPU cores, from tens to hundreds of gigabytes of memory, a limited amount of local disk space, and access to shared disk space with no performance guarantees.
 Algorithms using a GPU or a large amount of fast disk space require special-purpose hardware.
-
-\item \textbf{Efficiency.} Large \BWT{}s can be built by doing a lot of redundant work on multiple nodes.
+\textbf{Efficiency.} Large \BWT{}s can be built by doing a lot of redundant work on multiple nodes.
 As most computer clusters do not have large amounts of unused capacity, such inefficient algorithms are not suitable for repeated use.
 
-\end{itemize}
-
-The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} for the sequences using a fast general-purpose algorithm \cite{Mori2008,Nong2011}.
+The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} using a fast general-purpose algorithm \cite{Mori2008,Nong2011}.
 Given the sequences and their suffix array, we can easily build the \BWT{} of the sequences.
 These algorithms cannot be used with large datasets, as they require several times more memory than the sequences themselves.
-Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s \cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Kaerkkaeinen2015a}.
+Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s
+% DCC:
+\cite{Kaerkkaeinen2015a}.
+% arXiv:
+%\cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Nong2015,Kaerkkaeinen2015a,Liu2015a}.
 
-There are several algorithms for constructing the \BWT{} \emph{directly} without having to build the suffix array first.
-They may require only a limited amount of working space on top of a plain or compressed representation of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}, or use the disk as additional working space \cite{Ferragina2012,Beller2013}.
-General-purpose implementations of these algorithms rarely exceed 1\nobreakdash--2~Mbp/s with large datasets.
-
+There are algorithms for constructing the \BWT{} \emph{directly} without having to build the suffix array first.
+Some require only a limited amount of working space on top of a plain or compressed representation of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}, or use the disk as additional working space \cite{Ferragina2012,Beller2013}.
+These general-purpose algorithms rarely exceed 1\nobreakdash--2~Mbp/s with large datasets.
 \emph{Specialized algorithms} for DNA sequences achieve better time/space trade-offs.
-Some algorithms can index 5\nobreakdash--10~Mbps on general-purpose hardware, with their memory usage becoming a major bottleneck for datasets larger than about 1~Tbp \cite{Bauer2013,Li2014a}.
+Some algorithms can index 5\nobreakdash--10~Mbps using ordinary hardware, with their memory usage becoming a bottleneck for datasets larger than about 1~Tbp \cite{Bauer2013,Li2014a}.
 GPU-based algorithms can be much faster, but their memory usage is also higher \cite{Liu2014,Pantaleoni2014}.
-Distributing the \BWT{} construction on multiple nodes can remove the obvious bottlenecks, but such algorithms generally do not use the hardware as efficiently as those running on  a single system \cite{Wang2015}.
+Distributing the \BWT{} construction to multiple nodes can remove the obvious bottlenecks, but such algorithms generally use the hardware less efficiently than those running on a single system \cite{Wang2015}.
 
 In this paper, we propose a practical algorithm for building the \BWT{} for terabases of sequence data.
-The algorithm is based on dividing the sequence collection into a number of smaller subcollections, building the \BWT{} for each subcollection, and \emph{merging} the \BWT{}s into a single structure \cite{Siren2009}.
+The algorithm is based on dividing the sequence collection into a number of subcollections, building the \BWT{} for each subcollection, and \emph{merging} the \BWT{}s into a single structure \cite{Siren2009}.
 The merging algorithm is faster than \BWT{} construction for the subcollections, while having a relatively small memory overhead on top of the final \BWT-based index.
-As the index must be loaded in memory before it can be used, index construction can be done on the same system as the index is going to be used.
+As the index must be loaded in memory for use, it can be built on the same system as it is going to be used.
 
@@
 
-The \emph{suffix array} (\SA) \cite{Manber1993} is a simple full-text index of a string.
-Given a text $T$, its suffix array $\mSA_{T}[1,n]$ is an array of pointers to the suffixes of the text in \emph{lexicographic order}.
-(If the text is evident from the context, we will omit it and write just \SA.) We can build the suffix array in $\Oh(n)$ time using approximately $n(\log n + \log \sigma + 2)$ bits of space, or just $2n$ bits of working space over the text and the suffix array \cite{Nong2011}.
-Given a \emph{pattern} $P$, we can binary search the suffix array to find the \emph{lexicographic range} $[sp,ep]$ of suffixes prefixed by the pattern in $\Oh(\abs{P} \log n)$ time.
-The corresponding range of pointers $\mSA[sp,ep]$ lists the \emph{occurrences} of the pattern in the text.
+The \emph{suffix array} (\SA) \cite{Manber1993} is a simple full-text index.
+Given a text $T$, its suffix array $\mSA_{T}[1,n]$ is an array of pointers to the suffixes of the text in \emph{lexicographic order}.\footnote{If the text is evident from the context, we will omit the subscript and write just \SA{}, \BWT{}, etc.} We can build the suffix array in $\Oh(n)$ time using $2n$ bits of working space on top of the text and the suffix array \cite{Nong2011}.
+Given a \emph{pattern} $P$, we can find the \emph{lexicographic range} $[sp,ep]$ of suffixes prefixed by the pattern in $\Oh(\abs{P} \log n)$ time.
+The range of pointers $\mSA[sp,ep]$ lists the \emph{occurrences} of the pattern in the text.
 
-While the suffix array is simple, it requires several times more memory than the original text.
+The suffix array requires several times more memory than the original text.
 For large texts, this can be a serious drawback.
@@
 
-A key observation is that if suffix $X$ precedes suffix $Y$ in lexicographic order, suffix $cX$ will also precede suffix $cY$, for any character $c$.
-Therefore, if $\mBWT[i]$ is the $j$\nobreakdash-th occurrence of character $c$ in the \BWT and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order.
+If suffix $X$ precedes suffix $Y$ in lexicographic order, suffix $cX$ will also precede suffix $cY$, for any character $c$.
+If $\mBWT[i]$ is the $j$\nobreakdash-th occurrence of character $c$ in the \BWT{} and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order.
 This forms the essence of \LF\emph{\nobreakdash-mapping}.
-The pointer to suffix $T[\mSA[i]-1,n]$ is stored at $\mSA[\mLF(i)]$, where
+Let $\mC[c]$ bet the number of suffixes starting with a character smaller than $c$, and let $S.\mrank(i,c)$ be the number of occurrences of $c$ in the prefix $S[1,i]$.
+Then $\mSA[\mLF(i)]$, where
 $$
-\mLF(i) = \mC[\mBWT[i]] + \mBWT.\mrank(i, \mBWT[i]),
-$$
-$\mC[c]$ is the number of suffixes starting with a character smaller than $c$, and $S.\mrank(i,c)$ is the number of occurrences of character $c$ in the prefix $S[1,i]$.
-For searching, we use a more general definition of \LF\nobreakdash-mapping:
-$$
+\mLF(i) = \mLF(i,\mBWT[i])
+\,\, \textrm{and} \,\,
 \mLF(i,c) = \mC[c] + \mBWT.\mrank(i, c),
 $$
-which tells the number of suffixes $X$ of text $T$ with $X \le c T[\mSA[i],n]$ in lexicographic order.
-This is known as the \emph{lexicographic rank} $\mrank(cT[\mSA[i],n], T)$ of string $cT[\mSA[i],n]$ among the suffixes of text $T$.
+points to the suffix $T[\mSA[i]-1,n]$.
+The more general version of \LF-mapping $\mLF(i,c)$ tells the number of suffixes $X$ of text $T$ with $X \le c T[\mSA[i],n]$.
+This is the \emph{lexicographic rank} $\mrank(cT[\mSA[i],n], T)$ of $cT[\mSA[i],n]$ among the suffixes of text $T$.
 
 The \emph{FM-index} (\FMI) \cite{Ferragina2005a} uses the Burrows-Wheeler transform as a full-text index by augmenting it with a few additional structures.
-The lexicographic range $[sp,ep]$ corresponding to pattern $P$ is found through \emph{backward searching}.
+The lexicographic range $[sp,ep]$ corresponding to pattern $P$ is found by using \emph{backward searching}.
 Let $[sp_{i},ep_{i}]$ be the lexicographic range of suffixes of text $T$ matching suffix $P[i, \abs{P}]$ of the pattern.
-We can find $[sp_{i-1},ep_{i-1}]$ as
-$$
-[sp_{i-1},ep_{i-1}] = [\mLF(sp_{i}-1, P[i-1]) + 1, \mLF(ep_{i}, P[i-1])].
-$$
+We find $[sp_{i-1},ep_{i-1}]$ as $[\mLF(sp_{i}-1, P[i-1]) + 1, \mLF(ep_{i}, P[i-1])]$.
 By starting from $[sp_{\abs{P}}, ep_{\abs{P}}] = [\mC[P[\abs{P}]]+1, \mC[P[\abs{P}]+1]]$, we can find the lexicographic range of suffixes starting with the pattern in $\Oh(\abs{P} \cdot t_{r})$ time, where $t_{r}$ is the time required to answer \rank{} queries on the \BWT.
@@
 
-In order to find the occurrences of the pattern in the text, the FM-index \emph{samples} some suffix array pointers, including the one to the beginning of the text.
-Unsampled pointers can be derived by using \LF\nobreakdash-mapping.
+The FM-index \emph{samples} some suffix array pointers, including the one to the beginning of the text.
+When unsampled pointers are needed, they are derived by using \LF\nobreakdash-mapping.
 If $\mSA[i]$ is not sampled, the FM-index proceeds to $\mLF(i)$ and continues from there.
 If $\mSA[\mLF^{k}(i)]$ is the first sample encountered, $\mSA[i] = \mSA[\mLF^{k}(i)] + k$.
-Depending on the way the samples are selected, we may need a binary sequence to mark the pointers that have been sampled.
+% arXiv:
+%Depending on the way the samples are selected, we may need a binary sequence to mark the pointers that have been sampled.
 
@@
 We want to build a (generalized) \BWT{} for the collection.
-The usual way is to assume that the endmarkers of all texts are distinct, with the one at the end of text $T_{i}$ having character value $(0,i)$.
-This guarantees that each suffix has a unique lexicographic rank among the suffixes of the collection.
+The usual way is to make all endmarkers distinct, giving the one at the end of text $T_{i}$ character value $(0,i)$.
+This also makes all suffixes of the collection distinct.
 To save space, we still encode each endmarker as a $0$ in the \BWT{}.
@@
 When the texts are short (e.g.~reads), there are more space-efficient alternatives to sampling.
-Because all endmarkers have distinct character values during sorting, we know that $\mSA[i]$ (with $i \le m)$ points to the endmarker of text $T_{i}$.
-To reach the endmarker, we iterate
-$$
-\Psi(i) = \mBWT.\mselect(i - \mC[c], c),
-$$
-where $c$ is the largest character value with $\mC[c] < i$ and $S.\mselect(i,c)$ finds the $i$\nobreakdash-th occurrence of character $c$ in string $S$.
+Because all endmarkers have distinct values during sorting, we know that $\mSA[i]$ (with $i \le m)$ points to the end of text $T_{i}$.
+To find the end, we iterate $\Psi(i) = \mBWT.\mselect(i - \mC[c], c)$, where $c$ is the largest value with $\mC[c] < i$ and $S.\mselect(i,c)$ finds the $i$\nobreakdash-th occurrence of character $c$ in string $S$.
 If $k \ge 0$ is the smallest value for which $j = \Psi^{k}(i) \le m$, we know that $\mSA[i]$ points to offset $\abs{T_{j}} - k$ in text $T_{j}$.
@@
 We can \emph{extract} text $T_{i}$ in $\Oh(\abs{T_{i}} \cdot t_{r})$ time by using \LF\nobreakdash-mapping \cite{Burrows1994}.
-We start from the endmarker at $\mBWT[i]$ and extract the text backwards as
-$$
-T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)] \,\, \textrm{for} \,\, 1 \le j < \abs{T_{i}}.
-$$
-As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also extract the lexicographic ranks of all suffixes of text $T_{i}$ in the process.
+We start from the endmarker at $\mBWT[i]$ and extract the text backwards as $T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)]$ for $1 \le j < \abs{T_{i}}$.
+As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also find the lexicographic ranks of all suffixes of text $T_{i}$ in the process.
 
@@
 The FM-index was introduced as a more space-efficient alternative to the suffix array.
-If we have to build a suffix array first in order to construct an FM-index, we lose a large part of the benefits, and index construction becomes the overall bottleneck.
-To overcome this bottleneck, we can use \emph{incremental construction algorithms} that build the FM-index directly.
-Some of them use an adjustable amount of working space on top of the FM-index, making it possible to index text collections that are too large to fit in memory in an uncompressed form.
+If we need the suffix array to build the FM-index, a large part of this benefit is lost, and index construction becomes the bottleneck.
+To overcome the bottleneck, we can use \emph{incremental construction algorithms} that build the FM-index directly.
+Some of them use an adjustable amount of working space on top of the FM-index, making it possible to index text collections larger than the size of the memory.
 
-Assume that we have built the \BWT{} of text $T$, and we want to \emph{transform} the \BWT{} into that of text $cT$, where $c$ is a character.
-The transformation \cite{Hon2007} is quite straightforward.
-We first find the pointer $\mSA[i]$ to the beginning of text $T$ (the position $i$, where $\mBWT[i] = 0$).
+Assume that we have built the \BWT{} of text $T$, and we want to \emph{transform} the \BWT{} into that of text $cT$, where $c$ is a character \cite{Hon2007}.
+We find the pointer $\mSA[i]$ to the beginning of text $T$ (where $\mBWT[i] = 0$).
 Then we determine the lexicographic rank $j = \mrank(cT, T) = \mC[c] + \mBWT.\mrank(i, c)$ of text $cT$ among the suffixes of text $T$.
-Finally we transform the \BWT{} by \emph{replacing} the endmarker at $\mBWT[i]$ with the inserted character $c$, and by \emph{inserting} a new endmarker between $\mBWT[j]$ and $\mBWT[j+1]$.
+Finally \emph{replace} the endmarker at $\mBWT[i]$ with the inserted character $c$ and \emph{insert} a new endmarker between $\mBWT[j]$ and $\mBWT[j+1]$.
 
-There are six basic ways of using the transformation in a \BWT{} construction algorithm.
-Instead of inserting a single character at a time, we may use \emph{batch updates} and transform the \BWT{} of text $T$ directly into that of text $XT$, where $X$ is any string \cite{Hon2007}.
+We can use the transformation for \BWT{} construction in several ways.
+We can do \emph{batch updates} and transform the \BWT{} of text $T$ into that of text $XT$, where $X$ is a string \cite{Hon2007}.
 We can start with the \BWT{}s of text collections $\Acoll$ and $\Bcoll$, and \emph{merge} them into the \BWT{} of collection $\Acoll \cup \Bcoll$ \cite{Siren2009}.
-We can also have the \BWT{} of a text collection, and \emph{extend} multiple texts by inserting a new character to the beginning of each of them in a single update \cite{Bauer2013}.
-
-In all three cases, we may use either \emph{static} or \emph{dynamic} \cite{Chan2007} structures for the \BWT.
-Dynamic representations increase the size of the \BWT{} by a constant factor to facilitate fast updates (around 1.5x for \ropebwtii{} \cite{Li2014a}), while static representations require more space overhead for buffering the updates.
-
-\Section{Merging \BWT{}s of text collections}
+We can also \emph{extend} multiple texts at once by inserting a new character to the beginning of each of them \cite{Bauer2013}.
+In all cases, we can use either \emph{static} or \emph{dynamic} \cite{Chan2007} structures for the \BWT.
+Dynamic representations increase the size of the \BWT{} (e.g.~by around 1.5x in \ropebwtii{} \cite{Li2014a}), while static representations require more space overhead for buffering the updates.
 
@@
 Interleaving bitvector $B_{RS}$ tells whether the source of a character in the merged \BWT{} is in $\mBWT_{R}$ or $\mBWT_{S}$.
-We build it by setting bits $i+\mRA[i]$ to 1 for all $i$.}\label{fig:merge}
+We build it by setting bits $i+\mRA[i]$ to 1 for all $i$.}\label{fig:merge}%
+\vspace{-6pt}
 \end{figure}
 
-Assume that we have two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively, and that we have already built the \BWT{}s for them.
-Assume that we store the \BWT{}s in two-level arrays, where the first level contains pointers $b$\nobreakdash-bit \emph{blocks}.
-If the \BWT{} takes $x$ bits, the two-level array imposes $\frac{x}{b} \log x + \Oh(b)$ bits of space overhead.
-With $b = \sqrt{x \log x}$, the overhead becomes $\Oh(\sqrt{x \log x})$ bits.
+Assume that we want to merge the \BWT{}s of two text collections $\Acoll$ and $\Bcoll$ of total length $n_{\Acoll}$ and $n_{\Bcoll}$, respectively \cite{Siren2009}.
+We store the \BWT{}s in two-level arrays, where the first level contains pointers to $b$\nobreakdash-bit \emph{blocks}.
+If a \BWT{} takes $x$ bits, the space overhead from the array is $\frac{x}{b} \log x + \Oh(b)$ bits.
+This becomes $\Oh(\sqrt{x \log x})$ bits with $b = \sqrt{x \log x}$.
+The merging algorithm has three phases: search, sort, and merge.
+It uses $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r})$ time and $\min(n_{\Bcoll} \log n_{\Acoll}, n_{\Acoll} + n_{\Bcoll}) + \Oh(\sqrt{x \log x})$ bits of working space in addition to the \BWT{}s and the structures required to use them as FM-indexes.
+See Figure~\ref{fig:merge} for an example with two texts.
 
-The algorithm for merging $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ into $\mBWT_{\Acoll \cup \Bcoll}$ has three phases \cite{Siren2009}:
-\begin{enumerate}
-
-\item \textbf{Search.} We search for all texts in collection $\Bcoll$ by using $\mBWT_{\Acoll}$, and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of collection $\Bcoll$.
+\smallbreak\noindent\textbf{Search.} We search for all texts of collection $\Bcoll$ in $\mBWT_{\Acoll}$, and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of collection $\Bcoll$.
 This takes $\Oh(n_{\Bcoll} t_{r})$ time.
@@
 
-\item \textbf{Sort.} We sort the ranks produced in the search phase to build the \emph{rank array} (\RA) of collection $\Bcoll$ relative to collection $\Acoll$.
+\smallbreak\noindent\textbf{Sort.} We build the \emph{rank array} (\RA) of $\Bcoll$ relative to $\Acoll$ by sorting the ranks.
 The rank array is defined as $\mRA_{\Bcoll \mid \Acoll}[i] = \mrank(X, \Acoll)$, where $\mSA_{\Bcoll}[i]$ points to suffix $X$.
 The array requires $n_{\Bcoll} \log n_{\Acoll}$ bits of space, and we can build it in $\Oh(sort(n_{\Bcoll}, n_{\Acoll}))$ time, where $sort(n, u)$ is the time required to sort $n$ integers from universe $[0,u]$.
-If we extract the texts from $\mBWT_{\Bcoll}$ in the search phase, we can write the ranks directly into the rank array, making this phase unnecessary.
+If we extracted the texts from $\mBWT_{\Bcoll}$, we can write the ranks directly into the rank array, making this phase trivial.
+We can also encode the rank array as a binary sequence $B_{\Acoll \cup \Bcoll}$ of length $n_{\Acoll} + n_{\Bcoll}$.
+This \emph{interleaving bitvector} is built by setting $B_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]] = 1$ for $1 \le i \le n_{\Bcoll}$.
+If $B_{\Acoll \cup \Bcoll}[j] = 1$, we know that $\mSA_{\Acoll \cup \Bcoll}[j]$ points to a suffix of $\Bcoll$.
 
-\item \textbf{Merge.} We interleave $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ according to the rank array.
+\smallbreak\noindent\textbf{Merge.} We interleave $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ according to the rank array.
 If $\mRA_{\Bcoll \mid \Acoll}[i] = j$, the merged \BWT{} will have $j$ characters from $\mBWT_{\Acoll}$ before $\mBWT_{\Bcoll}[i]$.
 This phase takes $\Oh(n_{\Acoll} + n_{\Bcoll})$ time.
-By reusing the blocks of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ for $\mBWT_{\Acoll \cup \Bcoll}$ once we have processed them, we can merge the \BWT{}s almost in-place.
+By reusing the blocks of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ for $\mBWT_{\Acoll \cup \Bcoll}$, we can merge the \BWT{}s almost in-place.
 The total working space is $\Oh(\sqrt{x \log x})$ bits, where $x$ is the maximum of the sizes of $\mBWT_{\Acoll}$ and $\mBWT_{\Bcoll}$ in bits.
-
-\end{enumerate}
-See Figure~\ref{fig:merge} for an example with two texts.
-The overall time complexity of \BWT{} merging is $\Oh(n_{\Acoll} + n_{\Bcoll} t_{r})$.
-The algorithm uses $n_{\Bcoll} \log n_{\Acoll} + \Oh(\sqrt{x \log x})$ bits of working space in addition to the \BWT{}s and the structures required to use them as FM-indexes.
-
-Alternatively, we can build the \emph{interleaving bitvector} $B_{\Acoll \cup \Bcoll}$.
-This bitvector is a binary sequence of length $n_{\Acoll} + n_{\Bcoll}$ that encodes the same information as the rank array.
-We set $B_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]] = 1$ for $1 \le i \le n_{\Bcoll}$, denoting that $\mSA_{\Acoll \cup \Bcoll}[i + \mRA_{\Bcoll \mid \Acoll}[i]]$ points to a suffix of collection $\Bcoll$.
-This approach uses $n_{\Acoll} + n_{\Bcoll} + \Oh(\sqrt{x \log x})$ bits of working space.
 
@@
 
-If we split text collection $\Acoll$ of total length $n$ into $p$ \emph{subcollections} of equal size, we can build $\mBWT_{\Acoll}$ incrementally by merging the \BWT{}s of the subcollections.
-Merging the subcollections takes $\Oh((p+t_{r})n)$ time and uses (essentially) $\min(n, \frac{n}{p} \log n)$ bits of working space.
+If we split a text collection $\Acoll$ of total length $n$ into $p$ \emph{subcollections} of equal size, we can build $\mBWT_{\Acoll}$ incrementally by merging the \BWT{}s of the subcollections.
+This takes $\Oh((p+t_{r})n)$ time and uses essentially $\min(n, \frac{n}{p} \log n)$ bits of working space.
 
 When the collection is large, the space overhead of the construction algorithm often determines whether we can build the \BWT{} of the collection.
-Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large.
+% arXiv:
+%Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large.
 The space overhead from the rank array or the interleaving bitvector (or their equivalents in the other space-efficient algorithms) may also be too much.
 We can make the rank array fit in memory by increasing the number of subcollections, but that can make the construction too slow.
-
-The key to reducing the space overhead is writing the lexicographic ranks to \emph{disk} in the search phase.
-If we can perform the sort phase on disk, we just need to scan the rank array once during the merge phase.
-In practice, as fast disk space is often scarce, we may want to \emph{interleave} the sort phase with the search and merge phases, and to \emph{compress} the lexicographic ranks before writing them to disk.
-In the following, we describe the key ideas that make large-scale \BWT{} merging fast and space-efficient.
+%
+We can reduce the overhead by writing the lexicographic ranks to \emph{disk} in the search phase.
+If we can sort the ranks on disk, we just need to scan the rank array once during the merge phase.
+As fast disk space is often scarce, we may want to \emph{interleave} the sorting with the search and merge phases, and to \emph{compress} the ranks before writing them to disk.
+We now describe the key ideas that make \BWT{} merging fast and space-efficient.
 
@@
 Assume that we have a \emph{node} of the trie corresponding to suffix $X$, lexicographic rank $r$, and lexicographic range $[sp,ep]$.
-As suffix $X$ occurs $sp+1-ep$ times in collection $\Bcoll$, we can output a \emph{run} of lexicographic ranks $(r, sp+1-ep)$.
+As suffix $X$ occurs $sp+1-ep$ times in collection $\Bcoll$, we can output a \emph{run} of ranks $(r, sp+1-ep)$.
 Afterwards, we proceed to the \emph{children} of the node.
-For each character $c \in \Sigma$, we create a node corresponding to suffix $cX$, lexicographic rank $\mLF_{\Acoll}(r,c)$, and lexicographic range $[\mLF_{\Bcoll}(sp-1, c) + 1, \mLF_{\Bcoll}(ep, c)]$.
+For each character $c \in \Sigma$, we create a node corresponding to suffix $cX$, rank $\mLF_{\Acoll}(r,c)$, and range $[\mLF_{\Bcoll}(sp-1, c) + 1, \mLF_{\Bcoll}(ep, c)]$.
 Searching the branches of the trie can be done in parallel using multiple \emph{threads}.
@@
 Once the thread buffer becomes full, we merge it with the global \emph{merge buffers}.
-There are $k$ merge buffers $M_{1}$ to $M_{k}$, with buffer $M_{i}$ containing $2^{i-1}$ merged thread buffers.
-The merging starts from buffer $M_{1}$.
+There are $k$ merge buffers $M_{1}$ to $M_{k}$, with buffer $M_{i}$ containing $2^{i-1}$ thread buffers.
+The merging starts from $M_{1}$.
 If $M_{i}$ is empty, the thread swaps its thread buffer with the empty buffer and returns to the search phase.
-Otherwise it swaps $M_{i}$ with an empty buffer, merges the former $M_{i}$ with its thread buffer, and proceeds to $M_{i+1}$.
-If the merging reaches $M_{k+1}$, the thread writes its thread buffer to disk and returns back to work.
+Otherwise it merges $M_{i}$ with its thread buffer, clearing $M_{i}$, and proceeds to $M_{i+1}$.
+If the a thread reaches $M_{k+1}$, it writes its thread buffer to disk and returns back to work.
 
@@
 
-We have implemented the improved merging algorithm as a tool for merging the \BWT{}s of large collections of short reads.
-The tool, \BWTmerge{}, is written in C++, and the source code is available at GitHub.\footnote{\url{https://github.com/jltsiren/bwt-merge}} The implementation is built on top of the \emph{SDSL library} \cite{Gog2014b} and uses the features of C++11 extensively.
-As a result, it needs a fairly recent C++11 compiler to compile.
+We have implemented the improved merging algorithm in a tool for merging the \BWT{}s of large collections of short reads.
+The tool, \BWTmerge{}, is written in C++, and the source code is available on GitHub.\footnote{\url{https://github.com/jltsiren/bwt-merge}} The implementation uses the \emph{SDSL library} \cite{Gog2014b} and the new features in C++11.
+As a result, it needs a fairly recent C++ compiler to compile.
 We have successfully built \BWTmerge{} on Linux and OS~X using g++.
@@
 The system should have tens of CPU cores and hundreds of gigabytes of memory.
-The amount of local disk space might not be much larger than memory size, while there can be plenty of shared disk space available.
+There should also be a few hundred gigabytes of reliable (i.e.~local) disk space for temporary files.
 The number of search threads is equal to the number of CPU cores, while the merge phase uses just one producer thread and one consumer thread.
@@
 We use two-level arrays with 8\nobreakdash-megabyte blocks to store the \BWT{}s and the compressed buffers, managing the blocks using \texttt{mmap()} and \texttt{munmap()}.
-In our experiments, this direct memory management reduced the space overhead by tens of gigabytes.
-When multiple threads allocate memory in small enough blocks, the multithreaded \emph{glibc} implementation of \texttt{malloc()} creates a number of additional heaps that will never grow larger than 64~MB.
+This reduced the space overhead by tens of gigabytes over using \texttt{malloc()} and \texttt{free()}.
+% arXiv:
+%When multiple threads allocate memory in small enough blocks, the multithreaded \emph{glibc} implementation of \texttt{malloc()} creates a number of additional heaps that will never grow larger than 64~MB.
 Each thread tries to reuse the heap it used for the last allocation.
@@
 
-We ran our experiments on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256 gigabytes of memory.
+We used a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256 gigabytes of memory.
 The system was running Ubuntu 12.04 on Linux kernel 3.2.0.
+We used a development version of \BWTmerge{}  equivalent to v0.3.
+For the other tools, we used the versions that were available on GitHub in October~2015.
 All software was compiled with gcc/g++ version 4.9.2.
 We stored the input/output files on a distributed Lustre file system and used a local 0.5~TB hard disk for temporary files.
-\BWTmerge{} was allowed to use 32 threads, while the other \BWT{} construction tools were limited to 4 or 5 parallel threads by design.
-All reported merging times include verification by querying the \BWT{}s with 2~million $32$\nobreakdash-mers extracted from the human reference genome.
+\BWTmerge{} was allowed to use 32 threads, while the other \BWT{} construction tools were limited to 4 or 5 threads by design.
+The reported merging times include verification by querying the \BWT{}s with 2~million $32$\nobreakdash-mers extracted from the human reference genome.
 
@@
 \begin{center}
-\caption{Datasets used in the experiments.
-The amount of sequence data and the number of reads in each file, and the size of the \BWT{} in the native format and in the Read Server format.
+\caption{Datasets.
+The amount of sequence data, the number of reads, and the size of the \BWT{} in the native format and in the Read Server format.
 RLO indicates that the reads are sorted in reverse lexicographic order.
@@
 \renewcommand{\baselinestretch}{1}\footnotesize
-\begin{tabular}{cc|cc|cc|cc}
+\begin{tabular}{c|cc|cc|cc}
 \hline
- & & \multicolumn{2}{c|}{\textbf{Data}} & \multicolumn{2}{c|}{\textbf{Native \BWT}} & \multicolumn{2}{c}{\textbf{Read Server}} \\
-\textbf{Dataset} & \textbf{File} & \textbf{Size} & \textbf{Reads} & \textbf{Unsorted} & \textbf{RLO} & \textbf{\BWT} & \textbf{\FMI} \\
+ & \multicolumn{2}{c|}{\textbf{Data}} & \multicolumn{2}{c|}{\textbf{Native \BWT}} &
+ \multicolumn{2}{c}{\textbf{Read Server}} \\
+\textbf{Dataset} & \textbf{Size} & \textbf{Reads} & \textbf{Unsorted} & \textbf{RLO} & \textbf{\BWT} & \textbf{\FMI} \\
 \hline
-\CEU & NA12878 &  284 Gbp & 2.81G & 50.3 GB & 25.5 GB &      -- &       -- \\
-     & NA12891 &  242 Gbp & 2.40G & 42.4 GB & 19.7 GB &      -- &       -- \\
-     & NA12892 &  245 Gbp & 2.42G & 43.8 GB & 20.7 GB &      -- &       -- \\
-     & Merged  &  771 Gbp & 7.63G &  129 GB & 58.9 GB &      -- &       -- \\
+\CEU: All           &  771 Gbp & 7.63G &  136 GB & 65.9 GB &      -- &       -- \\
+% arXiv:
+%NA12878             &  284 Gbp & 2.81G & 50.3 GB & 25.5 GB &      -- &       -- \\
+%NA12891             &  242 Gbp & 2.40G & 42.4 GB & 19.7 GB &      -- &       -- \\
+%NA12892             &  245 Gbp & 2.42G & 43.8 GB & 20.7 GB &      -- &       -- \\
+Merged              &  771 Gbp & 7.63G &  129 GB & 58.9 GB &      -- &       -- \\
 \hline
-\RS  & AA      &  433 Gbp & 4.69G &      -- & 38.5 GB & 39.9 GB &  48.3 GB \\
-     & TT      &  432 Gbp & 4.68G &      -- & 38.7 GB & 40.0 GB &  48.4 GB \\
-     & AT      &  275 Gbp & 2.98G &      -- & 26.6 GB & 26.9 GB &  32.6 GB \\
-     & TA      &  355 Gbp & 3.84G &      -- & 32.7 GB & 33.5 GB &  40.6 GB \\
-     & Merged  & 1.49 Tbp & 16.2G &      -- &  117 GB &  126 GB & (152 GB) \\
+\RS: AA, TT, AT, TA & 1.49 Tbp & 16.2G &      -- &  136 GB &  140 GB &   170 GB \\
+% arXiv:
+%AA                  &  433 Gbp & 4.69G &      -- & 38.5 GB & 39.9 GB &  48.3 GB \\
+%TT                  &  432 Gbp & 4.68G &      -- & 38.7 GB & 40.0 GB &  48.4 GB \\
+%AT                  &  275 Gbp & 2.98G &      -- & 26.6 GB & 26.9 GB &  32.6 GB \\
+%TA                  &  355 Gbp & 3.84G &      -- & 32.7 GB & 33.5 GB &  40.6 GB \\
+Merged              & 1.49 Tbp & 16.2G &      -- &  117 GB &  126 GB & (152 GB) \\
 \hline
-\RS  & *A, *C  & 2.45 Tbp & 26.5G &      -- &  225 GB &  232 GB &   281 GB \\
-     & Merged  & 2.45 Tbp & 26.5G &      -- &  181 GB &  197 GB & (239 GB) \\
+\RS: *A, *C         & 2.45 Tbp & 26.5G &      -- &  225 GB &  232 GB &   281 GB \\
+Merged              & 2.45 Tbp & 26.5G &      -- &  181 GB &  197 GB & (239 GB) \\
 \hline
-\RS  & *G, *T  & 2.44 Tbp & 26.5G &      -- &  226 GB &  232 GB &   281 GB \\
-     & Merged  & 2.44 Tbp & 26.5G &      -- &  180 GB &  197 GB & (238 GB) \\
+\RS: *G, *T         & 2.44 Tbp & 26.5G &      -- &  226 GB &  232 GB &   281 GB \\
+Merged              & 2.44 Tbp & 26.5G &      -- &  180 GB &  197 GB & (238 GB) \\
 \hline
 \end{tabular}}
-\end{center}
+\end{center}%
+\vspace{-16pt}
 \end{table}
 
-We used two datasets from phase 3 of the \emph{1000 Genomes Project} \cite{1000GP2015}:
-\begin{itemize}
+We used two datasets from phase 3 of the \emph{1000 Genomes Project} \cite{1000GP2015}.
+\CEU{} contains 101~bp reads from high-coverage sequencing of the \emph{CEU trio} (individuals NA12878, NA12891, and NA12892).
+We downloaded the gzipped FASTQ files (run accessions SRR622457, SRR622458, and SRR622459).
+For each individual, we concatenated the files and corrected the sequencing errors with BFC \cite{Li2015} (\texttt{bfc -s 3g -t 16}).
+\RS{} is from the \emph{Read Server} project, which uses all low-coverage and exome data from the phase 3.
+After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp.
+The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} (SGA) \cite{Simpson2012}, partitioned by the last two bases.
+See Table~\ref{table:datasets} for further details on the datasets.
 
-\item \CEU{} contains 101~bp reads from high-coverage sequencing of the \emph{CEU trio} (individuals NA12878, NA12891, and NA12892).
-We downloaded the gzipped FASTQ files (run accessions SRR622457, SRR622458, and SRR622459) and concatenated the files for each individual.
-Then we corrected the sequencing errors with BFC \cite{Li2015} (\texttt{bfc -s 3g -t 16}) separately for each individual.
-
-\item \RS{} is from the \emph{Read Server} project, which uses all low-coverage and exome data from the phase 3.
-After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp.
-The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} (SGA) \cite{Simpson2012}, partitioned according to the last two bases.
-
-\end{itemize}
-See Table~\ref{table:datasets} for further details on the datasets.
-We used a development version of \BWTmerge{} that was essentially equivalent to v0.3 for the experiments.
-For the other tools, we used the versions that were available on GitHub in October~2015.
-
-\smallbreak\noindent\textbf{Parameters.} For benchmarking with different parameter values, we took four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset, and converted them to the \emph{native format} of \BWTmerge.
-This format includes the \BWT{} and the \rank/\select{} structures required by the FM-index.
+\smallbreak\noindent\textbf{Parameters.} For testing different parameter values, we took four \BWT{} files (AA, TT, AT, and TA) containing a total of 1.49~Tbp from the \RS{} dataset, and converted them to the \emph{native format} of \BWTmerge.
+This format includes the \rank/\select{} structures required by the FM-index.
 We then merged the \BWT{}s (in the given order).
 We used 128~MB or 256~MB run buffers and 256~MB or 512~MB thread buffers.
-The number of merge buffers was 4 or 5 with 512~MB thread buffers and 5 or 6 with 256~MB thread buffers, so that the files on disk were always merged from either 8~GB or 16~GB of thread buffers.
+The number of merge buffers was chosen so that the files on disk were always merged from either 8~GB or 16~GB of thread buffers.
 
@@
 \includegraphics{comparison.pdf}
-\end{center}
-\caption{Left: Time/space trade-offs for merging four \BWT{} files containing a total of 1.49~Tbp.
+\end{center}%
+\vspace{-12pt}
+\caption{Time/space trade-offs.
+Left: Merging four \BWT{} files (1.49~Tbp).
 Label rXbYmZ denotes $X$~MB run buffers, $Y$~MB threads buffers, and $Z$ merge buffers.
-The dashed line represents the total size of the last pair of \BWT{}s to be merged.
-Right: Time/space trade-offs for building the \BWT{} of the 771~Gbp \CEU{} dataset.
-RLO indicates reverse lexicographic order.}\label{fig:benchmark}
+The dashed line marks the total size of the last pair of \BWT{}s to be merged.
+Right: Building the \BWT{} of the 771~Gbp \CEU{} dataset.
+RLO indicates reverse lexicographic order.}\label{fig:benchmark}%
+\vspace{-6pt}
 \end{figure}
@@
 Memory overhead was 21.1 to 41.5~GB on top of the 124.2~GB required by the last pair of \BWT{}s.
-The temporary files used 287 to 306~gigabytes of disk space.
+% arXiv:
+%The temporary files used 287 to 306~gigabytes of disk space.
 Thread buffer size and the number of merge buffers were the most important parameters.
@@
 Increasing run buffer size to 256~MB made the search phase faster with 512~MB thread buffers and slower with 256~MB thread buffers.
-
+%
 For the further experiments, we chose 128~MB run buffers, 256~MB thread buffers, and 6 merge buffers (overhead 30.8~GB).
 
-\smallbreak\noindent\textbf{Comparison.} In the next experiment, we compared the performance of \BWTmerge{} to the \ropebwt{} \cite{Li2011-2013} implementation of the BCR algorithm \cite{Bauer2013} and to \ropebwtii{} \cite{Li2014a}.
-These two are the fastest existing \BWT{} construction tools for read collections on general hardware \cite{Li2014a}.
-We ran \ropebwt{} with parameters \texttt{-btORf -abcr} and \ropebwtii{} with parameters \texttt{-bRm10g}.
-For some runs, we also used \texttt{-s} with both tools to sort the reads in \emph{reverse lexicographic order} (RLO) to improve compression.
-Due to memory constraints, we used the smaller \CEU{} dataset.
-
-We built the \BWT{} of the dataset in several ways.
-We concatenated all input files and built the \BWT{} directly for the entire collection using both \ropebwt{} and \ropebwtii{}.
-We also built the \BWT{} for each input file separately using \ropebwt{}, and then merged the results using \BWTmerge{}.
-Finally, we repeated everything with the reads sorted in reverse lexicographic order.
-In each case, the tools were allowed to output the \BWT{} files in their preferred formats.
+\smallbreak\noindent\textbf{Comparison.} In the next experiment, we compared \BWTmerge{} to the fastest \BWT{} construction tools on general hardware \cite{Li2014a}.
+We built the \BWT{} of the \CEU{} dataset using \ropebwt{} \cite{Li2011-2013} with parameters \texttt{-btORf -abcr} and \ropebwtii{} \cite{Li2014a} with parameters \texttt{-bRm10g}.
+We also built individual \BWT{}s using \ropebwt{} and merged them using \BWTmerge.
+All tools wrote the \BWT{} files in their preferred formats.
 
@@
 When the reads are in the original order, \BWTmerge{} is 1.85x slower and 1.46x more space-efficient than building the final \BWT{} directly with \ropebwt.
-\ropebwt2{} ran out of memory when processing the last 10~GB batch of reads.
-\BWTmerge{} would have been about 1.2x slower and 1.5x more space-efficient than it.
-The running time of \BWTmerge{} was split almost evenly between building the \BWT{}s of the individual files and merging them.
-
-When we tell \ropebwt{} and \ropebwt2{} to sort the reads in reverse lexicographic order, all tools improve their performance.
+\ropebwt2{} ran out of memory just before finishing.
+\BWTmerge{} was about 1.2x slower and 1.5x more space-efficient than it.
+The running time of \BWTmerge{} was split almost evenly between building the individual \BWT{}s and merging them.
+% arXiv:
+%
+When \ropebwt{} and \ropebwt2{} sort the reads in \emph{reverse lexicographic order} (RLO, parameter \texttt{-s}) to improve compression, all tools improve their performance.
 \BWTmerge{} is now 1.70x slower and 2.12x more space-efficient than \ropebwt{} and 1.21x \emph{faster} and 1.09x more space-efficient than \ropebwt2.
-Again, \BWTmerge{} spent around half of the time building the individual \BWT{}s and another half merging them.
+% arXiv:
+%Again, \BWTmerge{} spent around half of the time building the individual \BWT{}s and another half merging them.
 Note that \BWTmerge{} builds a \BWT{} for the concatenation of three input files that are in RLO, while the \BWT{}s produced by the other tools are completely in RLO.
@@
 \smallbreak\noindent\textbf{Read Server.} In the last experiment, we merged the 16 \BWT{} files in the \RS{} dataset into two files (AA, CA, TA, GA, AC, CC, GC, and TC into the first file; TT, GT, CT, AT, TG, GG, CG, AG into the second one).
-Merging the \BWT{}s took 81.3~hours and 83.0~hours, required 221~gigabytes and 219~gigabytes of memory, and used 297~gigabytes and 300~gigabytes of disk space, respectively.
+Merging the \BWT{}s took 81.3~hours and 83.0~hours, required 221~GB and 219~GB of memory, and used 297~GB and 300~GB of disk space, respectively.
 This reduced the size of the FM-indexes from around 560~GB to 480~GB.
@@
 
-\Section{Discussion}
+\Section{Conclusions}
 
@@
 
-Read Server stores the reads in \emph{reverse lexicographic order} to improve compression \cite{Cox2012}.
+% arXiv:
+%Read Server stores the reads in \emph{reverse lexicographic order} to improve compression \cite{Cox2012}.
 We can maintain this order with a few changes to the search phase \cite{Li2014a}.
@@
 With \emph{lexicographic order}, we can determine $\mSA[i]$ without samples by using $\mLF$ instead of $\Psi$, which is often faster in practice.
-
-We can also sort the reads by their likely positions in a reference genome.
+%
+%We can also sort the reads by their likely positions in a reference genome.
 This \emph{position order} is useful for both compression and storing the \emph{pairing information} for the reads.

### 2015-11-06 paper/paper.tex
DCC 2016 submission.

@@
 \title
-{\large
+{\vspace{-12pt}\large
 \textbf{Burrows-Wheeler transform for terabases}
@@
 
-The dramatic decrease in the cost of DNA sequencing has flooded the world with \emph{sequence data}.
-The \emph{1000 Genomes Project} \cite{1000GP2015} sequenced the genomes of over 2500 humans, and there are many other projects that are similar or greater in scale.
-A sequencing machine produces a large number of \emph{reads} (short sequences) that cover the sequenced genome many times over.
+The decrease in the cost of DNA sequencing has flooded the world with \emph{sequence data}.
+The \emph{1000 Genomes Project} \cite{1000GP2015} sequenced the genomes of over 2500 humans, and there are other projects that are similar or greater in scale.
+A sequencing machine produces a large number of \emph{reads} (short sequences) that cover the genome many times over.
 For a 3~Gbp human genome, the total length of the reads is often 100~Gbp or more.
@@
 
-The \emph{Read Server} project at the Wellcome Trust Sanger Institute aims to avoid the reference bias entirely by developing tools for large-scale \emph{reference-free} genome analysis.
-Unique reads are compressed and indexed using the \emph{Burrows-Wheeler transform} (\BWT) \cite{Burrows1994}, while metadata databases contain information about the original reads.
-Initially, the project works with the low-coverage and exome data from phase 3 of the 1000 Genomes Project.
-After error correction and trimming the reads to either 73~bp or 100~bp, the 86~Tbp of raw data consisting of 922~billion original reads is reduced to 4.88~Tbp and 53.0~billion unique sequences.
+Preprocessing large datasets can take weeks.
+It is often not feasible to rebuild everything when new methods of analysis require new functionalities.
+Structures based on the \emph{Burrows-Wheeler transform} (\BWT) are often useful due to their versatility.
+A \emph{run-length encoded} \BWT{} compresses repetitive sequence collections quite well \cite{Maekinen2010}, while the similarities to the suffix tree and the suffix array make \BWT-based indexes suitable for many \emph{pattern matching} and \emph{sequence analysis} tasks \cite{Ohlebusch2013,Maekinen2015}.
+
+The \emph{Read Server} project at the Sanger Institute develops tools for large-scale \emph{reference-free} genome analysis, avoiding reference bias.
+Unique reads are compressed and indexed using the \BWT, while metadata databases contain information on the original reads.
+Initially, the project works with
+% arXiv:
+%the low-coverage and exome data from phase 3 of the 1000 Genomes Project.
+% DCC:
+data from the 1000 Genomes Project.
+After error correction and trimming the reads to either 73~bp or 100~bp, the 922~billion original reads (86~Tbp) are reduced to 53.0~billion unique sequences (4.88~Tbp).
 These sequences are stored in 16 \BWT-based indexes \cite{Ferragina2005a} taking a total of 561.5~gigabytes.
 
-Preprocessing large datasets can take days or even weeks.
-It is often not feasible to rebuild everything when new methods of analysis require new functionalities from the indexes.
-\BWT-based methods were chosen due to their versatility.
-A \emph{run-length encoded} \BWT{} compresses repetitive sequence collections quite well \cite{Maekinen2010}, while the similarities to the suffix tree and the suffix array make \BWT-based indexes suitable for a wide variety of \emph{pattern matching} and \emph{sequence analysis} tasks \cite{Ohlebusch2013,Maekinen2015}.
-
-Read Server partitions the unique reads based on the last two bases of the read.
-Every query must be repeated in all 16 indexes, and the \BWT{}s require more space, as we cannot compress the similarities between the reads in different indexes.
+The unique reads are partitioned between the \BWT{}s by the last two bases.
+Every query must be repeated in all 16 indexes.
+The \BWT{}s also require more space, as we cannot compress the similarities between the reads in different indexes.
 Reducing the number of indexes would improve both memory usage and query performance.
-However, the lack of \BWT{} construction algorithms that work with terabases of sequence data has made it infeasible so far.
+This requires \BWT{} construction algorithms that can handle terabases of data.
 
@@
 
-The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} using a fast general-purpose algorithm \cite{Mori2008,Nong2011}.
-Given the sequences and their suffix array, we can easily build the \BWT{} of the sequences.
-These algorithms cannot be used with large datasets, as they require several times more memory than the sequences themselves.
+The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} using a fast general-purpose algorithm \cite{Mori2008,Nong2011}, and to derive the \BWT{} from the suffix array.
+These algorithms cannot be used with large datasets, as they require much more memory than the sequences themselves.
 Suffix arrays can be built on disk \cite{Gonnet1992}, but even the fastest algorithms cannot index the data faster than 1\nobreakdash--2~Mbp/s
@@
 
-There are algorithms for constructing the \BWT{} \emph{directly} without having to build the suffix array first.
-Some require only a limited amount of working space on top of a plain or compressed representation of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}, or use the disk as additional working space \cite{Ferragina2012,Beller2013}.
-These general-purpose algorithms rarely exceed 1\nobreakdash--2~Mbp/s with large datasets.
+There are many \emph{direct} \BWT{} construction algorithms that do not need the suffix array.
+Some require a limited amount of working space on top of the \BWT{} \cite{Hon2007,Kaerkkaeinen2007,Siren2009,Okanohara2009}, while others use the disk as additional working space \cite{Ferragina2012,Beller2013}.
+These general-purpose algorithms rarely exceed 1\nobreakdash--2~Mbp/s.
 \emph{Specialized algorithms} for DNA sequences achieve better time/space trade-offs.
-Some algorithms can index 5\nobreakdash--10~Mbps using ordinary hardware, with their memory usage becoming a bottleneck for datasets larger than about 1~Tbp \cite{Bauer2013,Li2014a}.
-GPU-based algorithms can be much faster, but their memory usage is also higher \cite{Liu2014,Pantaleoni2014}.
-Distributing the \BWT{} construction to multiple nodes can remove the obvious bottlenecks, but such algorithms generally use the hardware less efficiently than those running on a single system \cite{Wang2015}.
+Some can index 5\nobreakdash--10~Mbps using ordinary hardware, with their memory usage becoming the bottleneck after about 1~Tbp \cite{Bauer2013,Li2014a}.
+GPU-based algorithms are even faster, but their memory usage is also higher \cite{Liu2014,Pantaleoni2014}.
+Distributing the \BWT{} construction to multiple nodes can remove the obvious bottlenecks, at the price of using more resources for the construction \cite{Wang2015}.
 
@@
 
-If suffix $X$ precedes suffix $Y$ in lexicographic order, suffix $cX$ will also precede suffix $cY$, for any character $c$.
-If $\mBWT[i]$ is the $j$\nobreakdash-th occurrence of character $c$ in the \BWT{} and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order.
-This forms the essence of \LF\emph{\nobreakdash-mapping}.
+If $X \le Y$ in lexicographic order, we also have $cX \le cY$ for any character $c$.
+If $\mBWT[i] = c$ is the $j$\nobreakdash-th occurrence of $c$ in the \BWT{} and $\mSA[i]$ points to suffix $X$, suffix $cX$ is the $j$\nobreakdash-th suffix starting with $c$ in lexicographic order.
+
 Let $\mC[c]$ bet the number of suffixes starting with a character smaller than $c$, and let $S.\mrank(i,c)$ be the number of occurrences of $c$ in the prefix $S[1,i]$.
-Then $\mSA[\mLF(i)]$, where
-$$
-\mLF(i) = \mLF(i,\mBWT[i])
-\,\, \textrm{and} \,\,
-\mLF(i,c) = \mC[c] + \mBWT.\mrank(i, c),
-$$
-points to the suffix $T[\mSA[i]-1,n]$.
-The more general version of \LF-mapping $\mLF(i,c)$ tells the number of suffixes $X$ of text $T$ with $X \le c T[\mSA[i],n]$.
-This is the \emph{lexicographic rank} $\mrank(cT[\mSA[i],n], T)$ of $cT[\mSA[i],n]$ among the suffixes of text $T$.
+We define \LF-\emph{mapping} as $\mLF(i,c) = \mC[c] + \mBWT.\mrank(i, c)$ and $\mLF(i) = \mLF(i,\mBWT[i])$.
+The general form $\mLF(i,c)$ is the number of suffixes $X$ of text $T$ with $X \le cT[\mSA[i],n]$.
+This is known as the  \emph{lexicographic rank} $\mrank(cT[\mSA[i],n],T)$ of text $cT[\mSA[i],n]$ among the suffixes of text $T$.
+The specific form $\mLF(i)$ gives the lexicographic rank of the previous suffix ($\mSA[\mLF(i)] = \mSA[i]-1$, if $\mSA[i] > 1$, and $\mSA[\mLF(i)] = n$ otherwise).
 
-The \emph{FM-index} (\FMI) \cite{Ferragina2005a} uses the Burrows-Wheeler transform as a full-text index by augmenting it with a few additional structures.
-The lexicographic range $[sp,ep]$ corresponding to pattern $P$ is found by using \emph{backward searching}.
-Let $[sp_{i},ep_{i}]$ be the lexicographic range of suffixes of text $T$ matching suffix $P[i, \abs{P}]$ of the pattern.
+The \emph{FM-index} (\FMI) \cite{Ferragina2005a} is a full-text index based on the \BWT{}.
+We use \emph{backward searching} in the FM-index to find the lexicographic range $[sp,ep]$ matching pattern $P$.
+Let $[sp_{i},ep_{i}]$ be the range of suffixes of text $T$ matching suffix $P[i, \abs{P}]$ of the pattern.
 We find $[sp_{i-1},ep_{i-1}]$ as $[\mLF(sp_{i}-1, P[i-1]) + 1, \mLF(ep_{i}, P[i-1])]$.
@@
 We can \emph{extract} text $T_{i}$ in $\Oh(\abs{T_{i}} \cdot t_{r})$ time by using \LF\nobreakdash-mapping \cite{Burrows1994}.
-We start from the endmarker at $\mBWT[i]$ and extract the text backwards as $T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)]$ for $1 \le j < \abs{T_{i}}$.
+We start from the endmarker at $\mBWT[i]$ and extract the text backwards as $T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)]$, for $1 \le j < \abs{T_{i}}$.
 As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also find the lexicographic ranks of all suffixes of text $T_{i}$ in the process.
@@
 Then we determine the lexicographic rank $j = \mrank(cT, T) = \mC[c] + \mBWT.\mrank(i, c)$ of text $cT$ among the suffixes of text $T$.
-Finally \emph{replace} the endmarker at $\mBWT[i]$ with the inserted character $c$ and \emph{insert} a new endmarker between $\mBWT[j]$ and $\mBWT[j+1]$.
+Finally we \emph{replace} the endmarker at $\mBWT[i]$ with the inserted character $c$ and \emph{insert} a new endmarker between $\mBWT[j]$ and $\mBWT[j+1]$.
 
 We can use the transformation for \BWT{} construction in several ways.
-We can do \emph{batch updates} and transform the \BWT{} of text $T$ into that of text $XT$, where $X$ is a string \cite{Hon2007}.
+We can use \emph{batch updates} and transform the \BWT{} of text $T$ into that of text $XT$, where $X$ is a string \cite{Hon2007}.
 We can start with the \BWT{}s of text collections $\Acoll$ and $\Bcoll$, and \emph{merge} them into the \BWT{} of collection $\Acoll \cup \Bcoll$ \cite{Siren2009}.
@@
 
-\smallbreak\noindent\textbf{Search.} We search for all texts of collection $\Bcoll$ in $\mBWT_{\Acoll}$, and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of collection $\Bcoll$.
+\smallbreak\noindent\textbf{Search.} We search for all texts of collection $\Bcoll$ in $\mBWT_{\Acoll}$, and output the lexicographic rank $\mrank(X, \Acoll)$ for each suffix $X$ of $\Bcoll$.
 This takes $\Oh(n_{\Bcoll} t_{r})$ time.
-If we do not have collection $\Bcoll$ in plain form, we can extract the texts from $\mBWT_{\Bcoll}$ in the same asymptotic time.
+We either need the collection in plain form, or extract the texts from $\mBWT_{\Bcoll}$ in the same asymptotic time.
 
@@
 If we split a text collection $\Acoll$ of total length $n$ into $p$ \emph{subcollections} of equal size, we can build $\mBWT_{\Acoll}$ incrementally by merging the \BWT{}s of the subcollections.
-This takes $\Oh((p+t_{r})n)$ time and uses essentially $\min(n, \frac{n}{p} \log n)$ bits of working space.
+This takes $\Oh((p+t_{r})n)$ time and uses essentially $\min(\frac{n}{p} \log n, n)$ bits of working space.
 
-When the collection is large, the space overhead of the construction algorithm often determines whether we can build the \BWT{} of the collection.
+When the collection is large, the space overhead of the construction algorithm often determines whether we can build the \BWT{}.
 % arXiv:
@@
 %
-We can reduce the overhead by writing the lexicographic ranks to \emph{disk} in the search phase.
-If we can sort the ranks on disk, we just need to scan the rank array once during the merge phase.
-As fast disk space is often scarce, we may want to \emph{interleave} the sorting with the search and merge phases, and to \emph{compress} the ranks before writing them to disk.
-We now describe the key ideas that make \BWT{} merging fast and space-efficient.
+We can reduce the overhead by writing the lexicographic ranks to \emph{disk}.
+If we sort the ranks on disk, we just need to scan the rank array once during the merge phase.
+We can also \emph{compress} the ranks before writing them to disk and \emph{interleave} the sorting with the search and merge phases.
+We now describe the key ideas for fast and space-efficient \BWT{}.
 
@@
 
-\smallbreak\noindent\textbf{Merge.} The merge phase starts with the rank array stored in multiple files on disk.
-Each of the files is sorted and compressed.
-For interleaving the \BWT{}s, we need to merge the files to scan through the rank array.
+\smallbreak\noindent\textbf{Merge.} The ranks are stored in sorted order in multiple files on disk.
+For interleaving the \BWT{}s, we need to merge the files and to scan through the rank array.
 We can also use multiple threads here.
@@
 
-We have implemented the improved merging algorithm in a tool for merging the \BWT{}s of large collections of short reads.
+We have implemented the improved merging algorithm as a tool for merging the \BWT{}s of large read collections.
 The tool, \BWTmerge{}, is written in C++, and the source code is available on GitHub.\footnote{\url{https://github.com/jltsiren/bwt-merge}} The implementation uses the \emph{SDSL library} \cite{Gog2014b} and the new features in C++11.
@@
 The target environment of \BWTmerge{} is a \emph{single node} of a \emph{computer cluster}.
-The system should have tens of CPU cores and hundreds of gigabytes of memory.
-There should also be a few hundred gigabytes of reliable (i.e.~local) disk space for temporary files.
+The system should have tens of CPU cores, hundreds of gigabytes of memory, and hundreds of gigabytes of local disk space for temporary files.
 The number of search threads is equal to the number of CPU cores, while the merge phase uses just one producer thread and one consumer thread.
-By adjusting the sizes of run buffers and thread buffers and the number of merge buffers, \BWTmerge{} should work reasonably well in different environments.
+\BWTmerge{} can be adapted to many other environments by adjusting the number and the size of the buffers.
 
@@
 
-For \rank/\select{} support, we split the \BWT{}s into 64\nobreakdash-byte blocks of compressed data, ensuring that no run continues from one block to the next one.
+For \rank/\select{} support, we divide the \BWT{}s into 64\nobreakdash-byte blocks of compressed data, ensuring that the runs do not cross block boundaries.
 For each block $i$, we store the total number of characters in blocks $1$ to $i-1$ as $n_{i}$, as well as the cumulative character counts $c_{i} = \mBWT.\mrank(n_{i},c)$ for $0 \le c \le 5$.
-Each of these seven increasing sequences is encoded using the \emph{sdarray} encoding \cite{Okanohara2007}.
-To compute $\mBWT.\mrank(j,c)$, we first use a \rank{} query on the $n_{i}$ sequence to find the correct block.
-A \select{} query on the same sequence transforms the position $j$ into an offset within the block, while a \select{} query on the $c_{i}$ sequence gives the rank up to the beginning of the block.
-Finally, we decompress the block to answer the query.
-\select{} queries and random accesses to the \BWT{} work in a similar way.
-There are also optimizations for e.g.~computing $\mrank(i,c)$ for all characters $c$, and for finding the children of a reverse trie node corresponding to a relatively short lexicographic range.
+These increasing sequences are stored using the \emph{sdarray} encoding \cite{Okanohara2007}.
+To compute $\mBWT.\mrank(j,c)$, we start with a \rank{} query on the $n_{i}$ sequence to find the block.
+A \select{} query on the same sequence transforms $j$ into a block offset, while a \select{} query on the $c_{i}$ sequence gives the rank up to the beginning of the block.
+We then decompress the block to answer the query.
+\select{} queries and accessing the \BWT{} work in a similar way.
+There are also optimizations for e.g.~computing $\mrank(i,c)$ for all characters $c$, and for finding the children of a reverse trie node corresponding to a short lexicographic range.
 
 We use two-level arrays with 8\nobreakdash-megabyte blocks to store the \BWT{}s and the compressed buffers, managing the blocks using \texttt{mmap()} and \texttt{munmap()}.
-This reduced the space overhead by tens of gigabytes over using \texttt{malloc()} and \texttt{free()}.
+This reduces the space overhead by tens of gigabytes over using \texttt{malloc()} and \texttt{free()}.
 % arXiv:
@@
 The system was running Ubuntu 12.04 on Linux kernel 3.2.0.
-We used a development version of \BWTmerge{}  equivalent to v0.3.
-For the other tools, we used the versions that were available on GitHub in October~2015.
+We used a development version of \BWTmerge{}  equivalent to v0.3, and the versions of the other tools that were available on GitHub in October 2015.
 All software was compiled with gcc/g++ version 4.9.2.
-We stored the input/output files on a distributed Lustre file system and used a local 0.5~TB hard disk for temporary files.
-\BWTmerge{} was allowed to use 32 threads, while the other \BWT{} construction tools were limited to 4 or 5 threads by design.
-The reported merging times include verification by querying the \BWT{}s with 2~million $32$\nobreakdash-mers extracted from the human reference genome.
+We stored the input/output files on a distributed Lustre file system and used a local 0.5~TB disk for temporary files.
+\BWTmerge{} used 32 threads, while the other \BWT{} construction tools were limited by design to 4 or 5 threads.
+% arXiv:
+%The merging times include verification by querying the \BWT{}s with 2~million $32$\nobreakdash-mers.
 
@@
 
-We used two datasets from phase 3 of the \emph{1000 Genomes Project} \cite{1000GP2015}.
+Our datasets come from phase 3 of the \emph{1000 Genomes Project} \cite{1000GP2015}.
 \CEU{} contains 101~bp reads from high-coverage sequencing of the \emph{CEU trio} (individuals NA12878, NA12891, and NA12892).
@@
 \RS{} is from the \emph{Read Server} project, which uses all low-coverage and exome data from the phase 3.
-After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp.
-The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} (SGA) \cite{Simpson2012}, partitioned by the last two bases.
+There are 53.0~billion unique reads for a total of 4.88~Tbp.
+The reads are in 16 run-length encoded \BWT{}s built by using the \emph{String Graph Assembler} (SGA) \cite{Simpson2012}, partitioned by the last two bases.
 See Table~\ref{table:datasets} for further details on the datasets.
@@
 The results can be seen in Figure~\ref{fig:benchmark} (left).
-The average speed for inserting 1.06~Tbp into file AA ranged from 8.27 to 9.40~Mbp/s, depending on the parameter values.
-Memory overhead was 21.1 to 41.5~GB on top of the 124.2~GB required by the last pair of \BWT{}s.
+The average speed for inserting 1.06~Tbp into file AA ranged from 8.27~Mbp/s to 9.40~Mbp/s, depending on the parameter values.
+Memory overhead was 21.1~GB to 41.5~GB on top of the 124.2~GB required by the last pair of \BWT{}s.
 % arXiv:
@@
 We built the \BWT{} of the \CEU{} dataset using \ropebwt{} \cite{Li2011-2013} with parameters \texttt{-btORf -abcr} and \ropebwtii{} \cite{Li2014a} with parameters \texttt{-bRm10g}.
-We also built individual \BWT{}s using \ropebwt{} and merged them using \BWTmerge.
-All tools wrote the \BWT{} files in their preferred formats.
+We also built individual \BWT{}s using \ropebwt{} and merged them with \BWTmerge.
+All tools were set to write the \BWT{}s in their preferred formats.
 
-The results can be seen in Figure~\ref{fig:benchmark} (right).
-When the reads are in the original order, \BWTmerge{} is 1.85x slower and 1.46x more space-efficient than building the final \BWT{} directly with \ropebwt.
+The results are in Figure~\ref{fig:benchmark} (right).
+When the reads are in the original order, \BWTmerge{} is 1.85x slower and 1.46x more space-efficient than \ropebwt.
 \ropebwt2{} ran out of memory just before finishing.
-\BWTmerge{} was about 1.2x slower and 1.5x more space-efficient than it.
-The running time of \BWTmerge{} was split almost evenly between building the individual \BWT{}s and merging them.
+It would have been about 1.2x faster and 1.5x less space-efficient than \BWTmerge.
+The running time of \BWTmerge{} was split evenly between \BWT{} construction and merging.
 % arXiv:
 %
-When \ropebwt{} and \ropebwt2{} sort the reads in \emph{reverse lexicographic order} (RLO, parameter \texttt{-s}) to improve compression, all tools improve their performance.
-\BWTmerge{} is now 1.70x slower and 2.12x more space-efficient than \ropebwt{} and 1.21x \emph{faster} and 1.09x more space-efficient than \ropebwt2.
+When \ropebwt{} and \ropebwt2{} sort the reads in \emph{reverse lexicographic order} (RLO) to improve compression, all tools improve their performance.
+\BWTmerge{} becomes 1.70x slower and 2.12x more space-efficient than \ropebwt{}, and 1.21x \emph{faster} and 1.09x more space-efficient than \ropebwt2.
 % arXiv:
@@
 By converting the \BWT{}s to the native format of \BWTmerge{}, we further reduced the size of the indexes to 360~GB.
-This makes it possible to host the indexes on two servers instead of the original three.
+% arXiv:
+%This makes it possible to host the indexes on two servers instead of the original three.
 
@@
 
-\BWTmerge{} can be used as a building block of a \BWT{} construction algorithm.
-We split the read collection into multiple subcollections, build the \BWT{}s of the subcollections with other tools, and merge the results.
+\BWTmerge{} can be used as a part of a \BWT{} construction algorithm.
+We split the read collection into subcollections, build the \BWT{}s of the subcollections, and merge the results.
 The resulting algorithm is typically slower but more space-efficient than the existing algorithms.

### 2016-01-05 paper/paper.tex
Changes to the final DCC paper.

@@
 Wellcome Trust Sanger Institute \\
-Wellcome Trust Genome Campus \\
+Wellcome Genome Campus \\
 Hinxton, Cambridge, CB10 1SA, UK \\
@@
 
-The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} using a fast general-purpose algorithm \cite{Mori2008,Nong2011}, and to derive the \BWT{} from the suffix array.
+The most straightforward approach to \BWT{} construction is to build a \emph{suffix array} using a fast general-purpose algorithm \cite{Mori2008,Nong2011}, and then derive the \BWT{} from the suffix array.
 These algorithms cannot be used with large datasets, as they require much more memory than the sequences themselves.
@@
 \emph{Specialized algorithms} for DNA sequences achieve better time/space trade-offs.
-Some can index 5\nobreakdash--10~Mbps using ordinary hardware, with their memory usage becoming the bottleneck after about 1~Tbp \cite{Bauer2013,Li2014a}.
+Some can index 5\nobreakdash--10~Mbp/s using ordinary hardware, with their memory usage becoming the bottleneck after about 1~Tbp \cite{Bauer2013,Li2014a}.
 GPU-based algorithms are even faster, but their memory usage is also higher \cite{Liu2014,Pantaleoni2014}.
@@
 When the texts are short (e.g.~reads), there are more space-efficient alternatives to sampling.
-Because all endmarkers have distinct values during sorting, we know that $\mSA[i]$ (with $i \le m)$ points to the end of text $T_{i}$.
+Because all endmarkers have distinct values during sorting, we know that $\mSA[i]$ with $i \le m$ points to the end of text $T_{i}$.
 To find the end, we iterate $\Psi(i) = \mBWT.\mselect(i - \mC[c], c)$, where $c$ is the largest value with $\mC[c] < i$ and $S.\mselect(i,c)$ finds the $i$\nobreakdash-th occurrence of character $c$ in string $S$.
@@
 We can \emph{extract} text $T_{i}$ in $\Oh(\abs{T_{i}} \cdot t_{r})$ time by using \LF\nobreakdash-mapping \cite{Burrows1994}.
-We start from the endmarker at $\mBWT[i]$ and extract the text backwards as $T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)]$, for $1 \le j < \abs{T_{i}}$.
+We start from the endmarker at $\mBWT[i]$ and extract the text backwards as $T_{i}[\abs{T_{i}} - j] = \mBWT[\LF^{j-1}(i)]$, for $1 \le j \le \abs{T_{i}}$.
 As $\mSA[\LF^{j}(i)]$ points to suffix $T_{i}[\abs{T_{i}}-j, \abs{T_{i}}]$, we also find the lexicographic ranks of all suffixes of text $T_{i}$ in the process.
@@
 We can also \emph{compress} the ranks before writing them to disk and \emph{interleave} the sorting with the search and merge phases.
-We now describe the key ideas for fast and space-efficient \BWT{}.
+We now describe the key ideas for fast and space-efficient \BWT{} construction.
 
@@
 Assume that we have a \emph{node} of the trie corresponding to suffix $X$, lexicographic rank $r$, and lexicographic range $[sp,ep]$.
-As suffix $X$ occurs $sp+1-ep$ times in collection $\Bcoll$, we can output a \emph{run} of ranks $(r, sp+1-ep)$.
+As suffix $X$ occurs $ep+1-sp$ times in collection $\Bcoll$, we can output a \emph{run} of ranks $(r, ep+1-sp)$.
 Afterwards, we proceed to the \emph{children} of the node.

### 2016-01-14 paper/paper.tex
Final pdf for DCC.
Revised full/arXiv version of the paper.

@@
 \usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 
@@
 % arXiv:
-%the low-coverage and exome data from phase 3 of the 1000 Genomes Project.
+the low-coverage and exome data from phase 3 of the 1000 Genomes Project.
 % DCC:
-data from the 1000 Genomes Project.
+%data from the 1000 Genomes Project.
 After error correction and trimming the reads to either 73~bp or 100~bp, the 922~billion original reads (86~Tbp) are reduced to 53.0~billion unique sequences (4.88~Tbp).
@@
 % DCC:
-\cite{Kaerkkaeinen2015a}.
+%\cite{Kaerkkaeinen2015a}.
 % arXiv:
-%\cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Nong2015,Kaerkkaeinen2015a,Liu2015a}.
+\cite{Bingmann2013,Kaerkkaeinen2014a,Nong2014,Nong2015,Kaerkkaeinen2015a,Liu2015a}.
 
@@
 % arXiv:
-%Depending on the way the samples are selected, we may need a binary sequence to mark the pointers that have been sampled.
+Depending on the way the samples are selected, we may need a binary sequence to mark the pointers that have been sampled.
 
@@
 % arXiv:
-%Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large.
+Even if a static encoding of the \BWT{} fits in memory, a dynamic encoding may already be too large.
 The space overhead from the rank array or the interleaving bitvector (or their equivalents in the other space-efficient algorithms) may also be too much.
 We can make the rank array fit in memory by increasing the number of subcollections, but that can make the construction too slow.
-%
+
+% DCC: continues the second paragraph of the section
 We can reduce the overhead by writing the lexicographic ranks to \emph{disk}.
@@
 % arXiv:
-%When multiple threads allocate memory in small enough blocks, the multithreaded \emph{glibc} implementation of \texttt{malloc()} creates a number of additional heaps that will never grow larger than 64~MB.
+When multiple threads allocate memory in small enough blocks, the multithreaded \emph{glibc} implementation of \texttt{malloc()} creates a number of additional heaps that will never grow larger than 64~MB.
 Each thread tries to reuse the heap it used for the last allocation.
@@
 % arXiv:
-%The merging times include verification by querying the \BWT{}s with 2~million $32$\nobreakdash-mers.
+The merging times include verification by querying the \BWT{}s with 2~million $32$\nobreakdash-mers.
 
@@
 \CEU: All           &  771 Gbp & 7.63G &  136 GB & 65.9 GB &      -- &       -- \\
-% arXiv:
-%NA12878             &  284 Gbp & 2.81G & 50.3 GB & 25.5 GB &      -- &       -- \\
-%NA12891             &  242 Gbp & 2.40G & 42.4 GB & 19.7 GB &      -- &       -- \\
-%NA12892             &  245 Gbp & 2.42G & 43.8 GB & 20.7 GB &      -- &       -- \\
+% arXiv: (individual datasets)
+NA12878             &  284 Gbp & 2.81G & 50.3 GB & 25.5 GB &      -- &       -- \\
+NA12891             &  242 Gbp & 2.40G & 42.4 GB & 19.7 GB &      -- &       -- \\
+NA12892             &  245 Gbp & 2.42G & 43.8 GB & 20.7 GB &      -- &       -- \\
 Merged              &  771 Gbp & 7.63G &  129 GB & 58.9 GB &      -- &       -- \\
@@
 \RS: AA, TT, AT, TA & 1.49 Tbp & 16.2G &      -- &  136 GB &  140 GB &   170 GB \\
-% arXiv:
-%AA                  &  433 Gbp & 4.69G &      -- & 38.5 GB & 39.9 GB &  48.3 GB \\
-%TT                  &  432 Gbp & 4.68G &      -- & 38.7 GB & 40.0 GB &  48.4 GB \\
-%AT                  &  275 Gbp & 2.98G &      -- & 26.6 GB & 26.9 GB &  32.6 GB \\
-%TA                  &  355 Gbp & 3.84G &      -- & 32.7 GB & 33.5 GB &  40.6 GB \\
+% arXiv: (individual datasets)
+AA                  &  433 Gbp & 4.69G &      -- & 38.5 GB & 39.9 GB &  48.3 GB \\
+TT                  &  432 Gbp & 4.68G &      -- & 38.7 GB & 40.0 GB &  48.4 GB \\
+AT                  &  275 Gbp & 2.98G &      -- & 26.6 GB & 26.9 GB &  32.6 GB \\
+TA                  &  355 Gbp & 3.84G &      -- & 32.7 GB & 33.5 GB &  40.6 GB \\
 Merged              & 1.49 Tbp & 16.2G &      -- &  117 GB &  126 GB & (152 GB) \\
@@
 % arXiv:
-%The temporary files used 287 to 306~gigabytes of disk space.
+The temporary files used 287 to 306~gigabytes of disk space.
 Thread buffer size and the number of merge buffers were the most important parameters.
@@
 Increasing run buffer size to 256~MB made the search phase faster with 512~MB thread buffers and slower with 256~MB thread buffers.
-%
+% (always continues the paragraph)
 For the further experiments, we chose 128~MB run buffers, 256~MB thread buffers, and 6 merge buffers (overhead 30.8~GB).
@@
 The running time of \BWTmerge{} was split evenly between \BWT{} construction and merging.
-% arXiv:
-%
+% arXiv: (paragraph break here)
+% DCC: (no paragraph break)
+
 When \ropebwt{} and \ropebwt2{} sort the reads in \emph{reverse lexicographic order} (RLO) to improve compression, all tools improve their performance.
@@
 % arXiv:
-%Again, \BWTmerge{} spent around half of the time building the individual \BWT{}s and another half merging them.
+Again, \BWTmerge{} spent around half of the time building the individual \BWT{}s and another half merging them.
 Note that \BWTmerge{} builds a \BWT{} for the concatenation of three input files that are in RLO, while the \BWT{}s produced by the other tools are completely in RLO.
@@
 % arXiv:
-%This makes it possible to host the indexes on two servers instead of the original three.
+This makes it possible to host the indexes on two servers instead of the original three.
 
@@
 % arXiv:
-%Read Server stores the reads in \emph{reverse lexicographic order} to improve compression \cite{Cox2012}.
+Read Server stores the reads in \emph{reverse lexicographic order} to improve compression \cite{Cox2012}.
 We can maintain this order with a few changes to the search phase \cite{Li2014a}.
@@
 With \emph{lexicographic order}, we can determine $\mSA[i]$ without samples by using $\mLF$ instead of $\Psi$, which is often faster in practice.
-%
-%We can also sort the reads by their likely positions in a reference genome.
+
+% arXiv:
+We can also sort the reads by their likely positions in a reference genome.
 This \emph{position order} is useful for both compression and storing the \emph{pairing information} for the reads.
@@
 
+% arXiv: (page break)
+\newpage
 \Section{References}
